---
title: "R Notebook"
output: html_notebook
---

```{r}
library(dplyr)
library(ggplot2)# Load the ape package
library(ape)
library(readxl)
library(tidyr)

# Read the .nwk file
tree <- read.tree("phylogeny_2022/22Nov_Naesiotus.nwk")
data <- read.nexus("phylogeny_2022/22Nov_Naesiotus.nex")

plot(tree)
```


parameters for plotting

```{r}
my_theme<-theme(axis.text=element_text(size=12),
        axis.title = element_text(size = 14),
        legend.text=element_text(size=10),
        legend.title = element_text(size=12),
        plot.title = element_text(face="bold",size=14,margin=margin(0,0,20,0),hjust = 0.5),
        axis.title.y = element_text(hjust = 0.5),
        axis.title.x = element_text(hjust = 0.5))
```


# compute phylogenetic distances


Read data

```{r}
df_rads <- read_excel("GPS RAD snails.xlsx")
head(df_rads)
```
```{r}
# Create a mapping between Sample and Taxon
sample_to_taxon <- setNames(df_rads$Taxon, df_rads$Sample)

# Replace tip labels in the tree with corresponding taxa
tree$tip.label <- sample_to_taxon[tree$tip.label]

plot(tree)
```



```{r}
df <- read_excel("04april24_datalab.xlsx", sheet = 2)
df$nematode_count[which(df$nematode_count == ">100")] <- "100"
df$nematode_count <- as.numeric(df$nematode_count)
hist(df$nematode_count)
```


```{r}

df <- df %>%
  mutate(nematode_count_categories = cut(nematode_count, 
                          breaks = c(-Inf, 0, 10, 30, Inf), 
                          labels = c("absence", "low", "moderate", "high"),
                          include.lowest = TRUE)) 

```

```{r}
# Filter out NA values from your data frame
filtered_df <- df %>%
  filter(!is.na(nematode_count_categories))

# Plotting the filtered data
total_counts <- ggplot(filtered_df, aes(x = nematode_count_categories, fill = nematode_count_categories)) +
  geom_bar(show.legend = FALSE) +
  labs(title = "Distribution of Categories",
       x = "Categories",
       y = "Count") +
  scale_fill_manual(values = c("absence" = "gray", "low" = "blue", "moderate" = "green", "high" = "red")) +
  xlab("") +
  theme_classic() +
  my_theme

total_counts

ggsave("output/exploratory/total_counts.jpeg", height = 7, width = 7)

```


# Species sampled per island

```{r}

# Count the number of species per island
species_counts <- filtered_df %>%
  group_by(island) %>%
  summarise(num_species = length(unique(spp)))

# Reorder the levels of the "island" factor based on the number of species
species_counts <- species_counts %>%
  mutate(island = factor(island, levels = island[order(num_species)]))


# Plot the number of species per island
nspp_island <- ggplot(species_counts, aes(x = island, y = num_species)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = num_species), vjust = -0.5, size = 5) +  # Add text labels
  labs(title = "Number of sampled species per island",
       x = "Island",
       y = "Number of Species") +
  theme(axis.text.x = element_text(angle = 45, vjust = c(1, 0))) +  # Adjust angle and vjust
  scale_x_discrete(guide = guide_axis(n.dodge=2))+
  theme_classic()+
  my_theme


nspp_island


ggsave("output/exploratory/nspp_island.jpeg", height = 7, width = 8)

```



# Parasitic load per species and island


```{r}

plot_counts_spp_island <- ggplot(df, aes(x = spp, fill = nematode_count_categories)) +
  geom_bar(stat = "count", position = "stack") +  # Use position = "stack" for stacked bar plots
  labs(title = "Nematode Counts per Species and Category",
       x = "Species",
       y = "Nematode Counts") +
  facet_wrap(~ island, scales = "free", strip.position = "top") +
  theme_classic() +
  theme(
    legend.position = "top",
    legend.box = "horizontal",
    legend.spacing.x = unit(0.2, "cm"),
    axis.text.x = element_text(angle = 45, hjust = 1) 
  ) +
 scale_fill_manual(values = c("absence" = "gray", "low" = "blue", "moderate" = "green", "high" = "red"))+
  xlab("")

plot_counts_spp_island

ggsave("output/exploratory/barplots_counts_spp_island_v2.jpeg", height = 9, width = 9)

```



```{r}
unique_pulls <- df %>%
  group_by(spp) %>%
  summarize(num_unique_pulls = n_distinct(id, na.rm = TRUE))


ggplot(unique_pulls, aes(x = num_unique_pulls)) +
  geom_bar(fill = "grey20", color = "black") +
  scale_y_continuous(breaks = seq(0, max(unique_pulls$num_unique_pulls), by = 1)) + 
  labs(title = "Distribution of n locations \nof examined specimens per species",
       x = "N locations in positive pulls",
       y = "Frequency") +
  theme_classic()+
  my_theme


ggsave("output/exploratory/barplots_nlocations_examinations.jpeg", height = 5, width = 9)

```



# compute a mean parasitic load

```{r}

df_load <- df %>%
  group_by(spp) %>%
  summarise(mean_load = mean(nematode_count, na.rm = TRUE)) %>%
  arrange(desc(mean_load))

colnames(df_load) = c("sp", "mean_load")

df_load

```


# shell brightness

```{r}
df_brightness <- read.csv("kraemer2019/kraemer2019_brightness_scores.csv")


df_brightness <- df_brightness[,c("sp", "brightness")]

# compute men brightness per spp

df_brightness <- df_brightness %>% 
  group_by(sp) %>% 
  summarise(mean_brightness = mean(brightness, na.rm = TRUE),
            sd_brightness = sd(brightness, na.rm = TRUE))

df_brightness <- df_brightness %>%
  filter(!(sp %in% c("AL sp.2", 
                     "AL sp.3",
                     "cf perspectivus FL",
                     "cf. nesioticus 'little'",
                     "cf. nesioticus SC E",
                     "cf. nesioticus SC W",
                     "unknownblack")))

```


```{r}
merged_df <- inner_join(df_load, df_brightness, by = "sp")

cor(merged_df$mean_brightness, merged_df$mean_load)


```



# n pools

```{r}

data <- read_excel("04april24_datalab.xlsx", sheet = "pools")


data <- mutate(data, id_negative_pull = ifelse(is.na(id_negative_pull), 0, id_negative_pull))

# Group by species (spp) and count the number of positive and negative pools
speciments_counts <- data %>%
  group_by(spp) %>%
  summarize(
    positive_pools = sum(!is.na(id_positive_pull)),
    negative_pools = sum(id_negative_pull > 0)
  )


speciments_counts


```
```{r}
# Assuming your data frame is called data
unique_pulls <- data %>%
  group_by(spp) %>%
  summarize(num_unique_pulls = n_distinct(id_positive_pull, na.rm = TRUE))

# View the result
print(unique_pulls)

ggplot(unique_pulls, aes(x = num_unique_pulls)) +
  geom_bar(fill = "grey20", color = "black") +
  labs(title = "Distribution of n locations of \npositive pulls per species",
       x = "N locations in positive pulls",
       y = "Frequency") +
  theme_classic()+
  my_theme


ggsave("output/exploratory/distrib_nlocations.jpeg", height = 6, width = 5)

```



```{r}

# Reshape the data into a longer format
speciments_counts_long <- pivot_longer(speciments_counts, cols = c(positive_pools, negative_pools), names_to = "pool_type", values_to = "count")

# Create a grouped bar plot
# Create a grouped bar plot with rotated x-axis labels and specified colors
# Create a grouped bar plot with rotated x-axis labels and specified colors
ggplot(speciments_counts_long, aes(x = spp, y = count, fill = ifelse(pool_type == "positive_pools" & count < 10, "less_than_10", pool_type))) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(title = "Number of Individuals in Pools per Species",
       x = "Species",
       y = "Number of Individuals",
       fill = "Pool Type") +
  scale_fill_manual(values = c("positive_pools" = "black", "negative_pools" = "gray", "less_than_10" = "red")) +  # Set colors for positive and negative pools
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))  # Rotate x-axis labels by 90 degrees


ggsave("output/exploratory/nind_pool_spp.png", height = 6, width = 8)



```





```{r}

unique(data$spp)

pool_counts <- data %>%
  group_by(spp) %>%
  summarize(
    positive_pool_types = n_distinct(id_positive_pull, na.rm = TRUE),  # Count distinct values, ignoring NA
    negative_pool_types = ifelse(any(id_negative_pull > 0), 1, 0)
  )

# Print the result
print(pool_counts)

```

Akamatus, Amastroides, and Basiplicatus do have negatives

Olla, Rabidensis, spp12 are those without negatives

Unifasciatus (from Floreana), and spp2 (from Wolf volcano in Isabela) are the unes used for the rarefraction analysis.




```{r}
# Print the result
n_samples <- sum(pool_counts$positive_pool_types) + sum(pool_counts$negative_pool_types) + 20


# budget calculation

n_samples * 27.15
```
















