---
title: "Analysis Centrality ~ speciation"

---


```{r}
library("here")
library("BAMMtools")
library("RPANDA")
#library("ggtree")
library("MetBrewer")
library("patchwork")
library("doParallel")
library("foreach")
library("MCMCglmm")
library(dplyr)
library(coda)
library(ggdist)
library(tidyverse)

here::i_am("centrality_speciation_regional.Rmd")
```


```{r}
## Importing tree
tree <- read.tree(here::here("clads/cleannames_tree_041024.txt"))

## Rescaling branch lengths by a scale of 10 to reflect time
tree$edge.length <- tree$edge.length * 10

## Making sure the tree is still ultrametric
is.ultrametric(tree)

## Replacing "." in species names for CLaDS not to complain
tree$tip.label <- gsub(".", "-", tree$tip.label, fixed = TRUE)

## Writing tree to file
#write.tree(tree, here::here("clads/naesiotus_rescaled.tre"))
#write.tree(tree, here::here("bamm/naesiotus_rescaled.tre"))

load(here::here("clads/naesiotus_clads_rates.RData"))

bamm.data <- getEventData(tree, here::here("bamm/event_data.txt"), burnin = 0.1)
```



distribution of trait states in the tips and speciation rates. The colors of the symbols at the tips describe the trait state, while both the icon size and the plot to the right show the distribution of speciation rates estimated using CLaDS:

```{r}
plotRateThroughTime(bamm.data, intervalCol = "lightgrey")

css <- credibleShiftSet(bamm.data, 1)
plot(css, lwd = 3)

tip.rates.bamm <- getTipRates(bamm.data, returnNetDiv = FALSE)
```



distribution of trait states in the tips and speciation rates. The colors of the symbols at the tips describe the trait state, while both the icon size and the plot to the right show the distribution of speciation rates estimated using CLaDS:

```{r}
plot_ClaDS_phylo(CladsOutput$tree, CladsOutput$lambdai_map, lwd = 3)
hist(CladsOutput$lambdatip_map)
```

distribution of trait states in the tips and speciation rates. The colors of the symbols at the tips describe the trait state, while both the icon size and the plot to the right show the distribution of speciation rates estimated using CLaDS:

```{r}
mean.tip.rates <- data.frame(
    species = tree$tip.label,
    lambda.bamm = round(setNames(apply(tip.rates.bamm[[1]], 1, mean), NULL), 5),
    mu.bamm = round(setNames(apply(tip.rates.bamm[[2]], 1, mean), NULL), 5),
    lambda.clads = round(CladsOutput$lambdatip_map, 5))

#write.table(mean.tip.rates, file = here::here("output/mean_tip_rates.csv"), sep = ",", quote = FALSE, row.names = FALSE)
```



# MCMCglmm


```{r}

## Branch indices
br.ind <- 4:(4+144)

## Tip indices
tip.ind <- (max(br.ind)+1):(max(br.ind)+Ntip(CladsOutput$tree))

## DTT indices
dtt.ind <- (max(tip.ind)+2):length(CladsOutput$chains[[1]])


## Checking parameter values
samp.ind <- sample(1:length(tip.ind), 1)
d_1 <- density(sapply(CladsOutput$chains,"[[", tip.ind[samp.ind])[251:1000,])
hist(sapply(CladsOutput$chains,"[[", tip.ind[samp.ind])[251:1000,], freq = FALSE, ylim = c(0, 1), breaks = 50)
lines(d_1)
abline(v = d_1$x[which.max(d_1$y)], col = 2, lwd = 3)
abline(v = CladsOutput$lambdatip_map[samp.ind], col = 3, lwd = 3)
legend("topright", legend = c("MAP", "lambda_tip"), lwd = 3, col = c(2, 3), cex = 2)
```



## compute dataset

```{r}
CladsOutput$tree$tip.label <- gsub("-", ".", 
                                   gsub("_", " ", CladsOutput$tree$tip.label))
```

PCA centrality, lineage size, and island area:

```{r}
centralities_df_regional <- readRDS(here::here("../output/tables/naesiotus_centralities_regional_df.rds"))
centralities_df_local <- readRDS(here::here("../output/tables/naesiotus_centralities_local_df.rds"))
metaweb <- readRDS(here::here("../output/temporal_output/bipartite_matrix.rds"))
sp_island_id <- readRDS(here::here("../output/temporal_output/sp_island_id.rds"))
species_names_checks.csv <- read.csv("../data/species_names_checks.csv")

```


Speciation rate and island ages are sampled from estimated distributions:

```{r}

###############  speciation rates

## Extracting posterior distributions for the tips
post.dist.tips <- lapply(1:length(tip.ind), function(x){sapply(CladsOutput$chains, "[[", tip.ind[x])})

## Discarding burnin
post.dist.tips <- lapply(post.dist.tips, function(x){x[251:1000,]})

## Creating data.frame with sampled rates for each tip
sampled.rates <- as.data.frame(t(plyr::ldply(1:50, .fun = function(x){sapply(post.dist.tips, function(y){sample(y, 1)})})))
names(sampled.rates) <- paste0("species_speciation_rate_", 1:50)



############### island ages

# Sample ages for each species on each island
sp_island_guild_size <- sp_island_id |>
  group_by(island) |>
  mutate(
    sampled_age = runif(n(), min = min_emergence, max = max_emergence)
  ) |>
  ungroup()

# Number of simulations
num_simulations <- 50

# Create the sampled.ages data frame with species_name
sampled.ages <- as.data.frame(
  replicate(num_simulations, {
    sp_island_guild_size |>
      group_by(island) |>
      mutate(
        sampled_age = runif(n(), min = min_emergence, max = max_emergence)
      ) |>
      ungroup() |>
      pull(sampled_age)
  })
)

# Rename columns for simulations
names(sampled.ages) <- paste0("island_age_", 1:num_simulations)

# Add the species_name column
sampled.ages$species_name <- sp_island_guild_size$species_name

# Reorder columns to have species_name first (optional)
sampled.ages <- sampled.ages |>
  dplyr::relocate(species_name)
               
sampled.ages

```




Regional

```{r}
centralities_df_regional <- centralities_df_regional |>
  dplyr::left_join(
    species_names_checks.csv |>
      dplyr::select(annotated_name, species_name) |>
      dplyr::rename(species = annotated_name),  # Rename for joining
    by = "species"
  )

# Filter sp_island_id to retain only species in centralities_df_regional
filtered_sp_island_id_regional <- sp_island_id |>
  dplyr::filter(species_name %in% centralities_df_regional$species_name)

# Add the guild_size column
sp_island_guild_size_regional <- filtered_sp_island_id_regional |>
  dplyr::group_by(island) |>
  dplyr::mutate(guild_size = n_distinct(species_name)) |>
  dplyr::ungroup()

# Add PCA_centrality to sp_island_guild_size_regional
sp_island_guild_size_regional <- sp_island_guild_size_regional |>
  dplyr::left_join(
    centralities_df_regional,
    by = "species_name"
  )

```


```{r}
centralities_df_local <- centralities_df_local |>
  dplyr::left_join(
    species_names_checks.csv |>
      dplyr::select(annotated_name, species_name) |>
      dplyr::rename(species = annotated_name),  # Rename for joining
    by = "species"
  )

# Filter sp_island_id to retain only species in centralities_df_regional
filtered_sp_island_id_local <- sp_island_id |>
  dplyr::filter(species_name %in% centralities_df_local$species_name)

# Add the guild_size column
sp_island_guild_size_local <- filtered_sp_island_id_local |>
  dplyr::group_by(island) |>
  dplyr::mutate(guild_size = n_distinct(species_name)) |>
  dplyr::ungroup()

# Add PCA_centrality to sp_island_guild_size_regional
sp_island_guild_size_local <- sp_island_guild_size_local |>
  dplyr::left_join(
    centralities_df_local,
    by = "species_name"
  )

```




Add species names to dataset sampled_rates

```{r}
# Assuming CladsOutput$tree$tip.label has species names in the same order as sampled_rates rows
species_order <- CladsOutput$tree$tip.label
# Add species names to sampled_rates
sampled.rates$species_name <- species_order

# REGIONAL

# Filter sampled_rates to keep only species present in sp_island_guild_size_regional
filtered_sampled_rates <- sampled.rates |>
  dplyr::filter(species_name %in% sp_island_guild_size_regional$species_name)

# Local
filtered_sampled_rates_local <- sampled.rates |>
  dplyr::filter(species_name %in% sp_island_guild_size_local$species_name)

```




```{r}
# Regional

# Add sampled rates to sp_island_guild_size_regional
sp_island_guild_size_regional <- sp_island_guild_size_regional |>
  dplyr::left_join(
    filtered_sampled_rates,
    by = "species_name"
  )

sp_island_guild_size_regional


# Local

# Add sampled rates to sp_island_guild_size_regional
sp_island_guild_size_local <- sp_island_guild_size_local |>
  dplyr::left_join(
    filtered_sampled_rates_local,
    by = "species_name"
  )

sp_island_guild_size_local
```

Add island ages sampled

```{r}

# regional

# Add sampled.ages to sp_island_guild_size_regional
sp_island_guild_size_regional <- sp_island_guild_size_regional |>
  dplyr::left_join(
    sampled.ages,
    by = "species_name"
  )

sp_island_guild_size_regional


# local

# Add sampled.ages to sp_island_guild_size_regional
sp_island_guild_size_local <- sp_island_guild_size_local |>
  dplyr::left_join(
    sampled.ages,
    by = "species_name"
  )

sp_island_guild_size_local

```


Gardner island merged with Espanola

```{r}
sp_island_guild_size_regional <- sp_island_guild_size_regional %>%
  mutate(island = ifelse(island == "Gardner", "Espanola", island))

sp_island_guild_size_local <- sp_island_guild_size_local %>%
  rename(island = island.x) |> 
  mutate(island = ifelse(island == "Gardner", "Espanola", island))
```




```{r}
missing_species <- setdiff(sp_island_guild_size_regional$species_name, CladsOutput$tree$tip.label)
if (length(missing_species) > 0) {
    print("These species are in the dataset but not in the tree:")
    print(missing_species)
}

```
Remove species not in the tree

```{r}
# regional

sp_island_guild_size_regional_filtered <- sp_island_guild_size_regional |>
    dplyr::filter(species_name %in% CladsOutput$tree$tip.label)

 Ainv <- inverseA(CladsOutput$tree)$Ainv

sp_island_guild_size_regional_filtered <- sp_island_guild_size_regional_filtered |>
    dplyr::filter(species_name %in% rownames(Ainv))


# local 

sp_island_guild_size_local_filtered <- sp_island_guild_size_local |>
    dplyr::filter(species_name %in% CladsOutput$tree$tip.label)

sp_island_guild_size_local_filtered <- sp_island_guild_size_local_filtered |>
    dplyr::filter(species_name %in% rownames(Ainv))

all(sp_island_guild_size_regional_filtered$species_name %in% rownames(Ainv))
```

Match levels in ginverse

```{r}
 Ainv <- inverseA(CladsOutput$tree)$Ainv

```


## CHECKS


Regional

```{r}

colnames(Ainv)  # Should return NULL
colnames(Ainv) <- rownames(Ainv)
common_species <- intersect(sp_island_guild_size_regional_filtered$species_name, rownames(Ainv))
Ainv_filtered <- Ainv[common_species, common_species]

dim(Ainv_filtered)  # Should match the number of species in common_species
length(common_species)  # Should be equal to the number of species in filtered Ainv

ginverse <- list(species_name = Ainv_filtered)

setdiff(sp_island_guild_size_regional_filtered$species_name, common_species)
all(sp_island_guild_size_regional_filtered$species_name %in% rownames(Ainv_filtered))
```


Local

```{r}
common_species <- intersect(sp_island_guild_size_local_filtered$species_name, rownames(Ainv))
colnames(Ainv) <- rownames(Ainv)
Ainv_filtered <- Ainv[common_species, common_species]

dim(Ainv_filtered)  # Should match the number of species in common_species
length(common_species)  # Should be equal to the number of species in filtered Ainv

ginverse <- list(species_name = Ainv_filtered)

setdiff(sp_island_guild_size_local_filtered$species_name, common_species)
all(sp_island_guild_size_local_filtered$species_name %in% rownames(Ainv_filtered))
```

Regional

```{r}
sp_island_guild_size_regional_filtered$species_name <- factor(
    sp_island_guild_size_regional_filtered$species_name,
    levels = rownames(ginverse$species_name)
)

setdiff(levels(sp_island_guild_size_regional_filtered$species_name), rownames(ginverse$species_name))
setdiff(rownames(ginverse$species_name), levels(sp_island_guild_size_regional_filtered$species_name))

anyDuplicated(levels(sp_island_guild_size_regional_filtered$species_name))
anyDuplicated(rownames(ginverse$species_name))

all(sp_island_guild_size_regional_filtered$species_name %in% rownames(ginverse$species_name))

required_columns <- c(
    "PCA_centrality", paste0("species_speciation_rate_", 1), 
    paste0("island_age_", 1), "guild_size", "island_area", "species_name"
)
missing_columns <- setdiff(required_columns, colnames(sp_island_guild_size_regional_filtered))
missing_columns

```


Local

```{r}
sp_island_guild_size_local_filtered$species_name <- factor(
    sp_island_guild_size_local_filtered$species_name,
    levels = rownames(ginverse$species_name)
)

setdiff(levels(sp_island_guild_size_local_filtered$species_name), rownames(ginverse$species_name))
setdiff(rownames(ginverse$species_name), levels(sp_island_guild_size_local_filtered$species_name))

anyDuplicated(levels(sp_island_guild_size_local_filtered$species_name))
anyDuplicated(rownames(ginverse$species_name))

all(sp_island_guild_size_local_filtered$species_name %in% rownames(ginverse$species_name))

required_columns <- c(
    "PCA_centrality", paste0("species_speciation_rate_", 1), 
    paste0("island_age_", 1), "guild_size", "island_area", "species_name"
)
missing_columns <- setdiff(required_columns, colnames(sp_island_guild_size_local_filtered))
missing_columns
```




```{r}
sp_island_guild_size_regional_filtered <- as.data.frame(sp_island_guild_size_regional_filtered)
sp_island_guild_size_local_filtered <- as.data.frame(sp_island_guild_size_local_filtered)
```



Standarize variables

Z score guild_size and island_area

```{r}
sp_island_guild_size_regional_filtered <- sp_island_guild_size_regional_filtered %>%
    mutate(
        guild_size_z = scale(guild_size) %>% as.numeric(),
        island_area_z = scale(island_area) %>% as.numeric()
    )


sp_island_guild_size_local_filtered <- sp_island_guild_size_local_filtered %>%
    mutate(
        guild_size_z = scale(guild_size) %>% as.numeric(),
        island_area_z = scale(island_area) %>% as.numeric()
    )

```

z score island ages and speciation rates (sets) REGIONAL

```{r}
# Identify the columns to z-score
age_columns <- grep("^island_age_", colnames(sp_island_guild_size_regional_filtered), value = TRUE)
speciation_rate_columns <- grep("^species_speciation_rate_", colnames(sp_island_guild_size_regional_filtered), value = TRUE)

# Z-score the columns and create new ones with a "_z" suffix
sp_island_guild_size_regional_filtered <- sp_island_guild_size_regional_filtered %>%
  mutate(across(
    all_of(c(age_columns, speciation_rate_columns)),
    ~ scale(.) %>% as.numeric(),
    .names = "{.col}_z"
  ))

sp_island_guild_size_regional_filtered
```


Z score speciation rates and centrality by island - LOCAL

```{r}
# Identify the columns to z-score
age_columns <- grep("^island_age_", colnames(sp_island_guild_size_local_filtered), value = TRUE)
speciation_rate_columns <- grep("^species_speciation_rate_", colnames(sp_island_guild_size_local_filtered), value = TRUE)

# Z-score speciation rates by island
sp_island_guild_size_local_filtered <- sp_island_guild_size_local_filtered %>%
  group_by(island) %>%
  mutate(across(
    all_of(speciation_rate_columns),
    ~ scale(.) %>% as.numeric(),
    .names = "{.col}_z"
  )) %>%
  ungroup()  # Ungroup after z-scoring speciation rates

# Z-score island ages across the entire dataset
sp_island_guild_size_local_filtered <- sp_island_guild_size_local_filtered %>%
  mutate(across(
    all_of(age_columns),
    ~ scale(.) %>% as.numeric(),
    .names = "{.col}_z"
  ))


sp_island_guild_size_local_filtered
```

------------------------------------------------------------------------------


# Models

## Full model with PCA centrality

### REGIONAL

```{r}

my_priors <- list(
  R = list(V = 1/2, nu = 0.002),
  G = list(G1 = list(V = 1/2, nu = 0.002))
)

# Initialize an empty list to store model fits
full_model.fit <- vector(mode = "list", length = 50)

# Loop through the 50 sets of speciation rates and island ages
for (i in 1:50) {
  # Construct the formula
  formula_string <- paste0(
    "PCA_centrality ~ species_speciation_rate_", i, "_z + ", 
    "island_age_", i, "_z + island_area_z + guild_size_z"
  )
  
  # Fit the model
  full_model.fit[[i]] <- MCMCglmm(
    as.formula(formula_string),  # Use as.formula() directly
    random = ~ species_name,  # Random effect for species_name
    family = "gaussian",  # Assuming a Gaussian response for PCA centrality
    ginverse = ginverse,  # Use species_name in ginverse
    data = as.data.frame(sp_island_guild_size_regional_filtered),  # Dataframe
    prior = my_priors,  # Priors
    nitt = 1000000,  # Number of iterations
    burnin = 500000,  # Burn-in
    thin = 100  # Thinning
  )
}




saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/full_model.pca_regional.rds"))

```



### LOCAL

```{r}

my_priors <- list(
  R = list(V = 1/2, nu = 0.002),
  G = list(G1 = list(V = 1/2, nu = 0.002))
)

# Initialize an empty list to store model fits
full_model.fit <- vector(mode = "list", length = 50)

# Loop through the 50 sets of speciation rates and island ages
for (i in 1:50) {
  # Construct the formula
  formula_string <- paste0(
    "PCA_centrality ~ species_speciation_rate_", i, "_z + ", 
    "island_age_", i, "_z + island_area_z + guild_size_z"
  )
  
  # Fit the model
  full_model.fit[[i]] <- MCMCglmm(
    as.formula(formula_string),  # Use as.formula() directly
    random = ~ species_name,  # Random effect for species_name
    family = "gaussian",  # Assuming a Gaussian response for PCA centrality
    ginverse = ginverse,  # Use species_name in ginverse
    data = as.data.frame(sp_island_guild_size_local_filtered),  # Dataframe
    prior = my_priors,  # Priors
    nitt = 1000000,  # Number of iterations
    burnin = 500000,  # Burn-in
    thin = 100  # Thinning
  )
}




saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/full_model.pca_local.rds"))

```



## Full model with specialization


```{r}

my_priors <- list(
  R = list(V = 1/2, nu = 0.002),
  G = list(G1 = list(V = 1/2, nu = 0.002))
)

# Initialize an empty list to store model fits
full_model.fit <- vector(mode = "list", length = 50)

# Loop through the 50 sets of speciation rates and island ages
for (i in 1:50) {
  # Construct the formula
  formula_string <- paste0(
    "d_z ~ species_speciation_rate_", i, "_z + ", 
    "island_age_", i, "_z + island_area_z + guild_size_z"
  )
  
  # Fit the model
  full_model.fit[[i]] <- MCMCglmm(
    as.formula(formula_string),  # Use as.formula() directly
    random = ~ species_name,  # Random effect for species_name
    family = "gaussian",  # Assuming a Gaussian response for PCA centrality
    ginverse = ginverse,  # Use species_name in ginverse
    data = as.data.frame(sp_island_guild_size_local_filtered),  # Dataframe
    prior = my_priors,  # Priors
    nitt = 1000000,  # Number of iterations
    burnin = 500000,  # Burn-in
    thin = 100  # Thinning
  )
}




saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/full_model.d_local.rds"))

```





```{r}
#model.fit <- readRDS(here::here("../output/models/network_role_speciation/full_model.pca_regional.rds"))
# model.fit <- readRDS(here::here("../output/models/network_role_speciation/full_model.d_regional.rds"))
# model.fit <- readRDS(here::here("../output/models/network_role_speciation/full_model.pca_local.rds"))
 model.fit <- readRDS(here::here("../output/models/network_role_speciation/full_model.d_local.rds"))

```



# Model results

```{r}
# Extract fixed effects (Sol) and variance components (VCV)
fixed_effects <- lapply(model.fit, function(model) as.mcmc(model$Sol))
variance_components <- lapply(model.fit, function(model) as.mcmc(model$VCV))
```



```{r}
# Rename the columns to standard names
fixed_effects_standardized <- lapply(seq_along(fixed_effects), function(i) {
  colnames(fixed_effects[[i]]) <- c("(Intercept)", "species_speciation_rate", 
                                    "guild_size", "island_age", "island_area")
  fixed_effects[[i]]
})


## combine chains

combined_fixed <- do.call(mcmc.list, fixed_effects_standardized)


```


## CHAINS CONVERGENCE

#### Traceplot

```{r}
traceplot(combined_fixed)
```

The traceplots show horizontal bands without trends or drifts, and they overlap and mix well, indicating that chains converged.

#### Autocorrelation plot

```{r}
autocorr.plot(combined_fixed)
```

There is a fast decay to zero in all cases, indicating that samples become effectively independent quickly, and there is no autocorrelation detected.

#### Gelman-Rubin diagnostic

```{r}
gelman_fixed <- gelman.diag(combined_fixed)
print(gelman_fixed)
```
The Gelman-Rubin diagnostic, also called the potential scale reduction factor (PSRF), is a method to assess the convergence of Markov Chain Monte Carlo (MCMC) simulations. It compares the variance within each chain to the variance between chains to determine whether the chains have converged to the same target distribution.

Interpretation:
Point Estimate: The closer this value is to 1, the better the convergence. Values above 1.1 suggest that the chains have not fully converged.
Upper C.I.: The upper bound of the confidence interval for the PSRF. This helps account for uncertainty in the diagnostic. Values significantly above 1.1 may indicate poor convergence.

Convergence is problematic for species_speciation_rate.


## Effective Sample Size

```{r}
effectiveSize(model.fit[[7]]$Sol[,2])
```


## Posterior distributions

```{r}
# Summarize fixed effects across all models
summary_fixed <- summary(combined_fixed)

print(summary_fixed)
```



```{r}
# Combine the posterior distributions into one long data frame
# Combine the posterior distributions into one long data frame
posterior_long <- do.call(rbind, lapply(seq_along(fixed_effects_standardized), function(i) {
  as.data.frame(as.matrix(fixed_effects_standardized[[i]])) %>%
    mutate(Model = paste0("Model_", i)) %>%
    pivot_longer(cols = -Model, names_to = "Predictor", values_to = "Value")  # Exclude Model from pivoting
}))

# Plot the posterior distributions as density curves
ggplot(posterior_long, aes(x = Value, color = Model, group = Model)) +
  geom_density(size = 0.5, alpha = 0.8) +  # Add density curves
  facet_wrap(~Predictor, scales = "fixed", ncol = 1) +  # Fixed scales for alignment
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +  # Vertical line at 0
  labs(
    x = "Posterior Value",
    y = "Density",
    title = "Posterior Distributions Across Models"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",  # Remove legend for clarity
    panel.grid.major.y = element_blank()
  )

#ggsave(here::here("../output/figures/posterior_pars/d_local.png"), height = 10, width = 5)
```

