---
title: "Analysis Centrality ~ speciation"

---


```{r}
library("here")
library("BAMMtools")
library("RPANDA")
#library("ggtree")
library("MetBrewer")
library("patchwork")
library("doParallel")
library("foreach")
library("MCMCglmm")
library(dplyr)

here::i_am("centrality_speciation.Rmd")
```


```{r}
## Importing tree
tree <- read.tree(here::here("clads/cleannames_tree_041024.txt"))

## Rescaling branch lengths by a scale of 10 to reflect time
tree$edge.length <- tree$edge.length * 10

## Making sure the tree is still ultrametric
is.ultrametric(tree)

## Replacing "." in species names for CLaDS not to complain
tree$tip.label <- gsub(".", "-", tree$tip.label, fixed = TRUE)

## Writing tree to file
write.tree(tree, here::here("clads/naesiotus_rescaled.tre"))
write.tree(tree, here::here("bamm/naesiotus_rescaled.tre"))

load(here::here("clads/naesiotus_clads_rates.RData"))

bamm.data <- getEventData(tree, here::here("bamm/event_data.txt"), burnin = 0.1)
```



distribution of trait states in the tips and speciation rates. The colors of the symbols at the tips describe the trait state, while both the icon size and the plot to the right show the distribution of speciation rates estimated using CLaDS:

```{r}
plotRateThroughTime(bamm.data, intervalCol = "lightgrey")

css <- credibleShiftSet(bamm.data, 1)
plot(css, lwd = 3)

tip.rates.bamm <- getTipRates(bamm.data, returnNetDiv = FALSE)
```



distribution of trait states in the tips and speciation rates. The colors of the symbols at the tips describe the trait state, while both the icon size and the plot to the right show the distribution of speciation rates estimated using CLaDS:

```{r}
plot_ClaDS_phylo(CladsOutput$tree, CladsOutput$lambdai_map, lwd = 3)
hist(CladsOutput$lambdatip_map)
```

distribution of trait states in the tips and speciation rates. The colors of the symbols at the tips describe the trait state, while both the icon size and the plot to the right show the distribution of speciation rates estimated using CLaDS:

```{r}
mean.tip.rates <- data.frame(
    species = tree$tip.label,
    lambda.bamm = round(setNames(apply(tip.rates.bamm[[1]], 1, mean), NULL), 5),
    mu.bamm = round(setNames(apply(tip.rates.bamm[[2]], 1, mean), NULL), 5),
    lambda.clads = round(CladsOutput$lambdatip_map, 5))

#write.table(mean.tip.rates, file = here::here("output/mean_tip_rates.csv"), sep = ",", quote = FALSE, row.names = FALSE)
```



# MCMCglmm


```{r}
## Set the number of cores to be used in the parallel processing
registerDoMC(4)

## Branch indices
br.ind <- 4:(4+144)

## Tip indices
tip.ind <- (max(br.ind)+1):(max(br.ind)+Ntip(CladsOutput$tree))

## DTT indices
dtt.ind <- (max(tip.ind)+2):length(CladsOutput$chains[[1]])


## Checking parameter values
samp.ind <- sample(1:length(tip.ind), 1)
d_1 <- density(sapply(CladsOutput$chains,"[[", tip.ind[samp.ind])[251:1000,])
hist(sapply(CladsOutput$chains,"[[", tip.ind[samp.ind])[251:1000,], freq = FALSE, ylim = c(0, 1), breaks = 50)
lines(d_1)
abline(v = d_1$x[which.max(d_1$y)], col = 2, lwd = 3)
abline(v = CladsOutput$lambdatip_map[samp.ind], col = 3, lwd = 3)
legend("topright", legend = c("MAP", "lambda_tip"), lwd = 3, col = c(2, 3), cex = 2)
```



## compute dataset

```{r}
CladsOutput$tree$tip.label <- gsub("-", ".", 
                                   gsub("_", " ", CladsOutput$tree$tip.label))
```




Speciation rate and island ages are sampled from estimated distributions:

```{r}

###############  speciation rates

## Extracting posterior distributions for the tips
post.dist.tips <- lapply(1:length(tip.ind), function(x){sapply(CladsOutput$chains, "[[", tip.ind[x])})

## Discarding burnin
post.dist.tips <- lapply(post.dist.tips, function(x){x[251:1000,]})

## Creating data.frame with sampled rates for each tip
sampled.rates <- as.data.frame(t(plyr::ldply(1:50, .fun = function(x){sapply(post.dist.tips, function(y){sample(y, 1)})})))
names(sampled.rates) <- paste0("species_speciation_rate_", 1:50)



############### island ages

# Sample ages for each species on each island
sp_island_guild_size <- sp_island_guild_size |>
  group_by(island) |>
  mutate(
    sampled_age = runif(n(), min = min_emergence, max = max_emergence)
  ) |>
  ungroup()

# Number of simulations
num_simulations <- 50

# Create the sampled.ages data frame with species_name
sampled.ages <- as.data.frame(
  replicate(num_simulations, {
    sp_island_guild_size |>
      group_by(island) |>
      mutate(
        sampled_age = runif(n(), min = min_emergence, max = max_emergence)
      ) |>
      ungroup() |>
      pull(sampled_age)
  })
)

# Rename columns for simulations
names(sampled.ages) <- paste0("island_age_", 1:num_simulations)

# Add the species_name column
sampled.ages$species_name <- sp_island_guild_size$species_name

# Reorder columns to have species_name first (optional)
sampled.ages <- sampled.ages |>
  dplyr::relocate(species_name)
               
sampled.ages

# Add sampled.ages to sp_island_guild_size
sp_island_guild_size <- sp_island_guild_size |>
  dplyr::left_join(
    sampled.ages,
    by = "species_name"
  )

```



PCA centrality, lineage size, and island area:

```{r}
centralities_df <- readRDS(here::here("../output/tables/centralities_df.rds")) |> dplyr::select(species, PCA_centrality)
metaweb <- readRDS(here::here("../output/temporal_output/bipartite_matrix.rds"))
sp_island_id <- readRDS(here::here("../output/temporal_output/sp_island_id.rds"))
species_names_checks.csv <- read.csv("../data/species_names_checks.csv")

```


```{r}
centralities_df <- centralities_df |>
  dplyr::left_join(
    species_names_checks.csv |>
      dplyr::select(annotated_name, species_name) |>
      dplyr::rename(species = annotated_name),  # Rename for joining
    by = "species"
  )

# Filter sp_island_id to retain only species in centralities_df
filtered_sp_island_id <- sp_island_id |>
  dplyr::filter(species_name %in% centralities_df$species_name)

# Add the guild_size column
sp_island_guild_size <- filtered_sp_island_id |>
  dplyr::group_by(island) |>
  dplyr::mutate(guild_size = n_distinct(species_name)) |>
  dplyr::ungroup()

# Add PCA_centrality to sp_island_guild_size
sp_island_guild_size <- sp_island_guild_size |>
  dplyr::left_join(
    centralities_df |>
      dplyr::select(species_name, PCA_centrality),
    by = "species_name"
  )

```


Add species names to dataset sampled_rates

```{r}
# Assuming CladsOutput$tree$tip.label has species names in the same order as sampled_rates rows
species_order <- CladsOutput$tree$tip.label
# Add species names to sampled_rates
sampled.rates$species_name <- species_order

# Filter sampled_rates to keep only species present in sp_island_guild_size
filtered_sampled_rates <- sampled.rates |>
  dplyr::filter(species_name %in% sp_island_guild_size$species_name)


```



```{r}
# Add sampled rates to sp_island_guild_size
sp_island_guild_size <- sp_island_guild_size |>
  dplyr::left_join(
    filtered_sampled_rates,
    by = "species_name"
  )

sp_island_guild_size
```



## Model


```{r}

my_priors <- list(R = list(V = 1/2, nu = 0.002),
                  G = list(G1 = list(V = 1/2, nu = 0.002)))

#registerDoMC(2)

# model.fit <-
#     foreach(i = 1:50) %dopar% {
#     MCMCglmm(formula(paste0("PCA_centrality ~ species_speciation_rate_", i, " + lineage_size + island_age_", i, " + island_area")), # Fixed effects 
#              random = ~ species, # Random intercept for clade and phylogenetic effect family = "gaussian", # Assuming a Gaussian response for PCA centrality 
#              ginverse = list(species = inverseA(CladsOutput$tree)$Ainv), # Inverse phylogenetic covariance matrix for species 
#              data = fulldata, 
#              prior = my_priors, 
#              nitt =  10000, 
#              burnin =  1000,
#              thin =  100)
#     }

model.fit <- vector(mode = "list", length = 50)
for(i in 1:50){
    model.fit[[i]] <-
        MCMCglmm(formula(paste0("PCA_centrality ~ species_speciation_rate_", i, " + lineage_size + island_age_", i, " + island_area")), # Fixed effects 
                 random = ~ species_name, # Random intercept for clade and phylogenetic effect family = "gaussian", # Assuming a Gaussian response for PCA centrality 
                 ginverse = list(species = inverseA(CladsOutput$tree)$Ainv), # Inverse phylogenetic covariance matrix for species 
                 data = sp_island_guild_size, 
                 prior = my_priors, 
                 nitt =  1000000, 
                 burnin =  500000,
                 thin =  100)
}
```

