---
title: "Analysis Centrality ~ speciation"

---


```{r}
library("here")
library("BAMMtools")
library("RPANDA")
library("ggtree")
library("MetBrewer")
library("patchwork")
library("doParallel")
library("foreach")
library("MCMCglmm")
library(dplyr)
library(coda)
library(ggdist)
library(tidyverse)

here::i_am("centrality_speciation.Rmd")
```


## Load data

```{r}
## Importing tree
tree <- read.tree(here::here("clads/cleannames_tree_041024.txt"))

## Rescaling branch lengths by a scale of 10 to reflect time
tree$edge.length <- tree$edge.length * 10

## Making sure the tree is still ultrametric
is.ultrametric(tree)

## Replacing "." in species names for CLaDS not to complain
tree$tip.label <- gsub(".", "-", tree$tip.label, fixed = TRUE)

## Writing tree to file
#write.tree(tree, here::here("clads/naesiotus_rescaled.tre"))
#write.tree(tree, here::here("bamm/naesiotus_rescaled.tre"))

load(here::here("clads/naesiotus_clads_rates.RData"))

bamm.data <- getEventData(tree, here::here("bamm/event_data.txt"), burnin = 0.1)
```



# Plot BAMM


## Rate through time BAMM


distribution of trait states in the tips and speciation rates. The colors of the symbols at the tips describe the trait state, while both the icon size and the plot to the right show the distribution of speciation rates estimated using CLaDS:

```{r}
plotRateThroughTime(bamm.data, intervalCol = "lightgrey")

css <- credibleShiftSet(bamm.data, 1)
plot(css, lwd = 3)

tip.rates.bamm <- getTipRates(bamm.data, returnNetDiv = FALSE)
```



## Posteriors BAMM

```{r}
# Extract the lambda matrix from BAMM output
lambda_rates <- tip.rates.bamm$lambda  # Matrix of speciation rates

# Identify tip species based on rownames (assuming tip species are present as rownames)
tip_species <- rownames(lambda_rates)

# Filter the posterior rates for the tips
# The assumption is that all rows in the lambda matrix represent species, including tips.
tip_rates <- lambda_rates[tip_species, ]

# Prepare the data for plotting
tip_rates_df <- as.data.frame(t(tip_rates))  # Transpose for long format
tip_species_names <- rownames(tip_rates)  # Species names for plotting

# Convert to long format for ggplot
tip_rates_long <- data.frame(
  Species = rep(tip_species_names, each = nrow(tip_rates_df)),
  Rate = as.vector(as.matrix(tip_rates_df)),
  Simulation = rep(1:nrow(tip_rates_df), times = length(tip_species_names))
)

# Generate individual density plots for each tip species
plots <- lapply(unique(tip_rates_long$Species), function(species) {
  ggplot(subset(tip_rates_long, Species == species), aes(x = Rate)) +
    geom_density(fill = "gold4", alpha = 0.5) +
    labs(
      title = species,
      x = "Speciation rate",
      y = "Density"
    ) +
    theme_minimal()
})


# Number of plots per page
plots_per_page <- 20

# Split the plots into pages
plot_pages <- split(plots, ceiling(seq_along(plots) / plots_per_page))

# Save or display each page
for (i in seq_along(plot_pages)) {
  gridExtra::grid.arrange(grobs = plot_pages[[i]], ncol = 4)
}

```



# Plot ClaDS


distribution of trait states in the tips and speciation rates. The colors of the symbols at the tips describe the trait state, while both the icon size and the plot to the right show the distribution of speciation rates estimated using CLaDS:

```{r}
plot_ClaDS_phylo(CladsOutput$tree, CladsOutput$lambdai_map, lwd = 3)
hist(CladsOutput$lambdatip_map)
```

distribution of trait states in the tips and speciation rates. The colors of the symbols at the tips describe the trait state, while both the icon size and the plot to the right show the distribution of speciation rates estimated using CLaDS:


```{r}
mean.tip.rates <- data.frame(
    species = tree$tip.label,
    lambda.bamm = round(setNames(apply(tip.rates.bamm[[1]], 1, mean), NULL), 5),
    mu.bamm = round(setNames(apply(tip.rates.bamm[[2]], 1, mean), NULL), 5),
    lambda.clads = round(CladsOutput$lambdatip_map, 5))

#write.table(mean.tip.rates, file = here::here("output/mean_tip_rates.csv"), sep = ",", quote = FALSE, row.names = FALSE)
```



## ClaDS rates


```{r}

## Branch indices
br.ind <- 4:(4+144)

## Tip indices
tip.ind <- (max(br.ind)+1):(max(br.ind)+Ntip(CladsOutput$tree))

## DTT indices
dtt.ind <- (max(tip.ind)+2):length(CladsOutput$chains[[1]])


## Checking parameter values
samp.ind <- sample(1:length(tip.ind), 1)
d_1 <- density(sapply(CladsOutput$chains,"[[", tip.ind[samp.ind])[251:1000,])
hist(sapply(CladsOutput$chains,"[[", tip.ind[samp.ind])[251:1000,], freq = FALSE, ylim = c(0, 1), breaks = 50)
lines(d_1)
abline(v = d_1$x[which.max(d_1$y)], col = 2, lwd = 3)
abline(v = CladsOutput$lambdatip_map[samp.ind], col = 3, lwd = 3)
legend("topright", legend = c("MAP", "lambda_tip"), lwd = 3, col = c(2, 3), cex = 2)
```



Speciation rate sampled from estimated distributions:

```{r}

###############  speciation rates

## Extracting posterior distributions for the tips
post.dist.tips <- lapply(1:length(tip.ind), function(x){sapply(CladsOutput$chains, "[[", tip.ind[x])})

## Discarding burnin
post.dist.tips <- lapply(post.dist.tips, function(x){x[251:1000,]})

## Creating data.frame with sampled rates for each tip
sampled.rates <- as.data.frame(t(plyr::ldply(1:50, .fun = function(x){sapply(post.dist.tips, function(y){sample(y, 1)})})))
names(sampled.rates) <- paste0("species_speciation_rate_", 1:50)




```

```{r}
# Assuming CladsOutput$tree$tip.label has species names in the same order as sampled_rates rows
#!!!!!
species_order <- CladsOutput$tree$tip.label
# Add species names to sampled_rates
sampled.rates$species_name <- species_order
```


```{r}
# Reshaping the data for ggplot
sampled.rates$Species <- rownames(sampled.rates)  # Adding species names as a column
sampled.rates_long <- tidyr::pivot_longer(
  sampled.rates,
  cols = starts_with("species_speciation_rate_"),
  names_to = "Sample",
  values_to = "Rate"
)

ggplot(sampled.rates_long, aes(x = Rate)) +
  geom_density(fill = "blue", alpha = 0.4) +
  facet_wrap(~ species_name, scales = "free_y", ncol = 4) +  # Facet by species
  labs(
    title = "Posterior Distributions of Tip-Specific Speciation Rates",
    x = "Speciation Rate",
    y = "Density"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 10),  # Adjust species label size
    panel.spacing = unit(0.5, "lines")     # Spacing between panels
  )

```

## compute dataset

```{r}
# CladsOutput$tree$tip.label <- gsub("-", ".", 
#                                    gsub("_", " ", CladsOutput$tree$tip.label))
```

PCA centrality, lineage size, and island area:

```{r}
centralities_df_regional <- readRDS(here::here("../output/tables/naesiotus_centralities_regional_df.rds"))
centralities_df_local <- readRDS(here::here("../output/tables/naesiotus_centralities_local_df.rds"))
metaweb <- readRDS(here::here("../output/temporal_output/bipartite_matrix.rds"))
sp_island_id <- readRDS(here::here("../output/temporal_output/sp_island_id.rds"))
species_names_checks.csv <- read.csv("../data/species_names_checks.csv")
```


```{r}
############### island ages

# Sample ages for each species on each island
sp_island_guild_size <- sp_island_id |>
  group_by(island) |>
  mutate(
    sampled_age = runif(n(), min = min_emergence, max = max_emergence)
  ) |>
  ungroup()

# Number of simulations
num_simulations <- 50

# Create the sampled.ages data frame with species_name
sampled.ages <- as.data.frame(
  replicate(num_simulations, {
    sp_island_guild_size |>
      group_by(island) |>
      mutate(
        sampled_age = runif(n(), min = min_emergence, max = max_emergence)
      ) |>
      ungroup() |>
      pull(sampled_age)
  })
)

# Rename columns for simulations
names(sampled.ages) <- paste0("island_age_", 1:num_simulations)

# Add the species_name column
sampled.ages$species_name <- sp_island_guild_size$species_name

# Reorder columns to have species_name first (optional)
sampled.ages <- sampled.ages |>
  dplyr::relocate(species_name)
               
sampled.ages

```


Regional

```{r}
centralities_df_regional <- centralities_df_regional |>
  dplyr::left_join(
    species_names_checks.csv |>
      dplyr::select(annotated_name, species_name) |>
      dplyr::rename(species = annotated_name),  # Rename for joining
    by = "species"
  )

# Filter sp_island_id to retain only species in centralities_df_regional
filtered_sp_island_id_regional <- sp_island_id |>
  dplyr::filter(species_name %in% centralities_df_regional$species_name)

# Add the guild_size column
sp_island_guild_size_regional <- filtered_sp_island_id_regional |>
  dplyr::group_by(island) |>
  dplyr::mutate(guild_size = n_distinct(species_name)) |>
  dplyr::ungroup()

# Add PCA_centrality to sp_island_guild_size_regional
sp_island_guild_size_regional <- sp_island_guild_size_regional |>
  dplyr::left_join(
    centralities_df_regional,
    by = "species_name"
  )

```


```{r}
centralities_df_local <- centralities_df_local |>
  dplyr::left_join(
    species_names_checks.csv |>
      dplyr::select(annotated_name, species_name) |>
      dplyr::rename(species = annotated_name),  # Rename for joining
    by = "species"
  )

# Filter sp_island_id to retain only species in centralities_df_regional
filtered_sp_island_id_local <- sp_island_id |>
  dplyr::filter(species_name %in% centralities_df_local$species_name)

# Add the guild_size column
sp_island_guild_size_local <- filtered_sp_island_id_local |>
  dplyr::group_by(island) |>
  dplyr::mutate(guild_size = n_distinct(species_name)) |>
  dplyr::ungroup()

# Add PCA_centrality to sp_island_guild_size_regional
sp_island_guild_size_local <- sp_island_guild_size_local |>
  dplyr::left_join(
    centralities_df_local%>% select(-island),
    by = "species_name"
  )

```




Add species names to dataset sampled_rates
!!!!

```{r}
# REGIONAL

# # Filter sampled_rates to keep only species present in sp_island_guild_size_regional
# filtered_sampled_rates <- sampled.rates |>
#   dplyr::filter(species_name %in% sp_island_guild_size_regional$species_name)
# 
# # Local
# filtered_sampled_rates_local <- sampled.rates |>
#   dplyr::filter(species_name %in% sp_island_guild_size_local$species_name)

```




```{r}
# Regional

# Add sampled rates to sp_island_guild_size_regional
sp_island_guild_size_regional <- sp_island_guild_size_regional |>
  dplyr::left_join(
    sampled.rates,
    by = "species_name"
  )

sp_island_guild_size_regional


# Local

# Add sampled rates to sp_island_guild_size_regional
sp_island_guild_size_local <- sp_island_guild_size_local |>
  dplyr::left_join(
    sampled.rates,
    by = "species_name"
  )

sp_island_guild_size_local
```

Add island ages sampled

```{r}

# regional

# Add sampled.ages to sp_island_guild_size_regional
sp_island_guild_size_regional <- sp_island_guild_size_regional |>
  dplyr::left_join(
    sampled.ages,
    by = "species_name"
  )

sp_island_guild_size_regional


# local

# Add sampled.ages to sp_island_guild_size_regional
sp_island_guild_size_local <- sp_island_guild_size_local |>
  dplyr::left_join(
    sampled.ages,
    by = "species_name"
  )

sp_island_guild_size_local

```


Gardner island merged with Espanola

```{r}
sp_island_guild_size_regional <- sp_island_guild_size_regional %>%
  mutate(island = ifelse(island == "Gardner", "Espanola", island))

sp_island_guild_size_local <- sp_island_guild_size_local %>%
  mutate(island = ifelse(island == "Gardner", "Espanola", island))
```






```{r}
missing_species <- setdiff(sp_island_guild_size_regional$species_name, CladsOutput$tree$tip.label)
if (length(missing_species) > 0) {
    print("These species are in the dataset but not in the tree:")
    print(missing_species)
}

```
Remove species not in the tree

```{r}
# regional

# sp_island_guild_size_regional_filtered <- sp_island_guild_size_regional |>
#     dplyr::filter(species_name %in% CladsOutput$tree$tip.label)
# 
#  Ainv <- inverseA(CladsOutput$tree)$Ainv
# 
# sp_island_guild_size_regional_filtered <- sp_island_guild_size_regional_filtered |>
#     dplyr::filter(species_name %in% rownames(Ainv))
# 
# 
# # local 
# 
# sp_island_guild_size_local_filtered <- sp_island_guild_size_local |>
#     dplyr::filter(species_name %in% CladsOutput$tree$tip.label)
# 
# sp_island_guild_size_local_filtered <- sp_island_guild_size_local_filtered |>
#     dplyr::filter(species_name %in% rownames(Ainv))
# 
# all(sp_island_guild_size_regional_filtered$species_name %in% rownames(Ainv))
```

Match levels in ginverse

```{r}
# Ainv <- inverseA(CladsOutput$tree)$Ainv

```


## CHECKS


Regional

```{r}

# colnames(Ainv)  # Should return NULL
# colnames(Ainv) <- rownames(Ainv)
# common_species <- intersect(sp_island_guild_size_regional_filtered$species_name, rownames(Ainv))
# Ainv_filtered <- Ainv[common_species, common_species]
# 
# dim(Ainv_filtered)  # Should match the number of species in common_species
# length(common_species)  # Should be equal to the number of species in filtered Ainv
# 
# ginverse <- list(species_name = Ainv_filtered)
# 
# setdiff(sp_island_guild_size_regional_filtered$species_name, common_species)
# all(sp_island_guild_size_regional_filtered$species_name %in% rownames(Ainv_filtered))
```


Local

```{r}
# common_species <- intersect(sp_island_guild_size_local_filtered$species_name, rownames(Ainv))
# colnames(Ainv) <- rownames(Ainv)
# Ainv_filtered <- Ainv[common_species, common_species]
# 
# dim(Ainv_filtered)  # Should match the number of species in common_species
# length(common_species)  # Should be equal to the number of species in filtered Ainv
# 
# ginverse <- list(species_name = Ainv_filtered)
# 
# setdiff(sp_island_guild_size_local_filtered$species_name, common_species)
# all(sp_island_guild_size_local_filtered$species_name %in% rownames(Ainv_filtered))
```

Regional

```{r}
# sp_island_guild_size_regional_filtered$species_name <- factor(
#     sp_island_guild_size_regional_filtered$species_name,
#     levels = rownames(ginverse$species_name)
# )
# 
# setdiff(levels(sp_island_guild_size_regional_filtered$species_name), rownames(ginverse$species_name))
# setdiff(rownames(ginverse$species_name), levels(sp_island_guild_size_regional_filtered$species_name))
# 
# anyDuplicated(levels(sp_island_guild_size_regional_filtered$species_name))
# anyDuplicated(rownames(ginverse$species_name))
# 
# all(sp_island_guild_size_regional_filtered$species_name %in% rownames(ginverse$species_name))
# 
# required_columns <- c(
#     "PCA_centrality", paste0("species_speciation_rate_", 1), 
#     paste0("island_age_", 1), "guild_size", "island_area", "species_name"
# )
# missing_columns <- setdiff(required_columns, colnames(sp_island_guild_size_regional_filtered))
# missing_columns

```


Local

```{r}
# sp_island_guild_size_local_filtered$species_name <- factor(
#     sp_island_guild_size_local_filtered$species_name,
#     levels = rownames(ginverse$species_name)
# )
# 
# setdiff(levels(sp_island_guild_size_local_filtered$species_name), rownames(ginverse$species_name))
# setdiff(rownames(ginverse$species_name), levels(sp_island_guild_size_local_filtered$species_name))
# 
# anyDuplicated(levels(sp_island_guild_size_local_filtered$species_name))
# anyDuplicated(rownames(ginverse$species_name))
# 
# all(sp_island_guild_size_local_filtered$species_name %in% rownames(ginverse$species_name))
# 
# required_columns <- c(
#     "PCA_centrality", paste0("species_speciation_rate_", 1), 
#     paste0("island_age_", 1), "guild_size", "island_area", "species_name"
# )
# missing_columns <- setdiff(required_columns, colnames(sp_island_guild_size_local_filtered))
# missing_columns
```




```{r}
sp_island_guild_size_regional <- as.data.frame(sp_island_guild_size_regional)
sp_island_guild_size_local<- as.data.frame(sp_island_guild_size_local)
```



Standarize variables

Z score guild_size and island_area

```{r}
sp_island_guild_size_regional <- sp_island_guild_size_regional %>%
    mutate(
        guild_size_z = scale(guild_size) %>% as.numeric(),
        island_area_z = scale(island_area) %>% as.numeric()
    )


sp_island_guild_size_local <- sp_island_guild_size_local %>%
    mutate(
        guild_size_z = scale(guild_size) %>% as.numeric(),
        island_area_z = scale(island_area) %>% as.numeric()
    )

```

z score island ages and speciation rates (sets) REGIONAL

```{r}
# Identify the columns to z-score
age_columns <- grep("^island_age_", colnames(sp_island_guild_size_regional), value = TRUE)
speciation_rate_columns <- grep("^species_speciation_rate_", colnames(sp_island_guild_size_regional), value = TRUE)

# Z-score the columns and create new ones with a "_z" suffix
sp_island_guild_size_regional <- sp_island_guild_size_regional %>%
  mutate(across(
    all_of(c(age_columns, speciation_rate_columns)),
    ~ scale(.) %>% as.numeric(),
    .names = "{.col}_z"
  ))

sp_island_guild_size_regional
```


Z score speciation rates and centrality by island - LOCAL

```{r}
# Identify the columns to z-score
age_columns <- grep("^island_age_", colnames(sp_island_guild_size_local), value = TRUE)
speciation_rate_columns <- grep("^species_speciation_rate_", colnames(sp_island_guild_size_local), value = TRUE)

# Z-score speciation rates by island
sp_island_guild_size_local <- sp_island_guild_size_local %>%
  group_by(island) %>%
  mutate(across(
    all_of(speciation_rate_columns),
    ~ scale(.) %>% as.numeric(),
    .names = "{.col}_z"
  )) %>%
  ungroup()  # Ungroup after z-scoring speciation rates

# Z-score island ages across the entire dataset
sp_island_guild_size_local <- sp_island_guild_size_local %>%
  mutate(across(
    all_of(age_columns),
    ~ scale(.) %>% as.numeric(),
    .names = "{.col}_z"
  ))


sp_island_guild_size_local
```


### Add mean speciation rates and scale it


```{r}

mean.tip.rates <- mean.tip.rates |>
  dplyr::left_join(
    species_names_checks.csv |>
      dplyr::select(annotated_name, species_name) |>
      dplyr::rename(species = annotated_name),  # Rename for joining
    by = "species"
  )

# regional

sp_island_guild_size_regional_meanrates <- sp_island_guild_size_regional %>%
  left_join(mean.tip.rates %>% select(species_name, lambda.clads), by = "species_name") %>%
  relocate(lambda.clads, .after = species_name) 

sp_island_guild_size_regional_meanrates <- sp_island_guild_size_regional_meanrates %>%
    mutate(
        lambda.clads_z = scale(lambda.clads) %>% as.numeric()
    )



# local

sp_island_guild_size_local_meanrates <- sp_island_guild_size_local %>%
  left_join(mean.tip.rates %>% select(species_name, lambda.clads), by = "species_name") %>%
  relocate(lambda.clads, .after = species_name) 


sp_island_guild_size_local_meanrates <- sp_island_guild_size_local_meanrates %>%
    mutate(
        lambda.clads_z = scale(lambda.clads) %>% as.numeric()
    )


```




### Add lineage ages


```{r}
## Replacing "." in species names for CLaDS not to complain
tree$tip.label <- gsub("-", ".", 
                                   gsub("_", " ", CladsOutput$tree$tip.label))
# Get the root age (maximum cumulative branch length to tips)
root_age <- max(node.depth.edgelength(tree))

# Calculate ages for each tip relative to the root
lineage_ages <- sapply(tree$tip.label, function(tip) {
  # Get the node number of the tip
  tip_node <- which(tree$tip.label == tip)
  
  # Calculate the cumulative branch length from root to the tip
  tip_age <- root_age - node.depth.edgelength(tree)[tip_node]
  return(tip_age)
})

# Combine tip labels with their ages
lineage_ages_df <- data.frame(
  lineage = tree$tip.label,
  age = lineage_ages*1e10
)

# View the lineage ages
lineage_ages_df
```

Homogenize names ages

```{r}


lineage_ages_df$species_name <- lineage_ages_df$lineage  # Temporary assignment

####################################

# Step 3: Add lineage ages to the regional dataset
sp_island_guild_size_regional_meanrates <- sp_island_guild_size_regional_meanrates %>%
  left_join(
    lineage_ages_df %>% select(species_name, age),
    by = "species_name"
  ) |> 
      mutate(
        age_z = scale(age) %>% as.numeric(),
    )

# Step 4: Add lineage ages to the local dataset
sp_island_guild_size_local_meanrates <- sp_island_guild_size_local_meanrates %>%
  left_join(
    lineage_ages_df %>% select(species_name, age),
    by = "species_name"
  ) |> 
      mutate(
        age_z = scale(age) %>% as.numeric(),
    )

# Step 5: Verify if all species now have ages
missing_ages_regional <- sp_island_guild_size_regional_meanrates %>% filter(is.na(age))
missing_ages_local <- sp_island_guild_size_local_meanrates %>% filter(is.na(age))

if (nrow(missing_ages_regional) > 0) {
  print("These species in the regional dataset still have missing ages:")
  print(missing_ages_regional$species_name)
}

if (nrow(missing_ages_local) > 0) {
  print("These species in the local dataset still have missing ages:")
  print(missing_ages_local$species_name)
}


```

The species above are also those for which I have interaction data but not phylogenetic - most of them are extinct, or in some cases had conflictive positions in the phylogeny (sp_nov_CA and pinzonopsis).


We will use minimum emergence for island age. It needs to be scaled:

```{r}

sp_island_guild_size_regional_meanrates <- sp_island_guild_size_regional_meanrates %>%
      mutate(
        min_emergence_z = scale(min_emergence) %>% as.numeric(),
    )

sp_island_guild_size_local_meanrates <- sp_island_guild_size_local_meanrates %>%
      mutate(
        min_emergence_z = scale(min_emergence) %>% as.numeric(),
    )
```


## create phylogenetic distance matrix

```{r}
Ainv <- inverseA(CladsOutput$tree)$Ainv

# Get current rownames of Ainv
current_ainv_names <- rownames(Ainv)

# Replace names only if they exist in the annotated_name column, otherwise keep original
corrected_names <- current_ainv_names

# Check which names need to be replaced
matched_indices <- match(current_ainv_names, species_names_checks.csv$annotated_name)

# Update only the matched names
corrected_names[!is.na(matched_indices)] <- species_names_checks.csv$species_name[matched_indices[!is.na(matched_indices)]]

# Apply corrected names to the Ainv matrix
rownames(Ainv) <- corrected_names

# Print the updated rownames for verification
print(rownames(Ainv))

colnames(Ainv) <- rownames(Ainv)

## filter the matrices for the species present in the datasets

# Filter Ainv for regional species
regional_species <- sp_island_guild_size_regional_meanrates$species_name
Ainv_regional <- Ainv[rownames(Ainv) %in% regional_species, colnames(Ainv) %in% regional_species]

# Filter Ainv for local species
local_species <- sp_island_guild_size_local_meanrates$species_name
Ainv_local <- Ainv[rownames(Ainv) %in% local_species, colnames(Ainv) %in% local_species]

```


------------------------------------------------------------------------------


Model 1: Does speciation rates influence centrality?

1.1
PCA ~ sampled spec. rate
1.2
PCA ~ mean spec. rate

Model 2: Does speciation rates influence specialization?

d ~ mean spec. rate

Model 3: Does lineage age influence centrality?

PCA ~ lineage age

Model 4: Does lineage age influence specialization?

d ~ lineage age




* Min emergence for island age (it is usually used the min. emergence, not an estimate between min and max)
------------------------------------------------------------------------------


# Models regional

## MODEL 1.1 - sampled speciation rates on centrality

Version accounting for variability in speciation rates:

```{r}

my_priors <- list(
  R = list(V = 1/2, nu = 0.002),
  G = list(G1 = list(V = 1/2, nu = 0.002))
)

# Initialize an empty list to store model fits
full_model.fit <- vector(mode = "list", length = 50)


for (i in 1:50) {
  # Construct the formula
  formula_string <- paste0(
    "PCA_centrality ~ species_speciation_rate_", i, "_z + ", 
    "min_emergence_z + island_area_z + guild_size_z"
  )
  
  # Define the relevant columns to check for missing values
  predictors <- c(
    paste0("species_speciation_rate_", i, "_z"),
    "min_emergence_z",
    "island_area_z",
    "guild_size_z",
    "lambda.clads_z"
  )
  
  # Filter out rows with NA values in the relevant columns
  data_filtered <- sp_island_guild_size_regional_meanrates |>
    dplyr::filter(across(all_of(predictors), ~ !is.na(.)))

  # Fit the model
  full_model.fit[[i]] <- MCMCglmm(
    as.formula(formula_string),  # Use as.formula() directly
    random = ~ species_name,  # Random effect for species_name
    family = "gaussian",  # Assuming a Gaussian response for PCA centrality
    ginverse = list(species_name = Ainv_regional),  # Use species_name in ginverse
    data = as.data.frame(data_filtered),  # Dataframe
    prior = my_priors,  # Priors
    nitt = 1000000,  # Number of iterations
    burnin = 500000,  # Burn-in
    thin = 100  # Thinning
  )
}



saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/model1_sampled_spec_centrality_regional.rds"))

```


## MODEL 1.2 - mean speciation rates on centrality

```{r}

my_priors <- list(
  R = list(V = 1/2, nu = 0.002),
  G = list(G1 = list(V = 1/2, nu = 0.002))
)

  # Construct the formula
  formula_string <- paste0(
    "PCA_centrality ~ lambda.clads_z + min_emergence_z + island_area_z + guild_size_z"
  )
  
  # Define the relevant columns to check for missing values
  predictors <- c(
    paste0("lambda.clads_z"),
    "min_emergence_z",
    "island_area_z",
    "guild_size_z"
  )
  
  # Filter out rows with NA values in the relevant columns
  data_filtered <- sp_island_guild_size_regional_meanrates |>
    dplyr::filter(across(all_of(predictors), ~ !is.na(.)))

  # Fit the model
  full_model.fit <- MCMCglmm(
    as.formula(formula_string),  # Use as.formula() directly
    random = ~ species_name,  # Random effect for species_name
    family = "gaussian",  # Assuming a Gaussian response for PCA centrality
    ginverse = list(species_name = Ainv_regional),  # Use species_name in ginverse
    data = as.data.frame(data_filtered),  # Dataframe
    prior = my_priors,  # Priors
    nitt = 1000000,  # Number of iterations
    burnin = 500000,  # Burn-in
    thin = 100  # Thinning
  )




saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/model1.2_mean_spec_centrality_regional.rds"))

```

## MODEL 2 - mean speciation on specialization

```{r}

my_priors <- list(
  R = list(V = 1/2, nu = 0.002),
  G = list(G1 = list(V = 1/2, nu = 0.002))
)

  # Construct the formula
  formula_string <- paste0(
    "d_z ~ lambda.clads_z + min_emergence_z + island_area_z + guild_size_z"
  )
  
  # Define the relevant columns to check for missing values
  predictors <- c(
    paste0("lambda.clads_z"),
    "min_emergence_z",
    "island_area_z",
    "guild_size_z"
  )
  
  # Filter out rows with NA values in the relevant columns
  data_filtered <- sp_island_guild_size_regional_meanrates |>
    dplyr::filter(across(all_of(predictors), ~ !is.na(.)))

  # Fit the model
  full_model.fit <- MCMCglmm(
    as.formula(formula_string),  # Use as.formula() directly
    random = ~ species_name,  # Random effect for species_name
    family = "gaussian",  # Assuming a Gaussian response for PCA centrality
    ginverse = list(species_name = Ainv_regional),  # Use species_name in ginverse
    data = as.data.frame(data_filtered),  # Dataframe
    prior = my_priors,  # Priors
    nitt = 1000000,  # Number of iterations
    burnin = 500000,  # Burn-in
    thin = 100  # Thinning
  )




saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/model2_meanrate_d_regional.rds"))

```


## MODEL 3 - lineage age on centrality

```{r}

my_priors <- list(
  R = list(V = 1/2, nu = 0.002),
  G = list(G1 = list(V = 1/2, nu = 0.002))
)

  # Construct the formula
  formula_string <- paste0(
    "PCA_centrality ~ age_z + min_emergence_z + island_area_z + guild_size_z"
  )
  
  # Define the relevant columns to check for missing values
  predictors <- c(
    paste0("age_z"),
    "min_emergence_z",
    "island_area_z",
    "guild_size_z"
  )
  
  # Filter out rows with NA values in the relevant columns
  data_filtered <- sp_island_guild_size_regional_meanrates |>
    dplyr::filter(across(all_of(predictors), ~ !is.na(.)))

  # Fit the model
  full_model.fit <- MCMCglmm(
    as.formula(formula_string),  # Use as.formula() directly
    random = ~ species_name,  # Random effect for species_name
    family = "gaussian",  # Assuming a Gaussian response for PCA centrality
    ginverse = list(species_name = Ainv_regional),  # Use species_name in ginverse
    data = as.data.frame(data_filtered),  # Dataframe
    prior = my_priors,  # Priors
    nitt = 1000000,  # Number of iterations
    burnin = 500000,  # Burn-in
    thin = 100  # Thinning
  )




saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/model3_age_centrality_regional.rds"))

```



## MODEL 4 - lineage age on specialization

```{r}

my_priors <- list(
  R = list(V = 1/2, nu = 0.002),
  G = list(G1 = list(V = 1/2, nu = 0.002))
)

  # Construct the formula
  formula_string <- paste0(
    "d_z ~ age_z + min_emergence_z + island_area_z + guild_size_z"
  )
  
  # Define the relevant columns to check for missing values
  predictors <- c(
    paste0("age_z"),
    "min_emergence_z",
    "island_area_z",
    "guild_size_z"
  )
  
  # Filter out rows with NA values in the relevant columns
  data_filtered <- sp_island_guild_size_regional_meanrates |>
    dplyr::filter(across(all_of(predictors), ~ !is.na(.)))

  # Fit the model
  full_model.fit <- MCMCglmm(
    as.formula(formula_string),  # Use as.formula() directly
    random = ~ species_name,  # Random effect for species_name
    family = "gaussian",  # Assuming a Gaussian response for PCA centrality
    ginverse = list(species_name = Ainv_regional),  # Use species_name in ginverse
    data = as.data.frame(data_filtered),  # Dataframe
    prior = my_priors,  # Priors
    nitt = 1000000,  # Number of iterations
    burnin = 500000,  # Burn-in
    thin = 100  # Thinning
  )




saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/model4_age_d_regional.rds"))

```







# Models lOCAL

## MODEL 1.1 - sampled speciation rates on centrality

```{r}

my_priors <- list(
  R = list(V = 1/2, nu = 0.002),
  G = list(G1 = list(V = 1/2, nu = 0.002))
)

# Initialize an empty list to store model fits
full_model.fit <- vector(mode = "list", length = 50)


for (i in 1:50) {
  # Construct the formula
  formula_string <- paste0(
    "PCA_centrality ~ species_speciation_rate_", i, "_z + ", 
    "min_emergence_z + island_area_z + guild_size_z"
  )
  
  # Define the relevant columns to check for missing values
  predictors <- c(
    paste0("species_speciation_rate_", i, "_z"),
    "min_emergence_z",
    "island_area_z",
    "guild_size_z",
    "lambda.clads_z"
  )
  
  # Filter out rows with NA values in the relevant columns
  data_filtered <- sp_island_guild_size_local_meanrates |>
    dplyr::filter(across(all_of(predictors), ~ !is.na(.)))

  # Fit the model
  full_model.fit[[i]] <- MCMCglmm(
    as.formula(formula_string),  # Use as.formula() directly
    random = ~ species_name,  # Random effect for species_name
    family = "gaussian",  # Assuming a Gaussian response for PCA centrality
    ginverse = list(species_name = Ainv_local),  # Use species_name in ginverse
    data = as.data.frame(data_filtered),  # Dataframe
    prior = my_priors,  # Priors
    nitt = 1000000,  # Number of iterations
    burnin = 500000,  # Burn-in
    thin = 100  # Thinning
  )
}



saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/model1_sampled_spec_centrality_local.rds"))

```



## MODEL 1.2 - mean speciation rates on centrality


```{r}

my_priors <- list(
  R = list(V = 1/2, nu = 0.002),
  G = list(G1 = list(V = 1/2, nu = 0.002))
)

  # Construct the formula
  formula_string <- paste0(
    "PCA_centrality ~ lambda.clads_z + min_emergence_z + island_area_z + guild_size_z"
  )
  
  # Define the relevant columns to check for missing values
  predictors <- c(
    paste0("lambda.clads_z"),
    "min_emergence_z",
    "island_area_z",
    "guild_size_z"
  )
  
  # Filter out rows with NA values in the relevant columns
  data_filtered <- sp_island_guild_size_local_meanrates |>
    dplyr::filter(across(all_of(predictors), ~ !is.na(.)))

  # Fit the model
  full_model.fit <- MCMCglmm(
    as.formula(formula_string),  # Use as.formula() directly
    random = ~ species_name,  # Random effect for species_name
    family = "gaussian",  # Assuming a Gaussian response for PCA centrality
    ginverse = list(species_name = Ainv_local),  # Use species_name in ginverse
    data = as.data.frame(data_filtered),  # Dataframe
    prior = my_priors,  # Priors
    nitt = 1000000,  # Number of iterations
    burnin = 500000,  # Burn-in
    thin = 100  # Thinning
  )




saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/model1.2_mean_spec_centrality_local.rds"))

```



## MODEL 2 - mean speciation on specialization

```{r}

my_priors <- list(
  R = list(V = 1/2, nu = 0.002),
  G = list(G1 = list(V = 1/2, nu = 0.002))
)

  # Construct the formula
  formula_string <- paste0(
    "d_z ~ lambda.clads_z + min_emergence_z + island_area_z + guild_size_z"
  )
  
  # Define the relevant columns to check for missing values
  predictors <- c(
    paste0("lambda.clads_z"),
    "min_emergence_z",
    "island_area_z",
    "guild_size_z"
  )
  
  # Filter out rows with NA values in the relevant columns
  data_filtered <- sp_island_guild_size_local_meanrates |>
    dplyr::filter(across(all_of(predictors), ~ !is.na(.)))

  # Fit the model
  full_model.fit <- MCMCglmm(
    as.formula(formula_string),  # Use as.formula() directly
    random = ~ species_name,  # Random effect for species_name
    family = "gaussian",  # Assuming a Gaussian response for PCA centrality
    ginverse = list(species_name = Ainv_local),  # Use species_name in ginverse
    data = as.data.frame(data_filtered),  # Dataframe
    prior = my_priors,  # Priors
    nitt = 1000000,  # Number of iterations
    burnin = 500000,  # Burn-in
    thin = 100  # Thinning
  )




saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/model2_meanrate_d_local.rds"))

```




## MODEL 3 - lineage age on centrality

```{r}

my_priors <- list(
  R = list(V = 1/2, nu = 0.002),
  G = list(G1 = list(V = 1/2, nu = 0.002))
)

  # Construct the formula
  formula_string <- paste0(
    "PCA_centrality ~ age_z + min_emergence_z + island_area_z + guild_size_z"
  )
  
  # Define the relevant columns to check for missing values
  predictors <- c(
    paste0("age_z"),
    "min_emergence_z",
    "island_area_z",
    "guild_size_z"
  )
  
  # Filter out rows with NA values in the relevant columns
  data_filtered <- sp_island_guild_size_local_meanrates |>
    dplyr::filter(across(all_of(predictors), ~ !is.na(.)))

  # Fit the model
  full_model.fit <- MCMCglmm(
    as.formula(formula_string),  # Use as.formula() directly
    random = ~ species_name,  # Random effect for species_name
    family = "gaussian",  # Assuming a Gaussian response for PCA centrality
    ginverse = list(species_name = Ainv_local),  # Use species_name in ginverse
    data = as.data.frame(data_filtered),  # Dataframe
    prior = my_priors,  # Priors
    nitt = 1000000,  # Number of iterations
    burnin = 500000,  # Burn-in
    thin = 100  # Thinning
  )




saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/model3_age_centrality_local.rds"))

```



## MODEL 4 - lineage age on specialization

```{r}

my_priors <- list(
  R = list(V = 1/2, nu = 0.002),
  G = list(G1 = list(V = 1/2, nu = 0.002))
)

  # Construct the formula
  formula_string <- paste0(
    "d_z ~ age_z + min_emergence_z + island_area_z + guild_size_z"
  )
  
  # Define the relevant columns to check for missing values
  predictors <- c(
    paste0("age_z"),
    "min_emergence_z",
    "island_area_z",
    "guild_size_z"
  )
  
  # Filter out rows with NA values in the relevant columns
  data_filtered <- sp_island_guild_size_local_meanrates |>
    dplyr::filter(across(all_of(predictors), ~ !is.na(.)))

  # Fit the model
  full_model.fit <- MCMCglmm(
    as.formula(formula_string),  # Use as.formula() directly
    random = ~ species_name,  # Random effect for species_name
    family = "gaussian",  # Assuming a Gaussian response for PCA centrality
    ginverse = list(species_name = Ainv_local),  # Use species_name in ginverse
    data = as.data.frame(data_filtered),  # Dataframe
    prior = my_priors,  # Priors
    nitt = 1000000,  # Number of iterations
    burnin = 500000,  # Burn-in
    thin = 100  # Thinning
  )




saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/model4_age_d_local.rds"))

```




```{r}
# REGIONAL

#model.fit <- readRDS(here::here("../output/models/network_role_speciation/model1_sampled_spec_centrality_regional.rds"))
#model.fit <- readRDS(here::here("../output/models/network_role_speciation/model1.2_mean_spec_centrality_regional.rds"))
#model.fit <- readRDS(here::here("../output/models/network_role_speciation/model2_meanrate_d_regional.rds"))
#model.fit <- readRDS(here::here("../output/models/network_role_speciation/model3_age_centrality_regional.rds"))
#model.fit <- readRDS(here::here("../output/models/network_role_speciation/model4_age_d_regional.rds"))

# LOCAL

#model.fit <- readRDS(here::here("../output/models/network_role_speciation/model1_sampled_spec_centrality_local.rds"))
#model.fit <- readRDS(here::here("../output/models/network_role_speciation/model1.2_mean_spec_centrality_local.rds"))
model.fit <- readRDS(here::here("../output/models/network_role_speciation/model2_meanrate_d_local.rds"))
#model.fit <- readRDS(here::here("../output/models/network_role_speciation/model3_age_centrality_local.rds"))
#model.fit <- readRDS(here::here("../output/models/network_role_speciation/model4_age_d_local.rds"))



```



# Model results

For sampled speciation

```{r}
# Extract fixed effects (Sol) and variance components (VCV)
fixed_effects <- lapply(model.fit, function(model) as.mcmc(model$Sol))
variance_components <- lapply(model.fit, function(model) as.mcmc(model$VCV))

# Rename the columns to standard names
fixed_effects_standardized <- lapply(seq_along(fixed_effects), function(i) {
  colnames(fixed_effects[[i]]) <- c("(Intercept)", "species_speciation_rate", 
                                    "island_age", "island_area","guild_size")
  fixed_effects[[i]]
})


## combine chains

combined_fixed <- do.call(mcmc.list, fixed_effects_standardized)
```


## CHAINS CONVERGENCE

#### Traceplot

```{r}
traceplot(combined_fixed)
```

The traceplots show horizontal bands without trends or drifts, and they overlap and mix well, indicating that chains converged.

#### Autocorrelation plot

```{r}
autocorr.plot(combined_fixed)
```

There is a fast decay to zero in all cases, indicating that samples become effectively independent quickly, and there is no autocorrelation detected.

#### Gelman-Rubin diagnostic

```{r}
gelman_fixed <- gelman.diag(combined_fixed)
print(gelman_fixed)
```
The Gelman-Rubin diagnostic, also called the potential scale reduction factor (PSRF), is a method to assess the convergence of Markov Chain Monte Carlo (MCMC) simulations. It compares the variance within each chain to the variance between chains to determine whether the chains have converged to the same target distribution.

Interpretation:
Point Estimate: The closer this value is to 1, the better the convergence. Values above 1.1 suggest that the chains have not fully converged.
Upper C.I.: The upper bound of the confidence interval for the PSRF. This helps account for uncertainty in the diagnostic. Values significantly above 1.1 may indicate poor convergence.

Convergence is problematic for species_speciation_rate.


## Effective Sample Size

```{r}
effectiveSize(model.fit[[7]]$Sol[,2])
```


## Posterior distributions

```{r}
# Summarize fixed effects across all models
summary_fixed <- summary(combined_fixed)

print(summary_fixed)
```



```{r}
# Combine the posterior distributions into one long data frame
# Combine the posterior distributions into one long data frame
posterior_long <- do.call(rbind, lapply(seq_along(fixed_effects_standardized), function(i) {
  as.data.frame(as.matrix(fixed_effects_standardized[[i]])) %>%
    mutate(Model = paste0("Model_", i)) %>%
    pivot_longer(cols = -Model, names_to = "Predictor", values_to = "Value")  # Exclude Model from pivoting
}))

# Define new labels and desired order for predictors
predictor_labels <- c(
  "species_speciation_rate" = "Speciation rate",
  "island_age" = "Island age",
  "island_area" = "Island area",
  "guild_size" = "N Island species",
  "(Intercept)" = "(Intercept)"
)

# Set the factor levels to control order in the plot
posterior_long$Predictor <- factor(posterior_long$Predictor, 
                                   levels = c("(Intercept)",
                                              "species_speciation_rate", 
                                              "island_age", 
                                              "island_area", 
                                              "guild_size"
                                              ))

# Plot the posterior distributions with improved predictor labels and order
ggplot(posterior_long, aes(x = Value, color = Model, group = Model)) +
  geom_density(size = 0.5, alpha = 0.8) +  # Add density curves
  facet_wrap(~Predictor, scales = "fixed", ncol = 1, 
             labeller = labeller(Predictor = predictor_labels)) +  # Apply new labels
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +  # Vertical line at 0
  labs(
    x = "Standardized posterior coefficient",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",  # Remove legend for clarity
    panel.grid.major.y = element_blank(),
    strip.text = element_text(size = 12),  # Increase facet label text size
    axis.title.x = element_text(size = 14),  # Increase x-axis label size
    axis.title.y = element_text(size = 14)   # Increase y-axis label size
  )



#ggsave(here::here("../output/figures/posterior_pars/model2_pca_d_local.png"), height = 6, width = 4)
```

------------------------


For mean speciation

```{r}
fixed_effects <- as.mcmc(model.fit$Sol)
variance_components <- as.mcmc(model.fit$VCV)

colnames(fixed_effects) <- c("(Intercept)", "species_speciation_rate", 
                             "island_age", "island_area", "guild_size")

# Summarize fixed effects for the single model
summary_fixed <- summary(fixed_effects)

# Convert the fixed effects to a data frame for plotting
posterior_long <- as.data.frame(as.matrix(fixed_effects)) %>%
  pivot_longer(cols = everything(), names_to = "Predictor", values_to = "Value") %>%
  mutate(Model = "Single Model")  # Add a column to distinguish the model

# Define new labels and desired order for predictors
predictor_labels <- c(
  "species_speciation_rate" = "Mean speciation rate",
  "island_age" = "Island age",
  "island_area" = "Island area",
  "guild_size" = "N Island species",
  "(Intercept)" = "(Intercept)"
)

# Set the factor levels to control order in the plot
posterior_long$Predictor <- factor(posterior_long$Predictor, 
                                   levels = c("(Intercept)",
                                              "species_speciation_rate", 
                                              "island_age", 
                                              "island_area", 
                                              "guild_size"))

ggplot(posterior_long, aes(x = Value)) +
  geom_density(fill = "#b3b3b3", color = "#b3b3b3", alpha = 0.8, size = 0.8) +  # Add density curves with custom color
  facet_wrap(~Predictor, scales = "fixed", ncol = 1, 
             labeller = labeller(Predictor = predictor_labels)) +  # Apply new labels
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +  # Vertical line at 0
  labs(
    x = "Standardized posterior coefficient",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",  # Remove legend for clarity
    panel.grid.major.y = element_blank(),
    strip.text = element_text(size = 12),  # Increase facet label text size
    axis.title.x = element_text(size = 14),  # Increase x-axis label size
    axis.title.y = element_text(size = 14)   # Increase y-axis label size
  )


ggsave(here::here("../output/figures/posterior_pars/model2_pca_d_local.png"), height = 6, width = 4)

```



