---
title: "Analysis Centrality ~ speciation"

---


```{r}
library("here")
library("BAMMtools")
library("RPANDA")
#library("ggtree")
library("MetBrewer")
library("patchwork")
library("doParallel")
library("foreach")
library("MCMCglmm")
library(dplyr)
library(coda)
library(ggdist)
library(tidyverse)

here::i_am("centrality_speciation_regional.Rmd")
```


```{r}
## Importing tree
tree <- read.tree(here::here("clads/cleannames_tree_041024.txt"))

## Rescaling branch lengths by a scale of 10 to reflect time
tree$edge.length <- tree$edge.length * 10

## Making sure the tree is still ultrametric
is.ultrametric(tree)

## Replacing "." in species names for CLaDS not to complain
tree$tip.label <- gsub(".", "-", tree$tip.label, fixed = TRUE)

## Writing tree to file
write.tree(tree, here::here("clads/naesiotus_rescaled.tre"))
write.tree(tree, here::here("bamm/naesiotus_rescaled.tre"))

load(here::here("clads/naesiotus_clads_rates.RData"))

bamm.data <- getEventData(tree, here::here("bamm/event_data.txt"), burnin = 0.1)
```



distribution of trait states in the tips and speciation rates. The colors of the symbols at the tips describe the trait state, while both the icon size and the plot to the right show the distribution of speciation rates estimated using CLaDS:

```{r}
plotRateThroughTime(bamm.data, intervalCol = "lightgrey")

css <- credibleShiftSet(bamm.data, 1)
plot(css, lwd = 3)

tip.rates.bamm <- getTipRates(bamm.data, returnNetDiv = FALSE)
```



distribution of trait states in the tips and speciation rates. The colors of the symbols at the tips describe the trait state, while both the icon size and the plot to the right show the distribution of speciation rates estimated using CLaDS:

```{r}
plot_ClaDS_phylo(CladsOutput$tree, CladsOutput$lambdai_map, lwd = 3)
hist(CladsOutput$lambdatip_map)
```

distribution of trait states in the tips and speciation rates. The colors of the symbols at the tips describe the trait state, while both the icon size and the plot to the right show the distribution of speciation rates estimated using CLaDS:

```{r}
mean.tip.rates <- data.frame(
    species = tree$tip.label,
    lambda.bamm = round(setNames(apply(tip.rates.bamm[[1]], 1, mean), NULL), 5),
    mu.bamm = round(setNames(apply(tip.rates.bamm[[2]], 1, mean), NULL), 5),
    lambda.clads = round(CladsOutput$lambdatip_map, 5))

#write.table(mean.tip.rates, file = here::here("output/mean_tip_rates.csv"), sep = ",", quote = FALSE, row.names = FALSE)
```



# MCMCglmm


```{r}

## Branch indices
br.ind <- 4:(4+144)

## Tip indices
tip.ind <- (max(br.ind)+1):(max(br.ind)+Ntip(CladsOutput$tree))

## DTT indices
dtt.ind <- (max(tip.ind)+2):length(CladsOutput$chains[[1]])


## Checking parameter values
samp.ind <- sample(1:length(tip.ind), 1)
d_1 <- density(sapply(CladsOutput$chains,"[[", tip.ind[samp.ind])[251:1000,])
hist(sapply(CladsOutput$chains,"[[", tip.ind[samp.ind])[251:1000,], freq = FALSE, ylim = c(0, 1), breaks = 50)
lines(d_1)
abline(v = d_1$x[which.max(d_1$y)], col = 2, lwd = 3)
abline(v = CladsOutput$lambdatip_map[samp.ind], col = 3, lwd = 3)
legend("topright", legend = c("MAP", "lambda_tip"), lwd = 3, col = c(2, 3), cex = 2)
```



## compute dataset

```{r}
CladsOutput$tree$tip.label <- gsub("-", ".", 
                                   gsub("_", " ", CladsOutput$tree$tip.label))
```

PCA centrality, lineage size, and island area:

```{r}
centralities_df <- readRDS(here::here("../output/tables/centralities_regional_df.rds.rds")) |> dplyr::select(species, PCA_centrality, degree)
metaweb <- readRDS(here::here("../output/temporal_output/bipartite_matrix.rds"))
sp_island_id <- readRDS(here::here("../output/temporal_output/sp_island_id.rds"))
species_names_checks.csv <- read.csv("../data/species_names_checks.csv")

```


Speciation rate and island ages are sampled from estimated distributions:

```{r}

###############  speciation rates

## Extracting posterior distributions for the tips
post.dist.tips <- lapply(1:length(tip.ind), function(x){sapply(CladsOutput$chains, "[[", tip.ind[x])})

## Discarding burnin
post.dist.tips <- lapply(post.dist.tips, function(x){x[251:1000,]})

## Creating data.frame with sampled rates for each tip
sampled.rates <- as.data.frame(t(plyr::ldply(1:50, .fun = function(x){sapply(post.dist.tips, function(y){sample(y, 1)})})))
names(sampled.rates) <- paste0("species_speciation_rate_", 1:50)



############### island ages

# Sample ages for each species on each island
sp_island_guild_size <- sp_island_id |>
  group_by(island) |>
  mutate(
    sampled_age = runif(n(), min = min_emergence, max = max_emergence)
  ) |>
  ungroup()

# Number of simulations
num_simulations <- 50

# Create the sampled.ages data frame with species_name
sampled.ages <- as.data.frame(
  replicate(num_simulations, {
    sp_island_guild_size |>
      group_by(island) |>
      mutate(
        sampled_age = runif(n(), min = min_emergence, max = max_emergence)
      ) |>
      ungroup() |>
      pull(sampled_age)
  })
)

# Rename columns for simulations
names(sampled.ages) <- paste0("island_age_", 1:num_simulations)

# Add the species_name column
sampled.ages$species_name <- sp_island_guild_size$species_name

# Reorder columns to have species_name first (optional)
sampled.ages <- sampled.ages |>
  dplyr::relocate(species_name)
               
sampled.ages

```






```{r}
centralities_df <- centralities_df |>
  dplyr::left_join(
    species_names_checks.csv |>
      dplyr::select(annotated_name, species_name) |>
      dplyr::rename(species = annotated_name),  # Rename for joining
    by = "species"
  )

# Filter sp_island_id to retain only species in centralities_df
filtered_sp_island_id <- sp_island_id |>
  dplyr::filter(species_name %in% centralities_df$species_name)

# Add the guild_size column
sp_island_guild_size <- filtered_sp_island_id |>
  dplyr::group_by(island) |>
  dplyr::mutate(guild_size = n_distinct(species_name)) |>
  dplyr::ungroup()

# Add PCA_centrality to sp_island_guild_size
sp_island_guild_size <- sp_island_guild_size |>
  dplyr::left_join(
    centralities_df |>
      dplyr::select(species_name, PCA_centrality, degree),
    by = "species_name"
  )

```


Add species names to dataset sampled_rates

```{r}
# Assuming CladsOutput$tree$tip.label has species names in the same order as sampled_rates rows
species_order <- CladsOutput$tree$tip.label
# Add species names to sampled_rates
sampled.rates$species_name <- species_order

# Filter sampled_rates to keep only species present in sp_island_guild_size
filtered_sampled_rates <- sampled.rates |>
  dplyr::filter(species_name %in% sp_island_guild_size$species_name)


```



```{r}
# Add sampled rates to sp_island_guild_size
sp_island_guild_size <- sp_island_guild_size |>
  dplyr::left_join(
    filtered_sampled_rates,
    by = "species_name"
  )

sp_island_guild_size
```

Add island ages sampled

```{r}
# Add sampled.ages to sp_island_guild_size
sp_island_guild_size <- sp_island_guild_size |>
  dplyr::left_join(
    sampled.ages,
    by = "species_name"
  )

sp_island_guild_size

```



```{r}
missing_species <- setdiff(sp_island_guild_size$species_name, CladsOutput$tree$tip.label)
if (length(missing_species) > 0) {
    print("These species are in the dataset but not in the tree:")
    print(missing_species)
}

```
Remove species not in the tree

```{r}
sp_island_guild_size_filtered <- sp_island_guild_size |>
    dplyr::filter(species_name %in% CladsOutput$tree$tip.label)

 Ainv <- inverseA(CladsOutput$tree)$Ainv

sp_island_guild_size_filtered <- sp_island_guild_size_filtered |>
    dplyr::filter(species_name %in% rownames(Ainv))

all(sp_island_guild_size_filtered$species_name %in% rownames(Ainv))
```

Match levels in ginverse

```{r}
 Ainv <- inverseA(CladsOutput$tree)$Ainv

```


```{r}
rownames(Ainv)  # Check the row names
colnames(Ainv)  # Should return NULL
colnames(Ainv) <- rownames(Ainv)
common_species <- intersect(sp_island_guild_size_filtered$species_name, rownames(Ainv))
Ainv_filtered <- Ainv[common_species, common_species]

dim(Ainv_filtered)  # Should match the number of species in common_species
length(common_species)  # Should be equal to the number of species in filtered Ainv

ginverse <- list(species_name = Ainv_filtered)

```


## CHECKS

```{r}
setdiff(sp_island_guild_size_filtered$species_name, common_species)

```
```{r}
all(sp_island_guild_size_filtered$species_name %in% rownames(Ainv_filtered))
```



```{r}
sp_island_guild_size_filtered$species_name <- factor(
    sp_island_guild_size_filtered$species_name,
    levels = rownames(ginverse$species_name)
)

setdiff(levels(sp_island_guild_size_filtered$species_name), rownames(ginverse$species_name))
setdiff(rownames(ginverse$species_name), levels(sp_island_guild_size_filtered$species_name))

anyDuplicated(levels(sp_island_guild_size_filtered$species_name))
anyDuplicated(rownames(ginverse$species_name))

all(sp_island_guild_size_filtered$species_name %in% rownames(ginverse$species_name))

required_columns <- c(
    "PCA_centrality", paste0("species_speciation_rate_", 1), 
    paste0("island_age_", 1), "guild_size", "island_area", "species_name"
)
missing_columns <- setdiff(required_columns, colnames(sp_island_guild_size_filtered))
missing_columns

```

```{r}
sp_island_guild_size_filtered <- as.data.frame(sp_island_guild_size_filtered)
```



Standarize variables

Z score speciation rates and centrality by island

```{r}
# Z-score PCA_centrality and speciation rates by island
sp_island_guild_size_filtered <- sp_island_guild_size_filtered %>%
  group_by(island) %>%
  mutate(
    # Z-score PCA_centrality by island
    z_PCA_centrality = scale(PCA_centrality) %>% as.numeric(),
    z_degree = scale(degree) %>% as.numeric(),
    
    # Identify the columns with speciation rates
    across(starts_with("species_speciation_rate_"), 
           ~ scale(.) %>% as.numeric(), 
           .names = "z_{.col}")
  ) %>%
  ungroup()

# Check the resulting dataset
head(sp_island_guild_size_filtered)
```


Z score guild_size and island_area

```{r}
sp_island_guild_size_filtered <- sp_island_guild_size_filtered %>%
    mutate(across(
        c(guild_size, 
          island_area), 
        ~ scale(.) %>% as.numeric()
    ))
```



```{r}
# sp_island_guild_size_filtered <- sp_island_guild_size_filtered %>%
#     mutate(across(
#         c(starts_with("species_speciation_rate_"), 
#           guild_size, 
#           island_area, 
#           starts_with("island_age"), 
#           PCA_centrality), 
#         ~ scale(.) %>% as.numeric()
#     ))

```




------------------------------------------------------------------------------


# Models

## Partial Model: Pca centrality

```{r}
my_priors <- list(R = list(V = 1/2, nu = 0.002),
                  G = list(G1 = list(V = 1/2, nu = 0.002)))


full_model.fit <- vector(mode = "list", length = 50)

for (i in 1:50) {
    full_model.fit[[i]] <- MCMCglmm(
        formula(paste0("z_PCA_centrality ~ z_species_speciation_rate_", i)),  # Fixed effects
        random = ~ species_name,  # Random effect for species_name
        family = "gaussian",  # Assuming a Gaussian response for PCA centrality
        ginverse = ginverse,  # Use species_name in ginverse
        data = as.data.frame(sp_island_guild_size_filtered),  # Dataframe
        prior = my_priors,  # Priors
        nitt = 1000000,  # Number of iterations
        burnin = 500000,  # Burn-in
        thin = 100  # Thinning
    )
}

saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/partial_model_pca_regional.rds"))
```



## Full model with PCA centrality


```{r}

my_priors <- list(R = list(V = 1/2, nu = 0.002),
                  G = list(G1 = list(V = 1/2, nu = 0.002)))


full_model.fit <- vector(mode = "list", length = 50)

for (i in 1:50) {
    full_model.fit[[i]] <- MCMCglmm(
        formula(paste0("z_PCA_centrality ~ z_species_speciation_rate_", i, 
                       " + guild_size + island_age_", i, 
                       " + island_area")),  # Fixed effects
        random = ~ species_name,  # Random effect for species_name
        family = "gaussian",  # Assuming a Gaussian response for PCA centrality
        ginverse = ginverse,  # Use species_name in ginverse
        data = as.data.frame(sp_island_guild_size_filtered),  # Dataframe
        prior = my_priors,  # Priors
        nitt = 1000000,  # Number of iterations
        burnin = 500000,  # Burn-in
        thin = 100  # Thinning
    )
}

saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/full_model.pca_local.rds"))

```


## Partial Model: degree

```{r}
my_priors <- list(R = list(V = 1/2, nu = 0.002),
                  G = list(G1 = list(V = 1/2, nu = 0.002)))


full_model.fit <- vector(mode = "list", length = 50)

for (i in 1:50) {
    full_model.fit[[i]] <- MCMCglmm(
        formula(paste0("z_degree ~ z_species_speciation_rate_", i)),  # Fixed effects
        random = ~ species_name,  # Random effect for species_name
        family = "gaussian",  # Assuming a Gaussian response for PCA centrality
        ginverse = ginverse,  # Use species_name in ginverse
        data = as.data.frame(sp_island_guild_size_filtered),  # Dataframe
        prior = my_priors,  # Priors
        nitt = 1000000,  # Number of iterations
        burnin = 500000,  # Burn-in
        thin = 100  # Thinning
    )
}

saveRDS(full_model.fit, here::here("../output/models/network_role_speciation/partial_model_degree_local.rds"))
```













-----------------------------------------------
## Model


Full model

```{r}

my_priors <- list(R = list(V = 1/2, nu = 0.002),
                  G = list(G1 = list(V = 1/2, nu = 0.002)))


full_model.fit <- vector(mode = "list", length = 50)

for (i in 1:50) {
    full_model.fit[[i]] <- MCMCglmm(
        formula(paste0("PCA_centrality ~ species_speciation_rate_", i, 
                       " + guild_size + island_age_", i, 
                       " + island_area")),  # Fixed effects
        random = ~ species_name,  # Random effect for species_name
        family = "gaussian",  # Assuming a Gaussian response for PCA centrality
        ginverse = ginverse,  # Use species_name in ginverse
        data = sp_island_guild_size_filtered,  # Dataframe
        prior = my_priors,  # Priors
        nitt = 1000000,  # Number of iterations
        burnin = 500000,  # Burn-in
        thin = 100  # Thinning
    )
}

saveRDS(full_model.fit, here::here("../output/models/network.role_speciation/full_model.fit.rds"))

```

Model with speciation rate only

```{r}
my_priors <- list(R = list(V = 1/2, nu = 0.002),
                  G = list(G1 = list(V = 1/2, nu = 0.002)))

model.fit_specrate <- vector(mode = "list", length = 50)

for (i in 1:50) {
    model.fit_specrate[[i]] <- MCMCglmm(
        formula(paste0("PCA_centrality ~ species_speciation_rate_", i)),  # Fixed effect: only species_speciation_rate
        random = ~ species_name,  # Random effect for species_name
        family = "gaussian",  # Assuming a Gaussian response for PCA centrality
        ginverse = ginverse,  # Use species_name in ginverse
        data = sp_island_guild_size_filtered,  # Dataframe
        prior = my_priors,  # Priors
        nitt = 1000000,  # Number of iterations
        burnin = 500000,  # Burn-in
        thin = 100  # Thinning
    )
}

saveRDS(model.fit_specrate, here::here("../output/models/network.role_speciation/model.fit_specrate.rds"))

```




```{r}
model.fit <- readRDS(here::here("../output/models/network.role_speciation/model.fit.rds"))
model.fit <- readRDS(here::here("../output/models/network.role_speciation/model.fit_specrate.rds"))

```



# Model results

```{r}
# Extract fixed effects (Sol) and variance components (VCV)
fixed_effects <- lapply(model.fit, function(model) as.mcmc(model$Sol))
variance_components <- lapply(model.fit, function(model) as.mcmc(model$VCV))
```



```{r}
# Rename the columns to standard names
fixed_effects_standardized <- lapply(seq_along(fixed_effects), function(i) {
  colnames(fixed_effects[[i]]) <- c("(Intercept)", "species_speciation_rate", 
                                    "guild_size", "island_age", "island_area")
  fixed_effects[[i]]
})

fixed_effects_standardized <- lapply(seq_along(fixed_effects), function(i) {
  colnames(fixed_effects[[i]]) <- c("(Intercept)", "species_speciation_rate")
  fixed_effects[[i]]
})


## combine chains

combined_fixed <- do.call(mcmc.list, fixed_effects_standardized)


```


## CHAINS CONVERGENCE

#### Traceplot

```{r}
traceplot(combined_fixed)
```

The traceplots show horizontal bands without trends or drifts, and they overlap and mix well, indicating that chains converged.

#### Autocorrelation plot

```{r}
autocorr.plot(combined_fixed)
```

There is a fast decay to zero in all cases, indicating that samples become effectively independent quickly, and there is no autocorrelation detected.

#### Gelman-Rubin diagnostic

```{r}
gelman_fixed <- gelman.diag(combined_fixed)
print(gelman_fixed)
```
The Gelman-Rubin diagnostic, also called the potential scale reduction factor (PSRF), is a method to assess the convergence of Markov Chain Monte Carlo (MCMC) simulations. It compares the variance within each chain to the variance between chains to determine whether the chains have converged to the same target distribution.

Interpretation:
Point Estimate: The closer this value is to 1, the better the convergence. Values above 1.1 suggest that the chains have not fully converged.
Upper C.I.: The upper bound of the confidence interval for the PSRF. This helps account for uncertainty in the diagnostic. Values significantly above 1.1 may indicate poor convergence.

Convergence is problematic for species_speciation_rate.


## Effective Sample Size

```{r}
effectiveSize(model.fit[[7]]$Sol[,2])
```


## Posterior distributions

```{r}
# Summarize fixed effects across all models
summary_fixed <- summary(combined_fixed)

print(summary_fixed)
```

```{r}
# Convert combined_fixed to a data frame
posterior_df <- as.data.frame(as.matrix(combined_fixed))

# Reshape the posterior data frame to long format
posterior_long <- posterior_df %>%
  pivot_longer(cols = everything(), names_to = "Predictor", values_to = "Value")

# Create the plot using stat_halfeye
ggplot(posterior_long, aes(x = Value, y = Predictor, fill = Predictor)) +
  stat_halfeye(
    aes(fill = after_stat(ifelse(x < 0, "negative", "positive"))),
    .width = c(0.5, 0.8, 0.95),
    alpha = 0.8
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + 
  scale_fill_manual(
    values = c("positive" = "skyblue", "negative" = "orange"),
    guide = "none"
  ) +
  labs(
    x = "Value",
    y = "Predictor"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",    
    panel.grid.major.y = element_blank()
  )

```


# Model with Z scores for speciation rates

```{r}
# Identify the columns with speciation rates
speciation_rate_columns <- grep("^species_speciation_rate_", names(sp_island_guild_size_filtered), value = TRUE)

# Z-score the speciation rates by island
sp_island_guild_size_filtered <- sp_island_guild_size_filtered %>%
  mutate(across(all_of(speciation_rate_columns), ~ scale(.) %>% as.numeric(), .names = "z_{.col}"))

# Check the resulting dataset
head(sp_island_guild_size_filtered)
```



```{r}

my_priors <- list(R = list(V = 1/2, nu = 0.002),
                  G = list(G1 = list(V = 1/2, nu = 0.002)))


model.fit_z <- vector(mode = "list", length = 50)

for (i in 1:50) {
    model.fit_z[[i]] <- MCMCglmm(
        formula(paste0("PCA_centrality ~ z_species_speciation_rate_", i, 
                       " + guild_size + island_age_", i, 
                       " + island_area")),  # Fixed effects
        random = ~ species_name,  # Random effect for species_name
        family = "gaussian",  # Assuming a Gaussian response for PCA centrality
        ginverse = ginverse,  # Use species_name in ginverse
        data = as.data.frame(sp_island_guild_size_filtered),  # Dataframe
        prior = my_priors,  # Priors
        nitt = 1000000,  # Number of iterations
        burnin = 500000,  # Burn-in
        thin = 100  # Thinning
    )
}

```



# Simplified Models - 1 variable at a time



```{r}

my_priors <- list(R = list(V = 1/2, nu = 0.002),
                  G = list(G1 = list(V = 1/2, nu = 0.002)))


model.fit_speciation <- vector(mode = "list", length = 50)

for (i in 1:50) {
    model.fit_speciation[[i]] <- MCMCglmm(
        formula(paste0("PCA_centrality ~ z_species_speciation_rate_", i)),  # Fixed effects
        random = ~ species_name,  # Random effect for species_name
        family = "gaussian",  # Assuming a Gaussian response for PCA centrality
        ginverse = ginverse,  # Use species_name in ginverse
        data = as.data.frame(sp_island_guild_size_filtered),  # Dataframe
        prior = my_priors,  # Priors
        nitt = 1000000,  # Number of iterations
        burnin = 500000,  # Burn-in
        thin = 100  # Thinning
    )
}

```

# Simplified Model - 2 variables at a time