---
title: "Load ~ brightness"
---

```{r}
library(dplyr)
library(ggplot2)# Load the ape package
library(ape)
library(readxl)
library(tidyr)
library(ggpubr)
library(ade4)
library(vegan)
library(stringr)
library(brms)
library(dplyr)
library(tidyverse)
library(cmdstanr)
library(bayesplot)
library(tidybayes)
library(posterior)

```

```{r}
my_theme<-theme(axis.text=element_text(size=12),
        axis.title = element_text(size = 14),
        legend.text=element_text(size=10),
        legend.title = element_text(size=12),
        plot.title = element_text(face="bold",size=14,margin=margin(0,0,20,0),hjust = 0.5),
        axis.title.y = element_text(hjust = 0.5),
        axis.title.x = element_text(hjust = 0.5))
```


# Load dataset


```{r}
df_load <- read_excel("../a_data/04april24_datalab.xlsx", sheet = 3)
df_load$nematode_count[which(df_load$nematode_count == ">100")] <- "100"
df_load$nematode_count <- as.numeric(df_load$nematode_count)

df_load <- df_load %>% 
  select(id, spp, nematode_count) %>% 
  rename(species = spp)

df_load
```




# brightness data

```{r}
df_kraemer <- read.csv("../a_data/kraemer2019_brightness_scores.csv")

df_kraemer <- df_kraemer %>% 
  select(location, spp, brightness) %>% 
  rename(id = location,
         species = spp)

df_kraemer

```


Names checks

```{r}
df_kraemer[which(df_kraemer$species == "albermarlensis"),"species"] <- "albemarlensis"
df_kraemer[which(df_kraemer$species == "cf. albemarlensis"),"species"] <- "albemarlensis"

df_kraemer[which(df_kraemer$species == "ustulatus pallescens"),"species"] <- "ustulatus"
df_kraemer[which(df_kraemer$species == "ustulatus phlegonis"),"species"] <- "ustulatus"
df_kraemer[which(df_kraemer$species == "ustulatus mahogany"),"species"] <- "ustulatus"

df_kraemer[which(df_kraemer$species == "invalidus 1"),"species"] <- "invalidus"
df_kraemer[which(df_kraemer$species == "invalidus 2"),"species"] <- "invalidus"

df_kraemer[which(df_kraemer$species == "sculpturatus 1"),"species"] <- "sculpturatus"
df_kraemer[which(df_kraemer$species == "sculpturatus 2"),"species"] <- "sculpturatus"
df_kraemer[which(df_kraemer$species == "sculpturatus 3"),"species"] <- "sculpturatus"

df_kraemer[which(df_kraemer$species == "wolfi 1"),"species"] <- "wolfi"
df_kraemer[which(df_kraemer$species == "wolfi 2"),"species"] <- "wolfi"
df_kraemer[which(df_kraemer$species == "wolfi 3"),"species"] <- "wolfi"

df_kraemer[which(df_kraemer$species == "cf. perspectivus"),"species"] <- "perspectivus"
df_kraemer[which(df_kraemer$species == "perspectivus 1"),"species"] <- "perspectivus"
df_kraemer[which(df_kraemer$species == "perspectivus 2"),"species"] <- "perspectivus"

df_kraemer[which(df_kraemer$species == "cf. amastroides"),"species"] <- "amastroides"

df_kraemer[which(df_kraemer$species == "canaliferus 1"),"species"] <- "canaliferus"
df_kraemer[which(df_kraemer$species == "canaliferus 2"),"species"] <- "canaliferus"

df_kraemer[which(df_kraemer$species == "simrothi 1"),"species"] <- "simrothi"
df_kraemer[which(df_kraemer$species == "simrothi 2"),"species"] <- "simrothi"

df_kraemer[which(df_kraemer$species == "cf. tortuganus"),"species"] <- "tortuganus"

df_kraemer[which(df_kraemer$species == "cf. nux"),"species"] <- "nux"
```


Only keep species with assessed load:

```{r}

# df_kraemer df_merged to only include species that appear in df_load
df_kraemer <- df_kraemer %>%
  semi_join(df_load, by = "species")

df_kraemer
```



### Create common species code to make sure species names coincide

```{r}

# Combine unique species from both dataframes
unique_species <- unique(c(unique(df_load$species), unique(df_kraemer$species)))

# Create a dictionary of species to letters
species_to_number <- setNames(1:length(unique_species), unique_species)

# Add the letter codes to df_load and df_habitat
df_load$species_code <- species_to_number[df_load$species]
df_kraemer$species_code <- species_to_number[df_kraemer$species]


df_load
df_kraemer

```




# MODEL STAN: load ~ brightness



```{r}

# Prepare the data list
data_list <- list(
  N_nem = nrow(df_load),
  N_bright = nrow(df_kraemer),
  N_sp_nem = length(unique(df_load$species)),
  N_sp_bright = length(unique(df_kraemer$species)),
  load = df_load$nematode_count,
  brightness = df_kraemer$brightness,
  species_i = df_load$species_code,
  species_k = df_kraemer$species_code
)


```





## model with log brightness and slope assumed positive

-> Truncation can't be done in poisson_log

```{r}

# Fit the model using cmdstanr
stan_load_brightness <- cmdstan_model(
  stan_file = "../c_codeStan/logbright_posslope.stan", 
  pedantic = TRUE
)

fit <- stan_load_brightness$sample(
  data = data_list,
  seed = 123,             
  chains = 4,               
  parallel_chains = 4,    
  iter_warmup = 1000,       
  iter_sampling = 2000   
)

```


## model with log brightness and slope assumed negative 



```{r}

# Fit the model using cmdstanr
stan_load_brightness <- cmdstan_model(
  stan_file = "../c_codeStan/logbright_negslope.stan", 
  pedantic = TRUE
)

fit <- stan_load_brightness$sample(
  data = data_list,
  seed = 123,             
  chains = 4,               
  parallel_chains = 4,    
  iter_warmup = 1000,       
  iter_sampling = 2000   
)

```





# Plotting results



## Posterior vs predicted

```{r}
stan_load_brightness |> tidybayes::gather_rvars(predicted_load[i])
```


## All posterior pars

```{r}
# Extract the posterior samples
posterior_samples <- fit$draws()

# Convert to a data frame for easier handling
posterior_df <- as_draws_df(posterior_samples)

# Rename mu_bright parameters for easier handling
posterior_df <- posterior_df %>%
  rename_with(~ gsub("\\[|\\]", "", .), starts_with("mu_bright")) %>%
  rename_with(~ gsub("\\.", "_", .), starts_with("mu_bright"))

# Extract the names of the parameters of interest
mu_bright_names <- grep("^mu_bright", names(posterior_df), value = TRUE)
sigma_bright_names <- grep("^sigma_bright", names(posterior_df), value = TRUE)

parameter_names <- c("intercept", "slope_bright", mu_bright_names, sigma_bright_names)
```


```{r}
# Plot posterior distributions of the parameters
mcmc_areas(
  posterior_df,
  pars = parameter_names,
  prob = 0.8  # 80% intervals
)+
  coord_cartesian(xlim=c(-2,11))


```




### Posterior brightness slope

```{r}
# Extract the posterior samples of the slope of brightness
slope_samples <- fit$draws("slope_bright")

# Plot the posterior distribution of the slope of brightness
mcmc_hist(slope_samples, prob = 0.8)

```



```{r}
draws <- fit$draws(format = "df")

library(ggdist)

draws |>
  ggplot(aes(x=slope_bright)) + 
  stat_dotsinterval() + 
  xlim(c(-3,2.5))
```




```{r}
fit$draws(variables = c("intercept", "slope_bright")) |>
  posterior::summarise_draws("mean", "sd", ~quantile(.x, probs = c(0.05, 0.95)), "rhat") #|>
  #flextable::flextable()
```


### Plot for predicted values


```{r}

# Extract the predicted values from the Stan fit object
predicted_brightness <- fit$draws("predicted_brightness")
predicted_load <- fit$draws("predicted_load")

# Convert the extracted predictions to a format suitable for bayesplot
predicted_load <- as_draws_matrix(predicted_load)
predicted_brightness <- as_draws_matrix(predicted_brightness)

# Posterior predictive check for nematode load
ppc_load <- ppc_load <- ppc_dens_overlay(y = df_load$nematode_count, yrep = predicted_load) +
  scale_color_manual(name = "Legend",
                     values = c("y" = "black", "yrep" = "blue"),
                     labels = c("Observed load", "Predicted load")) +
  scale_fill_manual(name = "Legend",
                    values = c("y" = "black", "yrep" = "blue"),
                    labels = c("Observed load", "Predicted load"))+
  my_theme

# Posterior predictive check for brightness
observed_brightness <- df_kraemer$brightness
ppc_brightness <- ppc_dens_overlay(y = observed_brightness, yrep = predicted_brightness) +
  scale_color_manual(name = "Legend",
                     values = c("y" = "black", "yrep" = "coral"),
                     labels = c("Observed brightness", "Predicted brightness")) +
  scale_fill_manual(name = "Legend",
                    values = c("y" = "black", "yrep" = "coral"),
                    labels = c("Observed brightness", "Predicted brightness"))+
  xlim(0,60000)
  my_theme


ppc_plots_load_bright <- ggarrange(
  
  ppc_load,
  ppc_brightness,
  labels = LETTERS[1:2],
  ncol = 1,
  nrow = 2
)

ppc_plots_load_bright

ggsave("../d_output/figures/predictions/ppc_load_bright_neg.png", height = 10, width = 15)

```


