---
title: "Load ~ ecological factors"
output: html_notebook
---

```{r}
library(dplyr)
library(ggplot2)# Load the ape package
library(ape)
library(readxl)
library(tidyr)
library(ggpubr)
library(ade4)
library(vegan)
library(stringr)
library(brms)
library(dplyr)
library(tidyverse)
library(cmdstanr)
library(bayesplot)
library(tidybayes)
library(posterior)

```

```{r}
my_theme<-theme(axis.text=element_text(size=12),
        axis.title = element_text(size = 14),
        legend.text=element_text(size=10),
        legend.title = element_text(size=12),
        plot.title = element_text(face="bold",size=14,margin=margin(0,0,20,0),hjust = 0.5),
        axis.title.y = element_text(hjust = 0.5),
        axis.title.x = element_text(hjust = 0.5))
```


# Load dataset

```{r}
df_island <- read_excel("../a_data04april24_datalab.xlsx", sheet = "island_characteristics")
df_load <- read_excel("../a_data04april24_datalab.xlsx", sheet = 3)
df_load$nematode_count[which(df_load$nematode_count == ">100")] <- "100"
df_load$nematode_count <- as.numeric(df_load$nematode_count)

df_load <- df_load %>% 
  select(id, spp, nematode_count, island) %>% 
  rename(species = spp)

df_load
```

Adding island names to df_load

```{r}
# Rename the column "mean_emergence" to "island_age" in df_island
df_island <- df_island %>% rename(island_age = mean_emergence,
                                  island_area = area)

# Merge the dataframes based on the "island" column
df_load <- df_load %>%
  left_join(df_island %>% select(island, island_age, island_area), by = "island")

df_load
```


# brightness data

```{r}
df_kraemer <- read.csv("../a_data/kraemer2019_brightness_scores.csv")

df_kraemer <- df_kraemer %>% 
  select(location, spp, brightness) %>% 
  rename(id = location,
         species = spp)

df_kraemer

```


Names checks

```{r}
df_kraemer[which(df_kraemer$species == "albermarlensis"),"species"] <- "albemarlensis"
df_kraemer[which(df_kraemer$species == "cf. albemarlensis"),"species"] <- "albemarlensis"

df_kraemer[which(df_kraemer$species == "ustulatus pallescens"),"species"] <- "ustulatus"
df_kraemer[which(df_kraemer$species == "ustulatus phlegonis"),"species"] <- "ustulatus"
df_kraemer[which(df_kraemer$species == "ustulatus mahogany"),"species"] <- "ustulatus"

df_kraemer[which(df_kraemer$species == "invalidus 1"),"species"] <- "invalidus"
df_kraemer[which(df_kraemer$species == "invalidus 2"),"species"] <- "invalidus"

df_kraemer[which(df_kraemer$species == "sculpturatus 1"),"species"] <- "sculpturatus"
df_kraemer[which(df_kraemer$species == "sculpturatus 2"),"species"] <- "sculpturatus"
df_kraemer[which(df_kraemer$species == "sculpturatus 3"),"species"] <- "sculpturatus"

df_kraemer[which(df_kraemer$species == "wolfi 1"),"species"] <- "wolfi"
df_kraemer[which(df_kraemer$species == "wolfi 2"),"species"] <- "wolfi"
df_kraemer[which(df_kraemer$species == "wolfi 3"),"species"] <- "wolfi"

df_kraemer[which(df_kraemer$species == "cf. perspectivus"),"species"] <- "perspectivus"
df_kraemer[which(df_kraemer$species == "perspectivus 1"),"species"] <- "perspectivus"
df_kraemer[which(df_kraemer$species == "perspectivus 2"),"species"] <- "perspectivus"

df_kraemer[which(df_kraemer$species == "cf. amastroides"),"species"] <- "amastroides"

df_kraemer[which(df_kraemer$species == "canaliferus 1"),"species"] <- "canaliferus"
df_kraemer[which(df_kraemer$species == "canaliferus 2"),"species"] <- "canaliferus"

df_kraemer[which(df_kraemer$species == "simrothi 1"),"species"] <- "simrothi"
df_kraemer[which(df_kraemer$species == "simrothi 2"),"species"] <- "simrothi"

df_kraemer[which(df_kraemer$species == "cf. tortuganus"),"species"] <- "tortuganus"

df_kraemer[which(df_kraemer$species == "cf. nux"),"species"] <- "nux"
```


Only keep species with assessed load:

```{r}

# df_kraemer df_merged to only include species that appear in df_load
df_kraemer <- df_kraemer %>%
  semi_join(df_load, by = "species")

df_kraemer
```



### Create common species code to make sure species names coincide

```{r}

# Combine unique species from both dataframes
unique_species <- unique(c(unique(df_load$species), unique(df_kraemer$species)))

# Create a dictionary of species to letters
species_to_number <- setNames(1:length(unique_species), unique_species)

# Add the letter codes to df_load and df_habitat
df_load$species_code <- species_to_number[df_load$species]
df_kraemer$species_code <- species_to_number[df_kraemer$species]

df_kraemer

```





# habitat data


### Habitat data from Kraemer 2019 database

```{r}
df_kraemer <- read.csv("../a_data/kraemer2019_brightness_scores.csv")

df_kraemer <- df_kraemer %>% 
  select(location, spp, microhab) %>% 
  rename(id = location,
         species = spp,
         habitat = microhab)

df_kraemer

```


### habitat data from RAD sequences database

```{r}
df_rad <- read_excel("../a_data/GPS RAD snails.xlsx")
df_rad <- df_rad %>% 
  select(Sample, Taxon, Microhabitat) %>% 
  rename(id = Sample,
         species = Taxon,
         habitat = Microhabitat)

# Semi-Arboreal is considered arboreal
df_rad <- df_rad %>%
  mutate(habitat = str_replace(habitat, "Semi-Arboreal", "arboreal"))

df_rad <- df_rad %>%
  mutate(habitat = str_to_lower(habitat)) %>% 
  filter(habitat %in% c("arboreal", "terrestrial"))

df_rad
  
```


### Merge data to obtain single habitat database

```{r}
# Remove rows from df_rad that have matching ids in df_kraemer
df_rad_filtered <- anti_join(df_rad, df_kraemer, by = "id")

# Combine df_kraemer with the filtered df_rad
df_habitat <- bind_rows(df_kraemer, df_rad_filtered)

df_habitat
```


Names checks

```{r}
df_habitat[which(df_habitat$species == "albermarlensis"),"species"] <- "albemarlensis"
df_habitat[which(df_habitat$species == "cf. albemarlensis"),"species"] <- "albemarlensis"

df_habitat[which(df_habitat$species == "ustulatus pallescens"),"species"] <- "ustulatus"
df_habitat[which(df_habitat$species == "ustulatus phlegonis"),"species"] <- "ustulatus"
df_habitat[which(df_habitat$species == "ustulatus mahogany"),"species"] <- "ustulatus"

df_habitat[which(df_habitat$species == "invalidus 1"),"species"] <- "invalidus"
df_habitat[which(df_habitat$species == "invalidus 2"),"species"] <- "invalidus"

df_habitat[which(df_habitat$species == "sculpturatus 1"),"species"] <- "sculpturatus"
df_habitat[which(df_habitat$species == "sculpturatus 2"),"species"] <- "sculpturatus"
df_habitat[which(df_habitat$species == "sculpturatus 3"),"species"] <- "sculpturatus"

df_habitat[which(df_habitat$species == "wolfi 1"),"species"] <- "wolfi"
df_habitat[which(df_habitat$species == "wolfi 2"),"species"] <- "wolfi"
df_habitat[which(df_habitat$species == "wolfi 3"),"species"] <- "wolfi"

df_habitat[which(df_habitat$species == "cf. perspectivus"),"species"] <- "perspectivus"
df_habitat[which(df_habitat$species == "perspectivus 1"),"species"] <- "perspectivus"
df_habitat[which(df_habitat$species == "perspectivus 2"),"species"] <- "perspectivus"

df_habitat[which(df_habitat$species == "cf. amastroides"),"species"] <- "amastroides"

df_habitat[which(df_habitat$species == "canaliferus 1"),"species"] <- "canaliferus"
df_habitat[which(df_habitat$species == "canaliferus 2"),"species"] <- "canaliferus"

df_habitat[which(df_habitat$species == "simrothi 1"),"species"] <- "simrothi"
df_habitat[which(df_habitat$species == "simrothi 2"),"species"] <- "simrothi"

df_habitat[which(df_habitat$species == "cf. tortuganus"),"species"] <- "tortuganus"

df_habitat[which(df_habitat$species == "cf. nux"),"species"] <- "nux"

```


Only keep species with assessed load:

```{r}

df_habitat1 <- df_habitat

# Filter df_merged to only include species that appear in df_load
df_habitat <- df_habitat %>%
  semi_join(df_load, by = "species")

df_habitat
```


which species were discarded?

```{r}
discarded_species <- setdiff(unique(df_habitat1$species), unique(df_habitat$species))
discarded_species
```


### Create common species code to make sure species names coincide

```{r}

# Combine unique species from both dataframes
unique_species <- unique(c(unique(df_load$species), unique(df_habitat$species)))

# Create a dictionary of species to letters
species_to_number <- setNames(1:length(unique_species), unique_species)

# Add the letter codes to df_load and df_habitat
df_load$species_code <- species_to_number[df_load$species]
df_habitat$species_code <- species_to_number[df_habitat$species]


df_habitat

```
# vegetation data

### vegetation zone data from RAD sequences database

```{r}
df_rad <- read_excel("../a_data/GPS RAD snails.xlsx")

# Standardize the names to lowercase
df_rad <- df_rad %>%
  mutate(`Vegetation Zone` = tolower(`Vegetation Zone`))

# Filter out rows with unwanted values in the Vegetation Zone column
df_rad <- df_rad %>%
  filter(`Vegetation Zone` %in% c("humid", "arid"))

# Rename the column
df_veg <- df_rad %>%
  rename(vegetation_zone = `Vegetation Zone`)
  
```

Names checks

```{r}
#df_veg[which(df_veg$Taxon == "albermarlensis"),] <- "albemarlensis"
df_veg[which(df_veg$Taxon == "cf. albemarlensis"),"Taxon"] <- "albemarlensis"

#df_veg[which(df_veg$Taxon == "ustulatus pallescens"),] <- "ustulatus"
#df_veg[which(df_veg$Taxon == "ustulatus phlegonis"),] <- "ustulatus"
#df_veg[which(df_veg$Taxon == "ustulatus mahogany"),] <- "ustulatus"

df_veg[which(df_veg$Taxon == "invalidus 1"),"Taxon"] <- "invalidus"
df_veg[which(df_veg$Taxon == "invalidus 2"),"Taxon"] <- "invalidus"

df_veg[which(df_veg$Taxon == "sculpturatus 1"),"Taxon"] <- "sculpturatus"
df_veg[which(df_veg$Taxon == "sculpturatus 2"),"Taxon"] <- "sculpturatus"
df_veg[which(df_veg$Taxon == "sculpturatus 3"),"Taxon"] <- "sculpturatus"

df_veg[which(df_veg$Taxon == "wolfi 1"),"Taxon"] <- "wolfi"
df_veg[which(df_veg$Taxon == "wolfi 2"),"Taxon"] <- "wolfi"
df_veg[which(df_veg$Taxon == "wolfi 3"),"Taxon"] <- "wolfi"

df_veg[which(df_veg$Taxon == "cf. perspectivus"),"Taxon"] <- "perspectivus"
df_veg[which(df_veg$Taxon == "perspectivus 1"),"Taxon"] <- "perspectivus"
df_veg[which(df_veg$Taxon == "perspectivus 2"),"Taxon"] <- "perspectivus"

df_veg[which(df_veg$Taxon == "cf. amastroides"),"Taxon"] <- "amastroides"

df_veg[which(df_veg$Taxon == "canaliferus 1"),"Taxon"] <- "canaliferus"
df_veg[which(df_veg$Taxon == "canaliferus 2"),"Taxon"] <- "canaliferus"

df_veg[which(df_veg$Taxon == "simrothi 1"),"Taxon"] <- "simrothi"
df_veg[which(df_veg$Taxon == "simrothi 2"),"Taxon"] <- "simrothi"

df_veg[which(df_veg$Taxon == "cf. tortuganus"),"Taxon"] <- "tortuganus"

df_veg[which(df_veg$Taxon == "cf. nux"),"Taxon"] <- "nux"


```

Only keep species with assessed load:

```{r}

# Filter df_merged to only include species that appear in df_load
df_veg <- df_veg %>%
  rename(species = Taxon) %>% 
  semi_join(df_load, by = "species")

df_veg
```


### Create common species code to make sure species names coincide

```{r}

# Combine unique species from both dataframes
unique_species <- unique(c(unique(df_load$species), unique(df_veg$species)))

# Create a dictionary of species to letters
species_to_number <- setNames(1:length(unique_species), unique_species)

# Add the letter codes to df_load and df_habitat
df_load$species_code <- species_to_number[df_load$species]
df_veg$species_code <- species_to_number[df_veg$species]

df_veg

```