---
title: "Load ~ ecological factors"
---

```{r}
library(dplyr)
library(ggplot2)# Load the ape package
library(ape)
library(readxl)
library(tidyr)
library(ggpubr)
library(ade4)
library(vegan)
library(stringr)
library(brms)
library(dplyr)
library(tidyverse)
library(cmdstanr)
library(bayesplot)
library(tidybayes)
library(posterior)

```

```{r}
my_theme<-theme(axis.text=element_text(size=12),
        axis.title = element_text(size = 14),
        legend.text=element_text(size=10),
        legend.title = element_text(size=12),
        plot.title = element_text(face="bold",size=14,margin=margin(0,0,20,0),hjust = 0.5),
        axis.title.y = element_text(hjust = 0.5),
        axis.title.x = element_text(hjust = 0.5))
```


# Load dataset

```{r}
df_island <- read_excel("../a_data/04april24_datalab.xlsx", sheet = "island_characteristics")
df_load <- read_excel("../a_data/04april24_datalab.xlsx", sheet = 3)
df_load$nematode_count[which(df_load$nematode_count == ">100")] <- "100"
df_load$nematode_count <- as.numeric(df_load$nematode_count)

df_load <- df_load %>% 
  select(id, spp, nematode_count, island) %>% 
  rename(species = spp)
```

Adding island names to df_load

```{r}
# Rename the column "mean_emergence" to "island_age" in df_island
df_island <- df_island %>% rename(island_age = mean_emergence,
                                  island_area = area)

# Merge the dataframes based on the "island" column
df_load <- df_load %>%
  left_join(df_island %>% select(island, island_age, island_area), by = "island")

df_load
```


# brightness data

```{r}
df_kraemer <- read.csv("../a_data/kraemer2019_brightness_scores.csv")

df_kraemer <- df_kraemer %>% 
  select(location, spp, brightness) %>% 
  rename(id = location,
         species = spp)

df_kraemer

```


Names checks

```{r}
df_kraemer[which(df_kraemer$species == "albermarlensis"),"species"] <- "albemarlensis"
df_kraemer[which(df_kraemer$species == "cf. albemarlensis"),"species"] <- "albemarlensis"

df_kraemer[which(df_kraemer$species == "ustulatus pallescens"),"species"] <- "ustulatus"
df_kraemer[which(df_kraemer$species == "ustulatus phlegonis"),"species"] <- "ustulatus"
df_kraemer[which(df_kraemer$species == "ustulatus mahogany"),"species"] <- "ustulatus"

df_kraemer[which(df_kraemer$species == "invalidus 1"),"species"] <- "invalidus"
df_kraemer[which(df_kraemer$species == "invalidus 2"),"species"] <- "invalidus"

df_kraemer[which(df_kraemer$species == "sculpturatus 1"),"species"] <- "sculpturatus"
df_kraemer[which(df_kraemer$species == "sculpturatus 2"),"species"] <- "sculpturatus"
df_kraemer[which(df_kraemer$species == "sculpturatus 3"),"species"] <- "sculpturatus"

df_kraemer[which(df_kraemer$species == "wolfi 1"),"species"] <- "wolfi"
df_kraemer[which(df_kraemer$species == "wolfi 2"),"species"] <- "wolfi"
df_kraemer[which(df_kraemer$species == "wolfi 3"),"species"] <- "wolfi"

df_kraemer[which(df_kraemer$species == "cf. perspectivus"),"species"] <- "perspectivus"
df_kraemer[which(df_kraemer$species == "perspectivus 1"),"species"] <- "perspectivus"
df_kraemer[which(df_kraemer$species == "perspectivus 2"),"species"] <- "perspectivus"

df_kraemer[which(df_kraemer$species == "cf. amastroides"),"species"] <- "amastroides"

df_kraemer[which(df_kraemer$species == "canaliferus 1"),"species"] <- "canaliferus"
df_kraemer[which(df_kraemer$species == "canaliferus 2"),"species"] <- "canaliferus"

df_kraemer[which(df_kraemer$species == "simrothi 1"),"species"] <- "simrothi"
df_kraemer[which(df_kraemer$species == "simrothi 2"),"species"] <- "simrothi"

df_kraemer[which(df_kraemer$species == "cf. tortuganus"),"species"] <- "tortuganus"

df_kraemer[which(df_kraemer$species == "cf. nux"),"species"] <- "nux"
```


Only keep species with assessed load:

```{r}

# df_kraemer df_merged to only include species that appear in df_load
df_bright <- df_kraemer %>%
  semi_join(df_load, by = "species")
```



### Create common species code to make sure species names coincide

```{r}

# Combine unique species from both dataframes
unique_species <- unique(c(unique(df_load$species), unique(df_bright$species)))

# Create a dictionary of species to letters
species_to_number <- setNames(1:length(unique_species), unique_species)

# Add the letter codes to df_load and df_habitat
df_load$species_code <- species_to_number[df_load$species]
df_bright$species_code <- species_to_number[df_bright$species]

df_bright

```





# habitat data


### Habitat data from Kraemer 2019 database

```{r}
df_kraemer <- read.csv("../a_data/kraemer2019_brightness_scores.csv")

df_kraemer <- df_kraemer %>% 
  select(location, spp, microhab) %>% 
  rename(id = location,
         species = spp,
         habitat = microhab)

```


### habitat data from RAD sequences database

```{r}
df_rad <- read_excel("../a_data/GPS RAD snails.xlsx")
df_rad <- df_rad %>% 
  select(Sample, Taxon, Microhabitat) %>% 
  rename(id = Sample,
         species = Taxon,
         habitat = Microhabitat)

# Semi-Arboreal is considered arboreal
df_rad <- df_rad %>%
  mutate(habitat = str_replace(habitat, "Semi-Arboreal", "arboreal"))

df_rad <- df_rad %>%
  mutate(habitat = str_to_lower(habitat)) %>% 
  filter(habitat %in% c("arboreal", "terrestrial"))
  
```


### Merge data to obtain single habitat database

```{r}
# Remove rows from df_rad that have matching ids in df_kraemer
df_rad_filtered <- anti_join(df_rad, df_kraemer, by = "id")

# Combine df_kraemer with the filtered df_rad
df_habitat <- bind_rows(df_kraemer, df_rad_filtered)
```


Names checks

```{r}
df_habitat[which(df_habitat$species == "albermarlensis"),"species"] <- "albemarlensis"
df_habitat[which(df_habitat$species == "cf. albemarlensis"),"species"] <- "albemarlensis"

df_habitat[which(df_habitat$species == "ustulatus pallescens"),"species"] <- "ustulatus"
df_habitat[which(df_habitat$species == "ustulatus phlegonis"),"species"] <- "ustulatus"
df_habitat[which(df_habitat$species == "ustulatus mahogany"),"species"] <- "ustulatus"

df_habitat[which(df_habitat$species == "invalidus 1"),"species"] <- "invalidus"
df_habitat[which(df_habitat$species == "invalidus 2"),"species"] <- "invalidus"

df_habitat[which(df_habitat$species == "sculpturatus 1"),"species"] <- "sculpturatus"
df_habitat[which(df_habitat$species == "sculpturatus 2"),"species"] <- "sculpturatus"
df_habitat[which(df_habitat$species == "sculpturatus 3"),"species"] <- "sculpturatus"

df_habitat[which(df_habitat$species == "wolfi 1"),"species"] <- "wolfi"
df_habitat[which(df_habitat$species == "wolfi 2"),"species"] <- "wolfi"
df_habitat[which(df_habitat$species == "wolfi 3"),"species"] <- "wolfi"

df_habitat[which(df_habitat$species == "cf. perspectivus"),"species"] <- "perspectivus"
df_habitat[which(df_habitat$species == "perspectivus 1"),"species"] <- "perspectivus"
df_habitat[which(df_habitat$species == "perspectivus 2"),"species"] <- "perspectivus"

df_habitat[which(df_habitat$species == "cf. amastroides"),"species"] <- "amastroides"

df_habitat[which(df_habitat$species == "canaliferus 1"),"species"] <- "canaliferus"
df_habitat[which(df_habitat$species == "canaliferus 2"),"species"] <- "canaliferus"

df_habitat[which(df_habitat$species == "simrothi 1"),"species"] <- "simrothi"
df_habitat[which(df_habitat$species == "simrothi 2"),"species"] <- "simrothi"

df_habitat[which(df_habitat$species == "cf. tortuganus"),"species"] <- "tortuganus"

df_habitat[which(df_habitat$species == "cf. nux"),"species"] <- "nux"

```


Only keep species with assessed load:

```{r}

# Filter df_merged to only include species that appear in df_load
df_habitat <- df_habitat %>%
  semi_join(df_load, by = "species")
```


### Create common species code to make sure species names coincide

```{r}

# Combine unique species from both dataframes
unique_species <- unique(c(unique(df_load$species), unique(df_habitat$species)))

# Create a dictionary of species to letters
species_to_number <- setNames(1:length(unique_species), unique_species)

# Add the letter codes to df_load and df_habitat
df_load$species_code <- species_to_number[df_load$species]
df_habitat$species_code <- species_to_number[df_habitat$species]


df_habitat

```
# vegetation data

### vegetation zone data from RAD sequences database

```{r}
df_rad <- read_excel("../a_data/GPS RAD snails.xlsx")

# Standardize the names to lowercase
df_rad <- df_rad %>%
  mutate(`Vegetation Zone` = tolower(`Vegetation Zone`))

# Filter out rows with unwanted values in the Vegetation Zone column
df_rad <- df_rad %>%
  filter(`Vegetation Zone` %in% c("humid", "arid"))

# Rename the column
df_veg <- df_rad %>%
  rename(vegetation_zone = `Vegetation Zone`)
  
```

Names checks

```{r}
#df_veg[which(df_veg$Taxon == "albermarlensis"),] <- "albemarlensis"
df_veg[which(df_veg$Taxon == "cf. albemarlensis"),"Taxon"] <- "albemarlensis"

#df_veg[which(df_veg$Taxon == "ustulatus pallescens"),] <- "ustulatus"
#df_veg[which(df_veg$Taxon == "ustulatus phlegonis"),] <- "ustulatus"
#df_veg[which(df_veg$Taxon == "ustulatus mahogany"),] <- "ustulatus"

df_veg[which(df_veg$Taxon == "invalidus 1"),"Taxon"] <- "invalidus"
df_veg[which(df_veg$Taxon == "invalidus 2"),"Taxon"] <- "invalidus"

df_veg[which(df_veg$Taxon == "sculpturatus 1"),"Taxon"] <- "sculpturatus"
df_veg[which(df_veg$Taxon == "sculpturatus 2"),"Taxon"] <- "sculpturatus"
df_veg[which(df_veg$Taxon == "sculpturatus 3"),"Taxon"] <- "sculpturatus"

df_veg[which(df_veg$Taxon == "wolfi 1"),"Taxon"] <- "wolfi"
df_veg[which(df_veg$Taxon == "wolfi 2"),"Taxon"] <- "wolfi"
df_veg[which(df_veg$Taxon == "wolfi 3"),"Taxon"] <- "wolfi"

df_veg[which(df_veg$Taxon == "cf. perspectivus"),"Taxon"] <- "perspectivus"
df_veg[which(df_veg$Taxon == "perspectivus 1"),"Taxon"] <- "perspectivus"
df_veg[which(df_veg$Taxon == "perspectivus 2"),"Taxon"] <- "perspectivus"

df_veg[which(df_veg$Taxon == "cf. amastroides"),"Taxon"] <- "amastroides"

df_veg[which(df_veg$Taxon == "canaliferus 1"),"Taxon"] <- "canaliferus"
df_veg[which(df_veg$Taxon == "canaliferus 2"),"Taxon"] <- "canaliferus"

df_veg[which(df_veg$Taxon == "simrothi 1"),"Taxon"] <- "simrothi"
df_veg[which(df_veg$Taxon == "simrothi 2"),"Taxon"] <- "simrothi"

df_veg[which(df_veg$Taxon == "cf. tortuganus"),"Taxon"] <- "tortuganus"

df_veg[which(df_veg$Taxon == "cf. nux"),"Taxon"] <- "nux"


```

Only keep species with assessed load:

```{r}

# Filter df_merged to only include species that appear in df_load
df_veg <- df_veg %>%
  rename(species = Taxon) %>% 
  semi_join(df_load, by = "species")

```


### Create common species code to make sure species names coincide

```{r}

# Combine unique species from both dataframes
unique_species <- unique(c(unique(df_load$species), unique(df_veg$species)))

# Create a dictionary of species to letters
species_to_number <- setNames(1:length(unique_species), unique_species)

# Add the letter codes to df_load and df_habitat
df_load$species_code <- species_to_number[df_load$species]
df_veg$species_code <- species_to_number[df_veg$species]

df_veg

```




# Model



## Data


```{r}
df_habitat$habitat_code <- ifelse(df_habitat$habitat == "terrestrial", 0, 1)
df_veg$veg_code <- ifelse(df_veg$vegetation_zone == "humid", 0, 1)
```



```{r}
# Calculate number of observations and species
N_nem <- nrow(df_load)             # Number of observations for nematode load
N_bright <- nrow(df_bright)       # Number of observations for brightness
N_habitat <- nrow(df_habitat)      # Number of observations for habitat
N_veg <- nrow(df_veg)              # Number of observations for vegetation zone

N_sp_nem <- length(unique(df_load$species))  # Number of species in the nematode data


# Data list for Stan model
data_list <- list(
  N_nem = N_nem,                                     # Number of observations for nematode load
  N_bright = N_bright,                               # Number of observations for brightness
  N_habitat = N_habitat,                             # Number of observations for habitat
  N_veg = N_veg,                                     # Number of observations for vegetation zone
  N_sp_nem = N_sp_nem,                               # Number of species in the nematode data
  load = df_load$nematode_count,                      # Nematode load counts
  brightness = df_bright$brightness,                 # Brightness scores
  habitat = df_habitat$habitat_code,  
  vegetation = df_veg$veg_code,
  island_age = df_load$island_age,                  # Island age for each observation
  island_area = df_load$island_area,                # Island area for each observation
  species_i = df_load$species_code,                   # Species id for load dataset
  species_k = df_bright$species_code,                       # Species id for brightness dataset
  species_h = df_habitat$species_code,                      # Species id for habitat dataset
  species_v = df_veg$species_code                           # Species id for vegetation dataset
)

```



## Run model

```{r}
# Fit the model using cmdstanr
stan_model <- cmdstan_model(
  stan_file = "../c_codeStan/full_model.stan", 
  pedantic = TRUE
)

fit <- stan_model$sample(
  data = data_list,
  seed = 123,             
  chains = 4,               
  parallel_chains = 4,    
  iter_warmup = 1000,       
  iter_sampling = 2000   
)

```

