---
title: "Data exploration"
---


DOUBT NAMES JOHN

Jacobinus (in my load dataset)
Jacobi (in other datasets)
Olla Jacobinus (?)

```{r}
library(dplyr)
library(ggplot2)# Load the ape package
library(ape)
library(readxl)
library(tidyr)
library(ggpubr)
library(ade4)
library(vegan)
library(stringr)
library(brms)
library(dplyr)
library(tidyverse)
library(cmdstanr)
library(viridis)
```


parameters for plotting

```{r}
my_theme<-theme(axis.text=element_text(size=12),
        axis.title = element_text(size = 14),
        legend.text=element_text(size=10),
        legend.title = element_text(size=12),
        plot.title = element_text(face="bold",size=14,margin=margin(0,0,20,0),hjust = 0.5),
        axis.title.y = element_text(hjust = 0.5),
        axis.title.x = element_text(hjust = 0.5))
```



# Load data

```{r}

#load
df_load <- read_excel("a_data/04april24_datalab.xlsx", sheet = 3)

df_load$nematode_count[which(df_load$nematode_count == ">100")] <- "100"
df_load$nematode_count <- as.numeric(df_load$nematode_count)

df_load <- df_load %>% 
  select(id, spp, nematode_count, island) %>% 
  rename(species = spp)

df_load

```


# island data

Add island data to df_load

```{r}
df_island <- read_excel("a_data/04april24_datalab.xlsx", sheet = "island_characteristics")
```

Adding island names to df_load

```{r}
# Rename the column "mean_emergence" to "island_age" in df_island
df_island <- df_island %>% rename(island_age = mean_emergence,
                                  island_area = area)

# Merge the dataframes based on the "island" column
df_load <- df_load %>%
  left_join(df_island %>% select(island, island_age, island_area), by = "island")

df_load
```





# brightness data

```{r}
df_kraemer <- read.csv("a_data/kraemer2019_brightness_scores.csv")

df_kraemer <- df_kraemer %>% 
  select(location, spp, brightness) %>% 
  rename(id = location,
         species = spp)

df_kraemer
```




Names checks

```{r}
df_kraemer[which(df_kraemer$species == "albermarlensis"),"species"] <- "albemarlensis"
df_kraemer[which(df_kraemer$species == "cf. albemarlensis"),"species"] <- "albemarlensis"

df_kraemer[which(df_kraemer$species == "ustulatus pallescens"),"species"] <- "ustulatus"
df_kraemer[which(df_kraemer$species == "ustulatus phlegonis"),"species"] <- "ustulatus"
df_kraemer[which(df_kraemer$species == "ustulatus mahogany"),"species"] <- "ustulatus"

df_kraemer[which(df_kraemer$species == "invalidus 1"),"species"] <- "invalidus"
df_kraemer[which(df_kraemer$species == "invalidus 2"),"species"] <- "invalidus"

df_kraemer[which(df_kraemer$species == "sculpturatus 1"),"species"] <- "sculpturatus"
df_kraemer[which(df_kraemer$species == "sculpturatus 2"),"species"] <- "sculpturatus"
df_kraemer[which(df_kraemer$species == "sculpturatus 3"),"species"] <- "sculpturatus"

df_kraemer[which(df_kraemer$species == "wolfi 1"),"species"] <- "wolfi"
df_kraemer[which(df_kraemer$species == "wolfi 2"),"species"] <- "wolfi"
df_kraemer[which(df_kraemer$species == "wolfi 3"),"species"] <- "wolfi"

df_kraemer[which(df_kraemer$species == "cf. perspectivus"),"species"] <- "perspectivus"
df_kraemer[which(df_kraemer$species == "perspectivus 1"),"species"] <- "perspectivus"
df_kraemer[which(df_kraemer$species == "perspectivus 2"),"species"] <- "perspectivus"

df_kraemer[which(df_kraemer$species == "cf. amastroides"),"species"] <- "amastroides"

df_kraemer[which(df_kraemer$species == "canaliferus 1"),"species"] <- "canaliferus"
df_kraemer[which(df_kraemer$species == "canaliferus 2"),"species"] <- "canaliferus"

df_kraemer[which(df_kraemer$species == "simrothi 1"),"species"] <- "simrothi"
df_kraemer[which(df_kraemer$species == "simrothi 2"),"species"] <- "simrothi"

df_kraemer[which(df_kraemer$species == "cf. tortuganus"),"species"] <- "tortuganus"

df_kraemer[which(df_kraemer$species == "cf. nux"),"species"] <- "nux"
```



Only keep species with assessed load:

```{r}

# df_kraemer df_merged to only include species that appear in df_load
df_brightness <- df_kraemer %>%
  semi_join(df_load, by = "species")

df_brightness
```




# habitat data


### Habitat data from Kraemer 2019 database

```{r}
df_kraemer <- read.csv("a_data/kraemer2019_brightness_scores.csv")

df_kraemer <- df_kraemer %>% 
  select(location, spp, microhab) %>% 
  rename(id = location,
         species = spp,
         habitat = microhab)

df_kraemer

```


### habitat data from RAD sequences database

```{r}
df_rad <- read_excel("a_data/GPS RAD snails.xlsx")
df_rad <- df_rad %>% 
  select(Sample, Taxon, Microhabitat) %>% 
  rename(id = Sample,
         species = Taxon,
         habitat = Microhabitat)

# Semi-Arboreal is considered arboreal
df_rad <- df_rad %>%
  mutate(habitat = str_replace(habitat, "Semi-Arboreal", "arboreal"))

df_rad <- df_rad %>%
  mutate(habitat = str_to_lower(habitat)) %>% 
  filter(habitat %in% c("arboreal", "terrestrial"))

df_rad
  
```


### Merge data to obtain single habitat database

```{r}
# Remove rows from df_rad that have matching ids in df_kraemer
df_rad_filtered <- anti_join(df_rad, df_kraemer, by = "id")

# Combine df_kraemer with the filtered df_rad
df_habitat <- bind_rows(df_kraemer, df_rad_filtered)

df_habitat
```


Names checks

```{r}
df_habitat[which(df_habitat$species == "albermarlensis"),"species"] <- "albemarlensis"
df_habitat[which(df_habitat$species == "cf. albemarlensis"),"species"] <- "albemarlensis"

df_habitat[which(df_habitat$species == "ustulatus pallescens"),"species"] <- "ustulatus"
df_habitat[which(df_habitat$species == "ustulatus phlegonis"),"species"] <- "ustulatus"
df_habitat[which(df_habitat$species == "ustulatus mahogany"),"species"] <- "ustulatus"

df_habitat[which(df_habitat$species == "invalidus 1"),"species"] <- "invalidus"
df_habitat[which(df_habitat$species == "invalidus 2"),"species"] <- "invalidus"

df_habitat[which(df_habitat$species == "sculpturatus 1"),"species"] <- "sculpturatus"
df_habitat[which(df_habitat$species == "sculpturatus 2"),"species"] <- "sculpturatus"
df_habitat[which(df_habitat$species == "sculpturatus 3"),"species"] <- "sculpturatus"

df_habitat[which(df_habitat$species == "wolfi 1"),"species"] <- "wolfi"
df_habitat[which(df_habitat$species == "wolfi 2"),"species"] <- "wolfi"
df_habitat[which(df_habitat$species == "wolfi 3"),"species"] <- "wolfi"

df_habitat[which(df_habitat$species == "cf. perspectivus"),"species"] <- "perspectivus"
df_habitat[which(df_habitat$species == "perspectivus 1"),"species"] <- "perspectivus"
df_habitat[which(df_habitat$species == "perspectivus 2"),"species"] <- "perspectivus"

df_habitat[which(df_habitat$species == "cf. amastroides"),"species"] <- "amastroides"

df_habitat[which(df_habitat$species == "canaliferus 1"),"species"] <- "canaliferus"
df_habitat[which(df_habitat$species == "canaliferus 2"),"species"] <- "canaliferus"

df_habitat[which(df_habitat$species == "simrothi 1"),"species"] <- "simrothi"
df_habitat[which(df_habitat$species == "simrothi 2"),"species"] <- "simrothi"

df_habitat[which(df_habitat$species == "cf. tortuganus"),"species"] <- "tortuganus"

df_habitat[which(df_habitat$species == "cf. nux"),"species"] <- "nux"

```


Only keep species with assessed load:

```{r}

# Filter df_merged to only include species that appear in df_load
df_habitat <- df_habitat %>%
  semi_join(df_load, by = "species")

df_habitat
```









# vegetation data

### vegetation zone data from RAD sequences database

```{r}
df_rad <- read_excel("a_data/GPS RAD snails.xlsx")

# Standardize the names to lowercase
df_rad <- df_rad %>%
  mutate(`Vegetation Zone` = tolower(`Vegetation Zone`))

# Filter out rows with unwanted values in the Vegetation Zone column
df_rad <- df_rad %>%
  filter(`Vegetation Zone` %in% c("humid", "arid"))

# Rename the column
df_veg <- df_rad %>%
  rename(vegetation_zone = `Vegetation Zone`)
  
```

Names checks

```{r}
#df_veg[which(df_veg$Taxon == "albermarlensis"),] <- "albemarlensis"
df_veg[which(df_veg$Taxon == "cf. albemarlensis"),"Taxon"] <- "albemarlensis"

#df_veg[which(df_veg$Taxon == "ustulatus pallescens"),] <- "ustulatus"
#df_veg[which(df_veg$Taxon == "ustulatus phlegonis"),] <- "ustulatus"
#df_veg[which(df_veg$Taxon == "ustulatus mahogany"),] <- "ustulatus"

df_veg[which(df_veg$Taxon == "invalidus 1"),"Taxon"] <- "invalidus"
df_veg[which(df_veg$Taxon == "invalidus 2"),"Taxon"] <- "invalidus"

df_veg[which(df_veg$Taxon == "sculpturatus 1"),"Taxon"] <- "sculpturatus"
df_veg[which(df_veg$Taxon == "sculpturatus 2"),"Taxon"] <- "sculpturatus"
df_veg[which(df_veg$Taxon == "sculpturatus 3"),"Taxon"] <- "sculpturatus"

df_veg[which(df_veg$Taxon == "wolfi 1"),"Taxon"] <- "wolfi"
df_veg[which(df_veg$Taxon == "wolfi 2"),"Taxon"] <- "wolfi"
df_veg[which(df_veg$Taxon == "wolfi 3"),"Taxon"] <- "wolfi"

df_veg[which(df_veg$Taxon == "cf. perspectivus"),"Taxon"] <- "perspectivus"
df_veg[which(df_veg$Taxon == "perspectivus 1"),"Taxon"] <- "perspectivus"
df_veg[which(df_veg$Taxon == "perspectivus 2"),"Taxon"] <- "perspectivus"

df_veg[which(df_veg$Taxon == "cf. amastroides"),"Taxon"] <- "amastroides"

df_veg[which(df_veg$Taxon == "canaliferus 1"),"Taxon"] <- "canaliferus"
df_veg[which(df_veg$Taxon == "canaliferus 2"),"Taxon"] <- "canaliferus"

df_veg[which(df_veg$Taxon == "simrothi 1"),"Taxon"] <- "simrothi"
df_veg[which(df_veg$Taxon == "simrothi 2"),"Taxon"] <- "simrothi"

df_veg[which(df_veg$Taxon == "cf. tortuganus"),"Taxon"] <- "tortuganus"

df_veg[which(df_veg$Taxon == "cf. nux"),"Taxon"] <- "nux"


```

Only keep species with assessed load:

```{r}

df_veg1 <- df_veg

# Filter df_merged to only include species that appear in df_load
df_veg <- df_veg %>%
  rename(species = Taxon) %>% 
  semi_join(df_load, by = "species")

df_veg
```


## save data

```{r}
#write.csv(df_load, "a_data/processed/df_load")
#write.csv(df_brightness, "a_data/processed/df_brightness")
#write.csv(df_habitat, "a_data/processed/df_habitat")
#write.csv(df_veg, "a_data/processed/df_veg")

df_load <- read.csv("a_data/processed/df_load")
df_brightness <- read.csv("a_data/processed/df_brightness")
df_habitat <- read.csv("a_data/processed/df_habitat")
df_veg <- read.csv("a_data/processed/df_veg")

```



# Plots raw data



## Load by species


```{r}
# Calculate mean nematode load per species
mean_load_per_species <- df_load %>%
  group_by(species) %>%
  summarise(mean_load = mean(nematode_count, na.rm = TRUE)) %>%
  arrange(mean_load)  # Arrange species by mean load in ascending order

# Reorder species factor levels based on mean_load
df_load$species <- factor(df_load$species, levels = mean_load_per_species$species)

# Plot nematode load by species with boxplot and jitter
plot_load_by_species <- ggplot(df_load, aes(x = species, y = nematode_count)) +
  geom_boxplot(outlier.shape = NA) +  # Boxplot without outliers
  geom_jitter(width = 0.2, alpha = 0.5, size = 1) +  # Jittered points
  labs(
    x = "Species",
    y = "Nematode Load"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))  # Rotate x-axis labels
my_theme

plot_load_by_species

```



## Load ~ brightness



Brightness data

```{r}
# Calculate mean brightness per species
mean_brightness <- df_brightness %>%
  group_by(species) %>%
  summarise(mean_brightness = mean(brightness)) %>%
  arrange(mean_brightness)  # Arrange species in ascending order of mean brightness

# Convert species to factor and reorder based on mean_brightness
df_brightness$species <- factor(df_brightness$species, levels = mean_brightness$species)

# Plot brightness by species
plot_brightness <- ggplot(data = df_brightness, aes(x = species, y = brightness)) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 2) +
  geom_point(data = mean_brightness, aes(y = mean_brightness), color = "red", size = 2, shape = 23, fill = "red") +
  labs(
    x = "Species",
    y = "Brightness"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))+
  my_theme

```


Load~brightness

```{r}
# Calculate the mean brightness per species
mean_brightness_per_species <- df_brightness %>%
  group_by(species) %>%
  summarize(mean_brightness = mean(brightness, na.rm = TRUE))

# Calculate the mean nematode counts per species
mean_nematode_per_species <- df_load %>%
  group_by(species) %>%
  summarize(mean_nematode_count = mean(nematode_count, na.rm = TRUE))

# Merge the datasets
merged_data <- inner_join(mean_brightness_per_species, mean_nematode_per_species, by = "species")

# Merge the individual nematode counts with the mean brightness per species
df_load_with_brightness <- df_load %>%
  inner_join(mean_brightness_per_species, by = "species")


# Plot
plot_load_brightness <- ggplot() +
  # Plot individual nematode counts
  geom_point(data = df_load_with_brightness, aes(x = mean_brightness, y = nematode_count), alpha = 0.5, color = "black") +
  # Plot mean nematode counts per species
  geom_point(data = merged_data, aes(x = mean_brightness, y = mean_nematode_count), color = "red", size = 2, shape = 23, fill = "red") +
  # Add dashed LOESS trend line
  geom_smooth(data = df_load_with_brightness, aes(x = mean_brightness, y = nematode_count), method = "loess", se = FALSE, color = "blue", linetype = "dashed") +
  labs(
       x = "Mean Brightness",
       y = "Nematode counts",
       color = "Mean nematode nounts") +
  theme_bw() +
  my_theme

```



## Load ~ habitat

```{r}

# Filter species in df_load that are also present in df_habitat
species_in_habitat <- unique(df_habitat$species)
df_load_filtered <- df_load %>%
  filter(species %in% species_in_habitat)

# Calculate mean nematode load per species
mean_load_per_species <- df_load_filtered %>%
  group_by(species) %>%
  summarise(mean_load = mean(nematode_count, na.rm = TRUE))

# Count individuals in each habitat per species
habitat_count_per_species <- df_habitat %>%
  group_by(species, habitat) %>%
  summarise(count = n(), .groups = 'drop')

# Merge mean load into habitat data
df_habitat_with_load <- habitat_count_per_species %>%
  left_join(mean_load_per_species, by = "species")

# Define colors manually from magma palette
colors <- c("red", "gray")




plot_load_habitat <- ggplot(data = df_habitat_with_load, aes(x = reorder(species, count), y = count, fill = habitat)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_point(aes(y = mean_load), color = "black", shape = 16, size = 2, alpha = 0.9,position = position_dodge(width = 0.9)) +
  labs(
       x = NULL,
       y = "Snail counts",
       fill = "Habitat") +
  scale_y_continuous(
    name = "Snail count",
    sec.axis = sec_axis(~ ., name = "Mean Nematode count")
  ) +
  scale_fill_manual(values = colors) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        axis.title.x = element_blank(),
        legend.position = "top")+
  my_theme


```

## Load ~ vegetation zone

```{r}
# Calculate mean nematode load per species
mean_load_per_species <- df_load_filtered %>%
  group_by(species) %>%
  summarise(mean_load = mean(nematode_count, na.rm = TRUE))

# Count individuals in each vegetation zone per species
veg_count_per_species <- df_veg %>%
  group_by(species, vegetation_zone) %>%
  summarise(count = n(), .groups = 'drop')

# Merge mean load into vegetation zone data
df_veg_with_load <- veg_count_per_species %>%
  left_join(mean_load_per_species, by = "species")

# Define colors manually from magma palette
colors <- c("red", "gray")  # Adjust the number of colors as needed

# Plot
plot_load_veg <- ggplot(data = df_veg_with_load, aes(x = reorder(species, count), y = count, fill = vegetation_zone)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_point(aes(y = mean_load), color = "black", shape = 16, size = 2, alpha = 0.9, position = position_dodge(width = 0.9)) +
  labs(
    x = NULL,
    y = "Snail counts",
    fill = "Vegetation Zone"
  ) +
  scale_y_continuous(
    name = "Snail count",
    sec.axis = sec_axis(~ ., name = "Mean Nematode count")
  ) +
  scale_fill_manual(values = colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        axis.title.x = element_blank(),
        legend.position = "top") +
  my_theme
```



## Load ~ island age

```{r}
back_transform_sqrt <- function(x) {
  return(x^2)  # Square back the transformed value
}
```


```{r}
# Calculate the mean nematode counts per species
mean_load_per_species <- df_load %>%
  group_by(species) %>%
  summarize(mean_load = mean(nematode_count, na.rm = TRUE))

# Calculate the mean island age and area per species
mean_age_area_per_species <- df_load %>%
  group_by(species) %>%
  summarize(mean_island_age = mean(island_age, na.rm = TRUE),
            mean_island_area = mean(island_area, na.rm = TRUE))

# Merge the datasets
merged_data <- inner_join(mean_age_area_per_species, mean_load_per_species, by = "species")

# Merge the individual nematode counts with the mean island age and area per species
df_load_with_age_area <- df_load %>%
  inner_join(mean_age_area_per_species, by = "species")

# Plot
plot_load_age <- ggplot() +
  # Plot individual nematode counts
  geom_point(data = df_load_with_age_area, aes(x = island_age, y = nematode_count), alpha = 0.5, color = "black") +
  # Plot mean nematode counts per species
  geom_point(data = merged_data, aes(x = mean_island_age, y = mean_load), color = "red", size = 2, shape = 23, fill = "red") +
  labs(
    x = "Island age (Myr)",
    y = "Nematode counts",
    color = "Mean nematode nounts"
  ) +
  theme_bw() +
  my_theme

plot_load_area <- ggplot() +
  # Plot individual nematode counts
  geom_point(data = df_load_with_age_area, aes(x = sqrt(island_area), y = nematode_count), alpha = 0.5, color = "black") +
  # Plot mean nematode counts per species
  geom_point(data = merged_data, aes(x = sqrt(mean_island_area), y = mean_load), color = "red", size = 2, shape = 23, fill = "red") +
  labs(
    x = expression("Island area ("~km^2~")"),
    y = "Nematode counts",
    color = "Mean nematode nounts"
  ) +
  scale_x_continuous(labels = back_transform_sqrt)+
  theme_bw() +
  my_theme

```






# Load ~ island

```{r}
# Calculate the mean nematode load per island
mean_load_per_island <- df_load %>%
  group_by(island) %>%
  summarize(mean_load = mean(nematode_count, na.rm = TRUE)) %>%
  arrange(mean_load)

# Order the island factor levels by mean load
df_load <- df_load %>%
  mutate(island = factor(island, levels = mean_load_per_island$island))

# Plot the mean nematode load per island
plot_mean_load_island <- ggplot(df_load, aes(x = island, y = nematode_count)) +
  geom_boxplot(outlier.shape = NA, fill = "skyblue", alpha = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "black", size = 1) +
  stat_summary(fun = mean, geom = "point", color = "red", size = 3, shape = 23, fill = "red") +
  labs(
    x = "Island",
    y = "Nematode Load",
    title = "Nematode Load per Island"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

plot_mean_load_island
```

```{r}

# Calculate the mean load for each island
mean_load_by_island <- df_load %>%
  group_by(island, species) %>%
  summarize(mean_load = mean(nematode_count, na.rm = TRUE))


# Plot with x-axis categories ordered by mean_load value
plot_island_name <- ggplot(mean_load_by_island, aes(x = factor(island), y = mean_load)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.5, size = 2) +  # Add points with jitter
  labs(
       x = NULL,
       y = "Mean nematode counts") +
  theme_bw() +
  my_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot_island_name


```


# Figure


```{r}
plots_load_ecol <- ggarrange(
  
  plot_load_brightness,
  plot_load_habitat,
  plot_load_veg,
  plot_load_age,
  plot_load_area,
  plot_island_name,
  
  ncol = 2,
  nrow = 3,
  
  labels = LETTERS[1:6]

  
)

plots_load_ecol


ggsave("plots_load_ecol.png", height = 19, width = 17)
```












----------------------------------------


# Phylogenetic distance


```{r}
# Read the .nwk file
tree <- read.tree("phylogeny_2022/22Nov_Naesiotus.nwk")
data <- read.nexus("phylogeny_2022/22Nov_Naesiotus.nex")
df_rads <- read_excel("GPS RAD snails.xlsx")

plot(tree)
```


```{r}
# Create a mapping between Sample and Taxon
sample_to_taxon <- setNames(df_rads$Taxon, df_rads$Sample)

# Replace tip labels in the tree with corresponding taxa
tree$tip.label <- sample_to_taxon[tree$tip.label]

plot(tree)
```


compute phylogenetic distances

```{r}
phylo_dist_matrix <- cophenetic.phylo(tree)

```


```{r}
df_load
```

```{r}
# Step 1: Calculate pairwise distances of mean_load values
load_dist_matrix <- as.matrix(dist(df_load$mean_load, method = "euclidean"))
rownames(load_dist_matrix) <- df_load$species
colnames(load_dist_matrix) <- df_load$species

# Step 2: Check and match species names in both matrices
common_species <- intersect(rownames(load_dist_matrix), rownames(phylo_dist_matrix))

common_species <- common_species[!is.na(common_species)]

# Step 3: Subset and reorder matrices
load_dist_matrix_subset <- load_dist_matrix[common_species, common_species]
phylo_dist_matrix_subset <- phylo_dist_matrix[common_species, common_species]

# Step 4: convert into distance objects
load_dist <- as.dist(load_dist_matrix_subset)
phylo_dist <- as.dist(phylo_dist_matrix_subset)

# Step 5: compute an ordination of the matrices

load_pco <- dudi.pco(load_dist, scannf = FALSE, nf = 2, full = TRUE)
phylo_pco <- dudi.pco(phylo_dist, scannf = FALSE, nf = 2, full = TRUE)

```

PCO plots

```{r}
# Create a data frame for ggplot2 (example for load_pco)
load_pco_data <- data.frame(Species = rownames(load_pco$li), 
                            PCoA1 = load_pco$li[, 1])

# Plot using ggplot2
ggplot(load_pco_data, aes(x = Species, y = PCoA1)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Principal Coordinate Analysis (PCoA) of Nematode Load", 
       x = "Species", y = "PCoA 1") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Procrustes

```{r}
protest <- protest(load_pco, phylo_pco)
protest

plot(protest)
```



## models test

```{r}
combined_df
```



```{r}

```










```{r}
# Convert categorical variables to factors
combined_df$microhabitat <- as.factor(combined_df$microhabitat)
combined_df$vegetation_zone <- as.factor(combined_df$vegetation_zone)

# Define priors
fixed_effects_prior <- set_prior("normal(0, 5)", class = "b")
random_effects_prior <- set_prior("cauchy(0, 2)", class = "sd")

# Fit the Bayesian model
bayesian_model <- brm(mean_load ~ microhabitat + vegetation_zone + mean_emergence + area + mean_brightness + (1|species),
                      data = combined_df, 
                      family = gaussian(),
                      prior = c(fixed_effects_prior, random_effects_prior),
                      chains = 4, iter = 2000, warmup = 1000, cores = 4,
                      control = list(adapt_delta = 0.95))

# Summarize the model
summary(bayesian_model)

# Get the marginal effects
marginal_effects <- marginal_effects(bayesian_model)

# Plot the marginal effects
plot(marginal_effects)
```

