---
title: "R Notebook"
---

```{r}
library(dplyr)
library(tidyverse)
library(ape)
library(qiime2R) 
library(igraph)
library(ggraph)
library(ggtree)
library(vegan)
library(gridExtra)
library(ggfortify)
library(FactoMineR) 
library(factoextra)  
library(centiserve)
library(bipartite)
library(readxl)
library(ggpubr)
```




```{r}
# Read the OTU table
otu_table <- read.delim(here::here("C:/Users/alexf/Desktop/PhD_Fuster-Calvo/naesiotus_nematodes/naesiotus_nematodes/data/OTUs/feature_table.tsv"), skip = 1)
colnames(otu_table) <- gsub("\\.", "-", colnames(otu_table))
colnames(otu_table)[1] <- "OTU_ID"
# Correct sample name in OTU table
otu_table <- otu_table %>%
  dplyr::rename_with(~ gsub("XI-2P", "X1-2P", .)) |> 
  dplyr::rename_with(~ gsub("U5-1P", "US-1P-b", .))

# Read your metadata file
metadata <- read.delim(here::here("C:/Users/alexf/Desktop/PhD_Fuster-Calvo/naesiotus_nematodes/naesiotus_nematodes/data/OTUs/metadata.tsv"), skip = 1)
# rename the sample ID column in either dataset to match
colnames(metadata)[1] <- "Sample_ID"
# Correct sample name in metadata
metadata <- metadata %>%
  dplyr::mutate(Sample_ID = gsub("XI-2P", "X1-2P", Sample_ID)) |> 
  dplyr::mutate(Sample_ID = gsub("U5-N", "US-N", Sample_ID)) |> 
  dplyr::mutate(Sample_ID = gsub("U5-1P", "US-1P-b", Sample_ID))

```



checks

```{r}
# Check if all sample IDs in otu_table are present in metadata
missing_samples <- setdiff(colnames(otu_table)[-1], metadata$Sample_ID)  # Exclude OTU_ID column

# Output the missing sample IDs, if any
if (length(missing_samples) > 0) {
  cat("These sample IDs in the OTU table are not found in the metadata:\n")
  print(missing_samples)
} else {
  cat("All sample IDs in the OTU table are present in the metadata.\n")
}

# Check if there are sample IDs in the metadata not present in the OTU table
extra_metadata_samples <- setdiff(metadata$Sample_ID, colnames(otu_table)[-1])

if (length(extra_metadata_samples) > 0) {
  cat("These sample IDs in the metadata are not found in the OTU table - i.e. nematodes werent found in them:\n")
  print(extra_metadata_samples)
} else {
  cat("All sample IDs in the metadata are present in the OTU table.\n")
}

```


```{r}
# Transpose OTU table so that OTUs are columns and samples are rows
otu_table_t <- t(otu_table[-1])  # Remove the first column which is the OTU IDs, then transpose
colnames(otu_table_t) <- otu_table$OTU_ID  # Set OTU IDs as column names
otu_table_t <- as.data.frame(otu_table_t)  # Convert to dataframe
otu_table_t$Sample_ID <- rownames(otu_table_t)  # Add SampleID as a column
```

```{r}
# Check the results
head(rownames(otu_table_t))
head(metadata$Sample_ID)

```

```{r}
# Join OTU table with metadata
otu_metadata <- dplyr::inner_join(metadata, otu_table_t, by = "Sample_ID")

# 1. Remove the 'categorical.1' column
otu_metadata <- otu_metadata %>%
  dplyr::select(-categorical.1)

# 2. Remove rows where 'Sample_ID' starts with "RAR" and save in a separate dataframe
otu_metadata_rar <- otu_metadata %>%
  dplyr::filter(grepl("^RAR", Sample_ID))

otu_metadata <- otu_metadata %>%
  dplyr::filter(!grepl("^RAR", Sample_ID))

# 3. Correctly assign 'naesiotus_sp' by extracting everything before the first dash
otu_metadata <- otu_metadata %>%
  dplyr::mutate(naesiotus_sp = sub("-.*", "", Sample_ID)) |> 
  rename(island = categorical,
         pool_id = categorical.2,
         pos_neg = categorical.3)

# Move the 'naesiotus_sp' column to the second position
otu_metadata <- otu_metadata %>%
  dplyr::select(Sample_ID, naesiotus_sp, everything())

# Verify the changes
head(otu_metadata)

```




What negatives contain nematodes?

```{r}
# Filter for species with OTUs greater than 0 and pos_neg is "negative"
species_with_otus_and_negative <- otu_metadata %>%
  dplyr::filter(pos_neg == "negative") %>%
  dplyr::filter(rowSums(.[, -(1:5)], na.rm = TRUE) > 0)

unique(species_with_otus_and_negative$naesiotus_sp)

paste(length(unique(species_with_otus_and_negative$naesiotus_sp)),"species with negative controls with nematodes")

```


##check rarefraction analysis

```{r}
# Rename metadata columns
otu_metadata_rar <- otu_metadata_rar |> 
  rename(island = categorical,
         pool_id = categorical.2,
         pos_neg = categorical.3)


# Create a column to identify the species based on Sample_ID
otu_metadata_rar <- otu_metadata_rar %>%
  dplyr::mutate(species = ifelse(grepl("^RAR-X", Sample_ID), "unifasciatus", "sp.nov.VolcanWolf_3"))

# Extract OTU columns (all columns that are not metadata)
otu_columns <- otu_metadata_rar %>% 
  dplyr::select(-Sample_ID, -island, -pool_id, -pos_neg, -species)

# Initialize an empty vector to store the cumulative unique OTU counts
cumulative_unique_otus <- numeric(nrow(otu_metadata_rar))

# Split the data by species
species_groups <- split(otu_metadata_rar, otu_metadata_rar$species)

# Initialize an empty list to store cumulative OTU data for each species
cumulative_list <- list()

# Loop through each species and calculate the cumulative unique OTUs
for (species in names(species_groups)) {
  # Subset data for this species
  otu_subset <- species_groups[[species]]
  otu_columns_subset <- otu_columns[rownames(otu_subset), ]
  
  # Initialize variables for cumulative unique OTUs
  unique_otus <- c()
  cumulative_unique_otus <- numeric(nrow(otu_subset))
  
  # Loop through each sample for this species
  for (i in 1:nrow(otu_subset)) {
    current_otus <- colnames(otu_columns_subset)[otu_columns_subset[i, ] > 0]
    unique_otus <- union(unique_otus, current_otus)
    cumulative_unique_otus[i] <- length(unique_otus)
  }
  
  # Store cumulative results for this species
  cumulative_list[[species]] <- data.frame(
    sample_num = 1:nrow(otu_subset),
    cumulative_unique_otus = cumulative_unique_otus,
    species = species
  )
}

# Combine the results from all species into one dataframe
cumulative_df <- do.call(rbind, cumulative_list)

ggplot(cumulative_df, aes(x = sample_num, y = cumulative_unique_otus, linetype = species)) +
  geom_line(size = 0.9) +
  geom_point(size = 2.3, shape = 21, fill = "black") + # Adds points for measurements
  labs(
    x = "Number of snail individuals",
    y = "Cumulative nematode OTUs"
  ) +
  theme_minimal() +
  scale_linetype_manual(values = c("solid", "dashed")) # Optional: specify line types


#ggsave(here::here("output/figures/rarefract_otus.png"), height = 4, width = 6)


```




## Read OTUS tree

```{r}
tree <- read_qza(here::here("C:/Users/alexf/Desktop/PhD_Fuster-Calvo/naesiotus_nematodes/naesiotus_nematodes/data/OTUs/iq-rooted-clustered.qza"))
# Extract the phylogenetic tree from the data
phylo_tree <- tree$data  # This extracts the phylogenetic tree in "phylo" format
plot(phylo_tree)
```
check that OTU tree tip names are present in the created metadata

```{r}
# Extract OTU names from the tree tips
otu_tree_tips <- phylo_tree$tip.label

# Get the OTU IDs from otu_metadata
otu_metadata_ids <- colnames(otu_metadata)[-(1:5)]  # Assuming OTU IDs start from column 6

# Check which OTU names in the tree are present in the otu_metadata
matching_otus <- otu_tree_tips %in% otu_metadata_ids

# Output the number of matching and non-matching OTUs
cat("Number of matching OTUs:", sum(matching_otus), "\n")
cat("Number of non-matching OTUs:", sum(!matching_otus), "\n")

```

Rename OTUS names in the tree and metadata

```{r}
# Combine OTU names from the tree and metadata to ensure all OTUs are included
all_otu_ids <- unique(c(otu_tree_tips, otu_metadata_ids))

# Create a mapping of new labels: OTU1, OTU2, OTU3, ...
otu_mapping <- data.frame(
  original_otu = all_otu_ids,
  new_otu_label = paste0("OTU", seq_along(all_otu_ids)),
  stringsAsFactors = FALSE
)

# View the mapping
head(otu_mapping)
```

Apply the mapping on the tree

```{r}
# Rename the tree tip labels based on the mapping
phylo_tree$tip.label <- otu_mapping$new_otu_label[match(phylo_tree$tip.label, otu_mapping$original_otu)]

# Check the updated tree tip labels
head(phylo_tree$tip.label)

```

Apply the mapping to the metadata

```{r}
# Rename the OTU columns in the metadata table based on the mapping
colnames(otu_metadata)[-(1:5)] <- otu_mapping$new_otu_label[match(colnames(otu_metadata)[-(1:5)], otu_mapping$original_otu)]

# Check the updated column names in the metadata table (optional)
head(colnames(otu_metadata))
```


```{r}
# Verify if the OTU names in the tree and metadata are consistent
all(phylo_tree$tip.label %in% colnames(otu_metadata)[-(1:5)])  # Should return TRUE if consistent
```

Aggregate data per naesiotus species

```{r}
# Aggregate OTU counts by naesiotus_sp, summing across all samples for each species
aggregated_otu_data <- otu_metadata %>%
  dplyr::group_by(naesiotus_sp) %>%
  dplyr::summarise(across(starts_with("OTU"), sum, na.rm = TRUE))

# Check the aggregated data
head(aggregated_otu_data)
head(otu_metadata)
```



# PLOT NETWORK AND NEMATODE PHYLOGENY

```{r}
# Convert the aggregated data to a bipartite matrix format
bipartite_matrix <- as.data.frame(aggregated_otu_data)

# Set rownames as the naesiotus_sp (snail species)
rownames(bipartite_matrix) <- bipartite_matrix$naesiotus_sp

# Remove the naesiotus_sp column from the matrix, leaving only OTU data
bipartite_matrix <- bipartite_matrix[, -1]

# Check the bipartite matrix
head(bipartite_matrix)

# Convert the bipartite matrix to a long format for graph creation
bipartite_long <- bipartite_matrix %>%
  rownames_to_column(var = "species") %>%
  pivot_longer(-species, names_to = "OTU", values_to = "count") %>%
  filter(count > 0)  # Only keep connections with counts greater than 0

# Create an edge list from the long data
edge_list <- bipartite_long %>% dplyr::select(species, OTU)

# Create a bipartite graph
bipartite_network <- graph_from_data_frame(edge_list, directed = FALSE)

# Get the number of species and OTUs
num_species <- length(unique(bipartite_long$species))
num_otus <- length(unique(bipartite_long$OTU))

# Assign bipartite types (TRUE for species, FALSE for OTUs)
V(bipartite_network)$type <- c(rep(TRUE, num_species), rep(FALSE, num_otus))


# Plot the bipartite network
ggraph(bipartite_network, layout = "bipartite") + 
  geom_edge_link(aes(width = 0.5), edge_color = "grey") + 
  geom_node_point(aes(color = factor(type)), size = 5) +  # Color by type (species or OTU)
  theme_void() +
  theme(legend.position = "none") +
  ggtitle("Bipartite Network of Snail Species and OTUs")
```


```{r}
# Create the bipartite graph from the matrix
g <- graph_from_incidence_matrix(as.matrix(bipartite_matrix))

# Assign types to nodes: type=TRUE for OTUs (from columns), type=FALSE for snails (from rows)
V(g)$type <- bipartite_mapping(g)$type

# Check the graph structure (optional)
summary(g)
```

```{r}

phylo_tree_otus <- phylo_tree
# Plot the phylogeny for nematode OTUs
phylo_plot_otus <- ggtree(phylo_tree, layout = "rectangular") +
  geom_tiplab()  # Add tip labels (OTU names)

# View the phylogeny
phylo_plot_otus
```

```{r}

# Create the customized bipartite plot
bipartite_plot <- ggraph(bipartite_network, layout = "bipartite") +
  
  # Change link color to a red gradient
  geom_edge_link(aes(color = ..index..), edge_alpha = 0.8, edge_width = 0.6) +
  scale_edge_color_gradient(low = "pink", high = "red4", guide = "none") +  # Remove edge color legend
  
  # Customize nodes with specific colors, transparency, and black borders
  geom_node_point(aes(shape = factor(V(bipartite_network)$type), 
                      color = factor(V(bipartite_network)$type)), 
                  size = 5, stroke = 1, alpha = 0.7, fill = "white") +
  
  # Customize node colors based on type: snails = #999900, nematodes = #00FFCC
  scale_color_manual(values = c("#999900", "#999900"), guide = "none") +  # Remove color legend
  
  # Show shape legend only
  scale_shape_manual(values = c(16, 17), labels = c("Nematode OTUs", "Naesiotus sp")) +
  
  labs(shape = "Guild") +  # Only show shape legend
  
  theme_void() +  # Clean background with no axes or gridlines
  theme(legend.position = "right")  # Adjust legend position

# Plot the bipartite network
#bipartite_plot

# Save the plot
#ggsave(here::here("output/figures/metanetwork_bipartiteplot.png"), height = 3, width = 12)

```


```{r}
#saveRDS(bipartite_matrix, here::here("output/temporal_output/bipartite_matrix.rds"))
```





# network per island

```{r}
# Step 1: Merge the two datasets by 'naesiotus_sp'
merged_data <- aggregated_otu_data %>%
  left_join(otu_metadata %>% dplyr::select(naesiotus_sp, island), by = "naesiotus_sp")

# Step 1.1: Aggregate the OTU data per snail species
merged_data_persnail <- merged_data %>%
  group_by(naesiotus_sp, island) %>%  # Group by both snail species and island
  summarise(across(starts_with("OTU"), sum, na.rm = TRUE)) %>%  # Sum the OTU columns
  ungroup()


#saveRDS(merged_data_persnail,here::here("output/temporal_output/merged_data_persnail.rds"))

# Step 2: Function to create a bipartite network for each island
create_bipartite_network <- function(island_name, data) {
  # Filter for the specific island
  island_data <- data %>% filter(island == !!island_name)
  
  # Reshape the data to long format to create the edges between snails and OTUs
  edges <- island_data %>%
    pivot_longer(cols = starts_with("OTU"), names_to = "OTU", values_to = "count") %>%
    filter(count > 0) %>%
    dplyr::select(naesiotus_sp, OTU) %>%
    mutate(island = island_name)  # Assign island to edges
  
  # Create a bipartite graph
  g <- graph_from_data_frame(edges, directed = FALSE)
  
  # Get the number of species (snails) and OTUs
  num_species <- length(unique(edges$naesiotus_sp))
  num_otus <- length(unique(edges$OTU))
  
  # Assign bipartite types (TRUE for species, FALSE for OTUs)
  V(g)$type <- c(rep(TRUE, num_species), rep(FALSE, num_otus))
  
  V(g)$island <- island_name # Add island as a node trait
  
  # Assign island attribute to edges
  E(g)$island <- edges$island

  return(g)
}

# Step 3: Function to plot the network with island-colored links
plot_custom_bipartite_network <- function(g, island_name) {
  ggraph(g, layout = "bipartite") +
    # Color edges based on island
    geom_edge_link(aes(color = island), edge_alpha = 0.8, edge_width = 0.6) +
    scale_edge_color_manual(values = island_colors, name = "Island") +  # Use island colors
    
    # Customize nodes with specific colors, transparency, and black borders
    geom_node_point(aes(shape = factor(V(g)$type), 
                        color = factor(V(g)$type)), 
                    size = 5, stroke = 1, alpha = 0.7, fill = "white") +
    
    # Customize node colors based on type: snails = #999900, nematodes = #00FFCC
    scale_color_manual(values = c("TRUE" = "grey30", "FALSE" = "grey30"), guide = "none") +
    
    # Show shape legend only
    scale_shape_manual(values = c(16, 17), labels = c("Nematode OTUs", "Naesiotus sp")) +
    
    labs(shape = "Guild") +  # Only show shape legend
    
    ggtitle(island_name) +  # Add the island name as a title
    
    theme_void() +  # Clean background with no axes or gridlines
    theme(legend.position = "right",  # Adjust legend position
          plot.title = element_text(hjust = 0.5))  # Center the title
}

# Step 4: Create and plot bipartite networks for all islands
islands <- unique(merged_data_persnail$island)
plots <- list()

#saveRDS(islands,here::here("output/temporal_output/islands.rds"))

for (island in islands) {
  g <- create_bipartite_network(island, merged_data_persnail)
  p <- plot_custom_bipartite_network(g, island)
  plots[[island]] <- p
}

# Step 5: Arrange the plots in a grid using ggarrange
ggarrange(plotlist = plots, ncol = 2, nrow = ceiling(length(plots)/2),
          common.legend = TRUE, legend = "top")

#ggsave(here::here("output/figures/islandnetworks_bipartiteplot.png"), height = 12, width = 12)

```

Floreana has included the islet Champion -> I think the clear module I see in Floreana correpsonds to this Champion module.






# COMPUTE CENTRALITY SNAILS



## Naesiotus CENTRALITY IN THE REGIONAL NETWORK

```{r}
# Convert bipartite_matrix to a proper matrix if not already
bipartite_matrix <- as.matrix(bipartite_matrix)

# Define types manually based on the bipartite structure of the matrix
num_snails <- nrow(bipartite_matrix)
num_otus <- ncol(bipartite_matrix)

# Create the bipartite graph
g <- graph_from_biadjacency_matrix(bipartite_matrix, directed = FALSE)

# Set types: FALSE for snail nodes (rows), TRUE for OTU nodes (columns)
V(g)$type <- c(rep(FALSE, num_snails), rep(TRUE, num_otus))

# Filter snail nodes
snail_nodes <- V(g)[V(g)$type == FALSE]

# Compute centrality measures
degree_centrality <- igraph::degree(g, v = snail_nodes)
closeness_centrality <- igraph::closeness(g, v = snail_nodes, normalized = TRUE)
katz_centrality <- katzcent(g, alpha = 0.05, vids = snail_nodes)

# Compute specialization index d' using dfun from bipartite package
dfun_results <- dfun(bipartite_matrix)
d_snails <- dfun_results$d  # Extract d' for the lower trophic level (snails)

# Combine the metrics into a data frame
naesiotus_centralities_regional_df <- data.frame(
  species = names(degree_centrality),
  degree = degree_centrality,
  closeness = closeness_centrality,
  katz_z = katz_centrality,
  d = d_snails
)

# Z-score standardization
naesiotus_centralities_regional_df <- naesiotus_centralities_regional_df %>%
  mutate(
    degree_z = scale(degree),
    closeness_z = scale(closeness),
    katz_z_z = scale(katz_z),
    d_z = scale(d)
  )

# Perform PCA on standardized centrality measures (excluding species column)
pca_result <- PCA(naesiotus_centralities_regional_df[, c("degree_z", "closeness_z", "katz_z_z")], scale.unit = FALSE, graph = FALSE)
pca_centrality <- as.data.frame(pca_result$ind$coord)[, 1]  # First principal component

# Add PCA centrality and z-scored PCA centrality to the data frame
naesiotus_centralities_regional_df <- naesiotus_centralities_regional_df %>%
  mutate(
    PCA_centrality = pca_centrality
  )

# Final centralities data frame
naesiotus_centralities_regional_df

# Save the final data frame
#saveRDS(naesiotus_centralities_regional_df, here::here("output/tables/naesiotus_centralities_regional_df.rds"))

```

```{r}
# Perform PCA using prcomp
pca_model <- prcomp(naesiotus_centralities_regional_df[, c("degree_z", "closeness_z", "katz_z_z")], 
                    center = FALSE, scale. = FALSE)

# Summary to get % of variance explained by each component
pca_summary <- summary(pca_model)
explained_variance <- pca_summary$importance[2, ] * 100  # % variance for each PC

# Plot PCA
autoplot(pca_model, data = naesiotus_centralities_regional_df, 
         loadings = TRUE, loadings.label = TRUE,
         loadings.label.size = 3, color = 'black') +
  labs(
    x = paste0("PC1 (", round(explained_variance[1], 1), "%)"),
    y = paste0("PC2 (", round(explained_variance[2], 1), "%)")
  ) +
  theme_minimal()

#ggsave(here::here("../output/figures/pca_plot_naesiotus.png"), height = 4, width = 4)
```


## CENTRALITY IN THE LOCAL NETWORKS


Gardner island merged with Espanola


```{r}

# Filter out islands with fewer than 2 snail species upfront
valid_islands <- otu_metadata %>%
  mutate(island = ifelse(island == "Gardner", "Espanola", island))  %>%
  group_by(island) %>%
  filter(n_distinct(naesiotus_sp) >= 2) %>%
  pull(island) %>%
  unique()

# Prepare an empty list to store centralities for each island
island_centralities <- list()

# Loop through each valid island
for (island in valid_islands) {
  print(paste("Processing island:", island))  # For debugging purposes
  
  # Filter metadata for the current island
  island_metadata <- otu_metadata %>% filter(island == !!island)
  
  # Get snail species from the current island
  island_snails <- unique(island_metadata$naesiotus_sp)
  
  # Subset bipartite matrix for snails from the current island
  island_matrix <- bipartite_matrix[rownames(bipartite_matrix) %in% island_snails, , drop = FALSE]
  
  # Check if the matrix is valid (non-zero rows and columns)
  if (nrow(island_matrix) == 0 || ncol(island_matrix) == 0 || 
      all(rowSums(island_matrix) == 0) || all(colSums(island_matrix) == 0)) {
    message("Skipping island due to invalid matrix: ", island)
    next
  }
  
  # Create bipartite graph
  num_snails <- nrow(island_matrix)
  num_otus <- ncol(island_matrix)
  g_island <- graph_from_biadjacency_matrix(island_matrix, directed = FALSE)
  
  # Set types: FALSE for snail nodes (rows), TRUE for OTU nodes (columns)
  V(g_island)$type <- c(rep(FALSE, num_snails), rep(TRUE, num_otus))
  
  # Filter snail nodes
  snail_nodes <- V(g_island)[V(g_island)$type == FALSE]
  
  # Compute centralities
  degree_centrality <- igraph::degree(g_island, v = snail_nodes)
  closeness_centrality <- igraph::closeness(g_island, v = snail_nodes, normalized = TRUE)
  katz_centrality <- katzcent(g_island, alpha = 0.05, vids = snail_nodes)
  
  # Compute specialization index d' using dfun from bipartite package
  dfun_results <- dfun(island_matrix)
  d_snails <- dfun_results$d  # Extract raw specialization index (d) for snails
  
  # Combine the metrics into a data frame
  centralities_df <- data.frame(
    species = names(degree_centrality),
    degree = degree_centrality,
    closeness = closeness_centrality,
    katz_z = katz_centrality,
    d = d_snails
  )
  
  # Replace missing values with 0
  centralities_df <- centralities_df %>%
    mutate(across(degree:d, ~ ifelse(is.na(.), 0, .)))
  
  # Standardize centralities (z-scores)
  centralities_df <- centralities_df %>%
    mutate(
      degree_z = scale(degree),
      closeness_z = scale(closeness),
      katz_z_z = scale(katz_z),
      d_z = scale(d)
    )
  
# Perform PCA on standardized centrality measures
  pca_variables <- if (nrow(centralities_df) > 2) {
    c("degree_z", "closeness_z", "katz_z_z")
  } else {
    c("degree_z", "katz_z_z")
  }
  
  if (length(pca_variables) > 1) {
    pca_result <- PCA(centralities_df[, pca_variables], scale.unit = FALSE, graph = FALSE)
    pca_centrality <- as.data.frame(pca_result$ind$coord)[, 1]  # First principal component
  } else {
    pca_centrality <- rep(NA, nrow(centralities_df))
  }

  
  # Add PCA centrality scores and island name to the data frame
  centralities_df <- centralities_df %>%
    mutate(PCA_centrality = pca_centrality, island = island)
  
  # Append to the list
  island_centralities[[island]] <- centralities_df
}

# Combine all island centralities into a single dataframe
naesiotus_centralities_local_df <- bind_rows(island_centralities)

# Save the results
#saveRDS(naesiotus_centralities_local_df, here::here("output/tables/naesiotus_centralities_local_df.rds"))

# View the final dataframe
head(naesiotus_centralities_local_df)


```



# Centrality Nematodes

```{r}
# Convert bipartite_matrix to a proper matrix if not already
bipartite_matrix <- as.matrix(bipartite_matrix)

# Define types manually based on the bipartite structure of the matrix
num_snails <- nrow(bipartite_matrix)
num_otus <- ncol(bipartite_matrix)

# Create the bipartite graph
g <- graph_from_biadjacency_matrix(bipartite_matrix, directed = FALSE)

  # Set types: FALSE for snail nodes (rows), TRUE for OTU nodes (columns)
  V(g)$type <- c(rep(FALSE, num_snails), rep(TRUE, num_otus))
  
  # Filter OTU nodes
  otu_nodes <- V(g)[V(g)$type == TRUE]

# Compute centrality measures
  degree_centrality <- igraph::degree(g, v = otu_nodes)
  closeness_centrality <- igraph::closeness(g, v = otu_nodes, normalized = TRUE)
  katz_centrality <- katzcent(g, alpha = 0.05, vids = otu_nodes)
  
  # Compute specialization index d' using dfun from bipartite package
  dfun_results <- dfun(t(island_matrix))  # Transpose for OTU focus
  d_otus <- dfun_results$d  # Extract raw specialization index (d) for OTUs
  
  # Ensure OTU names match across centrality measures and specialization index
  common_otus <- intersect(names(degree_centrality), names(d_otus))

# Filter centrality measures to include only OTUs with valid d values
degree_centrality <- degree_centrality[common_otus]
closeness_centrality <- closeness_centrality[common_otus]
katz_centrality <- katz_centrality[common_otus]
d_otus <- d_otus[common_otus]

# Combine the metrics into a data frame
otu_centralities_regional_df <- data.frame(
  species = common_otus,
  degree = degree_centrality,
  closeness = closeness_centrality,
  katz_z = katz_centrality,
  d = d_otus
)

# Z-score standardization
otu_centralities_regional_df <- otu_centralities_regional_df %>%
  mutate(
    degree_z = scale(degree),
    closeness_z = scale(closeness),
    katz_z_z = scale(katz_z),
    d_z = scale(d)
  )

# Perform PCA on standardized centrality measures (excluding species column)
pca_result <- PCA(otu_centralities_regional_df[, c("degree_z", "closeness_z", "katz_z_z")], scale.unit = FALSE, graph = FALSE)
pca_centrality <- as.data.frame(pca_result$ind$coord)[, 1]  # First principal component

# Add PCA centrality to the data frame
otu_centralities_regional_df <- otu_centralities_regional_df %>%
  mutate(PCA_centrality = pca_centrality)

# Save the results
#saveRDS(otu_centralities_regional_df, here::here("../output/tables/otu_centralities_regional_df.rds"))

# View the final dataframe
head(otu_centralities_regional_df)

```


```{r}
# Perform PCA using prcomp
pca_model <- prcomp(otu_centralities_regional_df[, c("degree_z", "closeness_z", "katz_z_z")], 
                    center = FALSE, scale. = FALSE)

# Summary to get % of variance explained by each component
pca_summary <- summary(pca_model)
explained_variance <- pca_summary$importance[2, ] * 100  # % variance for each PC

# Plot PCA
autoplot(pca_model, data = otu_centralities_regional_df, 
         loadings = TRUE, loadings.label = TRUE,
         loadings.label.size = 3, color = 'black') +
  labs(
    x = paste0("PC1 (", round(explained_variance[1], 1), "%)"),
    y = paste0("PC2 (", round(explained_variance[2], 1), "%)")
  ) +
  theme_minimal()
```

```{r}
# Ensure Gardner island is merged with Espanola
otu_metadata <- otu_metadata %>%
  mutate(island = ifelse(island == "Gardner", "Espanola", island))

# Filter out islands with fewer than 2 snail species upfront
valid_islands <- otu_metadata %>%
  group_by(island) %>%
  filter(n_distinct(naesiotus_sp) >= 2) %>%
  pull(island) %>%
  unique()

# Prepare an empty list to store centralities for each island
island_otu_centralities <- list()

# Loop through each valid island
for (island in valid_islands) {
  print(paste("Processing island:", island))  # For debugging purposes
  
  # Filter metadata for the current island
  island_metadata <- otu_metadata %>% filter(island == !!island)
  
  # Get snail species from the current island
  island_snails <- unique(island_metadata$naesiotus_sp)
  
  # Subset bipartite matrix for snails from the current island
  island_matrix <- bipartite_matrix[rownames(bipartite_matrix) %in% island_snails, , drop = FALSE]
  
  # Check if the matrix is valid (non-zero rows and columns)
  if (nrow(island_matrix) == 0 || ncol(island_matrix) == 0 || 
      all(rowSums(island_matrix) == 0) || all(colSums(island_matrix) == 0)) {
    message("Skipping island due to invalid matrix: ", island)
    next
  }
  
  # Create bipartite graph
  num_snails <- nrow(island_matrix)
  num_otus <- ncol(island_matrix)
  g_island <- graph_from_biadjacency_matrix(island_matrix, directed = FALSE)
  
  # Set types: FALSE for snail nodes (rows), TRUE for OTU nodes (columns)
  V(g_island)$type <- c(rep(FALSE, num_snails), rep(TRUE, num_otus))
  
  # Filter OTU nodes
  otu_nodes <- V(g_island)[V(g_island)$type == TRUE]
  
  # Compute centrality measures
  degree_centrality <- igraph::degree(g_island, v = otu_nodes)
  closeness_centrality <- igraph::closeness(g_island, v = otu_nodes, normalized = TRUE)
  katz_centrality <- katzcent(g_island, alpha = 0.05, vids = otu_nodes)
  
  # Compute specialization index d' using dfun from bipartite package
  dfun_results <- dfun(t(island_matrix))  # Transpose for OTU focus
  d_otus <- dfun_results$d  # Extract raw specialization index (d) for OTUs
  
  # Ensure OTU names match across centrality measures and specialization index
  common_otus <- intersect(names(degree_centrality), names(d_otus))
  
  # Filter centrality measures to include only OTUs with valid d values
  degree_centrality <- degree_centrality[common_otus]
  closeness_centrality <- closeness_centrality[common_otus]
  katz_centrality <- katz_centrality[common_otus]
  d_otus <- d_otus[common_otus]
  
  # Combine the metrics into a data frame
  centralities_df <- data.frame(
    species = common_otus,
    degree = degree_centrality,
    closeness = closeness_centrality,
    katz_z = katz_centrality,
    d = d_otus
  )
  
  # Replace missing values with 0
  centralities_df <- centralities_df %>%
    mutate(across(degree:d, ~ ifelse(is.na(.), 0, .)))
  
  # Standardize centralities (z-scores)
  centralities_df <- centralities_df %>%
    mutate(
      degree_z = scale(degree),
      closeness_z = scale(closeness),
      katz_z_z = scale(katz_z),
      d_z = scale(d)
    )
  
  # Perform PCA on standardized centrality measures
  pca_variables <- if (nrow(centralities_df) > 2) {
    c("degree_z", "closeness_z", "katz_z_z")
  } else {
    c("degree_z", "katz_z_z")
  }
  
  if (length(pca_variables) > 1) {
    pca_result <- PCA(centralities_df[, pca_variables], scale.unit = FALSE, graph = FALSE)
    pca_centrality <- as.data.frame(pca_result$ind$coord)[, 1]  # First principal component
  } else {
    pca_centrality <- rep(NA, nrow(centralities_df))
  }
  
  # Add PCA centrality scores and island name to the data frame
  centralities_df <- centralities_df %>%
    mutate(PCA_centrality = pca_centrality, island = island)
  
  # Append to the list
  island_otu_centralities[[island]] <- centralities_df
}

# Combine all island centralities into a single dataframe
otu_centralities_local_df <- bind_rows(island_otu_centralities)

# Save the results
#saveRDS(otu_centralities_local_df, here::here("output/tables/otu_centralities_local_df.rds"))

# View the final dataframe
head(otu_centralities_local_df)

```



# Network-ages


## island characteristics

```{r}
df_island <- read_excel(path = here::here("data/19july24_datalab.xlsx"), sheet = "island_characteristics")

df_island$min_emergence <- as.numeric(df_island$min_emergence)
df_island$max_emergence <- as.numeric(df_island$max_emergence)

# Calculate the mean difference between min_emergence and max_emergence
mean_diff <- df_island %>%
  filter(!is.na(min_emergence) & !is.na(max_emergence)) %>%
  mutate(diff = max_emergence - min_emergence) %>%
  summarise(mean_diff = mean(diff, na.rm = TRUE)) %>%
  pull(mean_diff)

# Estimate max_emergence for Champion by adding the mean difference to its min_emergence
df_island <- df_island %>%
  mutate(max_emergence = ifelse(is.na(max_emergence), min_emergence + mean_diff, max_emergence))

# Rename the column "mean_emergence" to "island_age" in df_island
# add an ID column
df_island <- df_island %>% dplyr::rename(island_area = area) |> 
  mutate(island_code = as.numeric(as.factor(island)))


# increase max. emergence of Sants Fe by a small amount - min. and max for this island are the same and this gives error in the model when trying to estimate the approximate age.

df_island <- df_island |> 
  dplyr::mutate(max_emergence = ifelse(island == "Santa_Fe", max_emergence + 0.1, max_emergence))


df_island 
```



```{r}
# Merge Gardner with Espanola
otu_metadata <- otu_metadata %>%
  mutate(island = ifelse(island == "Gardner", "Espanola", island))

# Filter islands with at least two snail species
valid_islands <- otu_metadata %>%
  group_by(island) %>%
  filter(n_distinct(naesiotus_sp) >= 2) %>%
  pull(island) %>%
  unique()

# Prepare data frame to store results
network_statistics <- data.frame()

# Loop through each valid island
for (island in valid_islands) {
  print(paste("Processing island:", island))
  
  # Filter metadata for the current island
  island_metadata <- otu_metadata %>% filter(island == !!island)
  
  # Get snail species for the island
  island_snails <- unique(island_metadata$naesiotus_sp)
  
  # Subset bipartite matrix
  island_matrix <- bipartite_matrix[rownames(bipartite_matrix) %in% island_snails, , drop = FALSE]
  
  # Skip invalid matrices
  if (nrow(island_matrix) == 0 || ncol(island_matrix) == 0 || 
      all(rowSums(island_matrix) == 0) || all(colSums(island_matrix) == 0)) {
    message("Skipping island due to invalid matrix: ", island)
    next
  }
  
  # Create graph
  g <- graph_from_biadjacency_matrix(island_matrix, directed = FALSE)
  
  # Compute modularity
  compute_modularity <- function(g) {
    mod_louvain <- cluster_louvain(g)$modularity
    mod_walktrap <- modularity(cluster_walktrap(g))
    mod_fast_greedy <- modularity(cluster_fast_greedy(g))
    mod_leading_eigen <- cluster_leading_eigen(g)$modularity
    mean_modularity <- mean(c(mod_louvain, mod_walktrap, mod_fast_greedy, mod_leading_eigen), na.rm = TRUE)
    return(mean_modularity)
  }
  modularity <- compute_modularity(g)
  
  # Compute nestedness
  nestedness <- nestednodf(island_matrix)$statistic["NODF"]
  
  # Compute H2 specialization
  H2_specialization <- H2fun(island_matrix)
  
  # Compute mean and SD specialization (d) for snails and OTUs
  snail_d <- dfun(island_matrix)$dprime
  otu_d <- dfun(t(island_matrix))$dprime
  
  mean_snail_d <- mean(snail_d, na.rm = TRUE)
  sd_snail_d <- sd(snail_d, na.rm = TRUE)
  mean_otu_d <- mean(otu_d, na.rm = TRUE)
  sd_otu_d <- sd(otu_d, na.rm = TRUE)
  
  # Append results to data frame
  network_statistics <- rbind(network_statistics, data.frame(
    island = island,
    modularity = modularity,
    nestedness = nestedness,
    H2 = H2_specialization,
    mean_snail_d = mean_snail_d,
    sd_snail_d = sd_snail_d,
    mean_otu_d = mean_otu_d,
    sd_otu_d = sd_otu_d
  ))
}

# Add island age and area to the statistics
df_island <- df_island %>%
  mutate(mean_age = (min_emergence + max_emergence) / 2)

network_statistics <- network_statistics %>%
  left_join(df_island %>% dplyr::select(island, mean_age, island_area), by = "island")

network_statistics

#saveRDS(network_statistics, here::here("output/tables/network_statistics.rds"))
```




```{r}

library(scales)

# Transform age and area in the data
network_statistics <- network_statistics %>%
  mutate(
    sqrt_mean_age = sqrt(mean_age),
    log_island_area = log(island_area)
  )

# Plot 1: Modularity vs. Network Age
plot_age_modularity <- ggplot(network_statistics) +
  geom_point(aes(x = sqrt_mean_age, y = modularity), shape = 16, color = "black") +
  scale_x_continuous(
    trans = "sqrt",
    breaks = scales::trans_breaks("sqrt", function(x) x^2),
    labels = scales::trans_format("sqrt", math_format(.x^2))
  ) +
  labs(
    x = "Mean island age (My)",
    y = "Modularity"
  ) +
  theme_minimal()

# Plot 2: Modularity vs. Island Area
plot_area_modularity <- ggplot(network_statistics) +
  geom_point(aes(x = log_island_area, y = modularity), shape = 16, color = "black") +
  scale_x_continuous(
    trans = "log10",
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", math_format(10^.x))
  ) +
  labs(
    x = expression("Island area (km"^2*")"),
    y = "Modularity"
  ) +
  theme_minimal()

# Plot 3: Nestedness vs. Network Age
plot_age_nestedness <- ggplot(network_statistics) +
  geom_point(aes(x = sqrt_mean_age, y = nestedness), shape = 16, color = "black") +
  scale_x_continuous(
    trans = "sqrt",
    breaks = scales::trans_breaks("sqrt", function(x) x^2),
    labels = scales::trans_format("sqrt", math_format(.x^2))
  ) +
  labs(
    x = "Mean island age (My)",
    y = "Nestedness (NODF)"
  ) +
  theme_minimal()

# Plot 4: Nestedness vs. Island Area
plot_area_nestedness <- ggplot(network_statistics) +
  geom_point(aes(x = log_island_area, y = nestedness), shape = 16, color = "black") +
  scale_x_continuous(
    trans = "log10",
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", math_format(10^.x))
  ) +
  labs(
    x = expression("Island area (km"^2*")"),
    y = "Nestedness (NODF)"
  ) +
  theme_minimal()

# Plot 5: H2 vs. Network Age
plot_age_H2 <- ggplot(network_statistics) +
  geom_point(aes(x = sqrt_mean_age, y = H2), shape = 16, color = "black") +
  scale_x_continuous(
    trans = "sqrt",
    breaks = scales::trans_breaks("sqrt", function(x) x^2),
    labels = scales::trans_format("sqrt", math_format(.x^2))
  ) +
  labs(
    x = "Mean island age (My)",
    y = "H2 specialization"
  ) +
  theme_minimal()

# Plot 6: H2 vs. Island Area
plot_area_H2 <- ggplot(network_statistics) +
  geom_point(aes(x = log_island_area, y = H2), shape = 16, color = "black") +
  scale_x_continuous(
    trans = "log10",
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", math_format(10^.x))
  ) +
  labs(
    x = expression("Island area (km"^2*")"),
    y = "H2 specialization"
  ) +
  theme_minimal()

# Mean Specialization (d') vs. Mean Island Age 
plot_age_mean_d <- ggplot(network_statistics) +
  geom_point(aes(x = sqrt_mean_age, y = mean_snail_d), shape = 16, color = "#009E73") +  # Snails
  # geom_ribbon(aes(x = sqrt_mean_age, ymin = mean_snail_d - sd_snail_d, ymax = mean_snail_d + sd_snail_d), 
  #             fill = "#009E73", alpha = 0.3) +  # Ribbon for Snails
  geom_point(aes(x = sqrt_mean_age, y = mean_otu_d), shape = 2, color = "#E69F00") +  # Nematodes
  # geom_ribbon(aes(x = sqrt_mean_age, ymin = mean_otu_d - sd_otu_d, ymax = mean_otu_d + sd_otu_d), 
  #             fill = "#E69F00", alpha = 0.3) +  # Ribbon for Nematodes
  scale_x_continuous(
    trans = "sqrt",
    breaks = scales::trans_breaks("sqrt", function(x) x^2),
    labels = scales::trans_format("sqrt", math_format(.x^2))
  ) +
  labs(
    x = "Mean island age (My)",
    y = "Mean specialization (d')"
  ) +
  theme_minimal()

# Mean Specialization (d') vs. Island Area 
plot_area_mean_d <- ggplot(network_statistics) +
  geom_point(aes(x = log_island_area, y = mean_snail_d), shape = 16, color = "#009E73") +  # Snails
  # geom_ribbon(aes(x = log_island_area, ymin = mean_snail_d - sd_snail_d, ymax = mean_snail_d + sd_snail_d), 
  #             fill = "#009E73", alpha = 0.3) +  # Ribbon for Snails
  geom_point(aes(x = log_island_area, y = mean_otu_d), shape = 2, color = "#E69F00") +  # Nematodes
  # geom_ribbon(aes(x = log_island_area, ymin = mean_otu_d - sd_otu_d, ymax = mean_otu_d + sd_otu_d), 
  #             fill = "#E69F00", alpha = 0.3) +  # Ribbon for Nematodes
  scale_x_continuous(
    trans = "log10",
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", math_format(10^.x))
  ) +
  labs(
    x = expression("Island area (km"^2*")"),
    y = "Mean specialization (d')"
  ) +
  theme_minimal()



ggarrange(
  plot_age_modularity,
  plot_area_modularity,
  plot_age_nestedness,
  plot_area_nestedness,
  plot_age_H2,
  plot_area_H2,
  plot_age_mean_d,
  plot_area_mean_d,
  ncol = 2,
  nrow = 4,
  common.legend = TRUE,
  labels = LETTERS[1:8]
)


#ggsave(here::here("output/figures/network_characteristics.png"), height = 12, width = 8)

```

```{r}
# Fit the model with sqrt_mean_age
nestedness_model_sqrt <- lm(modularity ~ log_island_area, data = network_statistics)

# Summarize the results
summary(nestedness_model_sqrt)

```

```{r}
nestedness_model_sqrt <- lm(mean_snail_d ~ log_island_area  , data = network_statistics)
summary(nestedness_model_sqrt)
```
```{r}
plot_area_modularity <- ggplot(network_statistics, aes(x = log_island_area, y = modularity)) +
  geom_point(shape = 16, color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "black", size = 0.8) +  # Add regression line
  scale_x_continuous(
    trans = "log10",
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", math_format(10^.x))
  ) +
  labs(
    x = expression("Island area (km"^2*")"),
    y = "Modularity"
  ) +
  theme_classic()
plot_area_modularity

ggsave("test_modularity_area.png", height = 5, width = 5)

```

```{r}
ggplot(network_statistics, aes(x = log_island_area, y = modularity)) +
  geom_point(shape = 16, color = "black") +
  geom_text(aes(label = island), hjust = -0.2, vjust = -0.5, size = 3) +  # Add island names
  geom_smooth(method = "lm", se = FALSE, color = "black", size = 0.8) +  # Add regression line
  scale_x_continuous(
    trans = "log10",
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", math_format(10^.x))
  ) +
  labs(
    x = expression("Island area (km"^2*")"),
    y = "Modularity"
  ) +
  theme_classic()

```

