---
title: "R Notebook"
output: html_notebook
---


```{r setup-packages}
#| message: false
#| warning: false
library(tidyverse)
library(readxl)
library(ggpubr)
library(phytools)
library(ape)
library(ggtree)

```

```{r}
tree <- read.newick(here::here("BAMM/cleannames_tree_doublenames_041024.newick"))
```

```{r}
species_names_df_raw <- read.csv(here::here("data/species_names_checks.csv"))

# we're dropping the species with no name in this database. 
# this indicates morphospecies which have no formal name
species_names_df <- species_names_df_raw %>% 
  dplyr::rename(species = annotated_name)  %>% 
  filter(species_name != "") |> 
  mutate(species_code = as.numeric(factor(species_name)))

species_code_df <- species_names_df |> 
  select(species_name, species_code) |> 
  distinct() |> 
  arrange(species_code)

species_code_df
```

```{r}

# load
df_load <- read_excel(path = here::here("data/18sept24_datalab.xlsx"), sheet = 3)
df_load$nematode_count[which(df_load$nematode_count == ">100")] <- "100"
df_load$nematode_count <- as.numeric(df_load$nematode_count)

df_load <- df_load %>% 
  select(id, spp, nematode_count, island) %>% 
  dplyr::rename(species = spp)%>%
  left_join(species_names_df |> select(species, species_code,species_name),
            by = "species")
```

Compute mean and se load per species

```{r}
df_meanload <- df_load %>%
  group_by(species_name) %>%
  summarise(n = n(),
            mean_load = mean(nematode_count, na.rm = TRUE),
            se_load = sd(nematode_count, na.rm = TRUE)/sqrt(length(nematode_count)), 
            sd_load = sd(nematode_count, na.rm = TRUE)) %>%
  arrange(desc(mean_load))

df_meanload
```

```{r}
meanload<-setNames(df_meanload$mean_load, df_meanload$species_name)
s.error<-setNames(df_meanload$se_load, df_meanload$species_name)
```


### diversity data

```{r}
faith_pd_table <- read.delim(here::here("../sequencing/OTUs clustered/alpha-diversity.tsv"), 
                             header = TRUE, 
                             row.names = 1)

faith_pd_table <- rownames_to_column(faith_pd_table, var = "species")
faith_pd_table

```



```{r}
# correct species names
faith_pd_table <- faith_pd_table %>%
  left_join(species_names_df |> select(species, species_code,species_name),
            by = "species") 
```

```{r}
faith_pd <- setNames(faith_pd_table$faith_pd, faith_pd_table$species_name)
```


# LOAD

Prune tree, discarting those species not assessed for load

```{r}
pruned.tree_load<-drop.tip(tree, setdiff(tree$tip.label, df_meanload$species_name))

plot(pruned.tree_load)
```


Also discard those species in load that are not present in the tree

```{r}
# Filter data to include only species present in the phylogenetic tree
tree_species <- pruned.tree_load$tip.label
meanload_filtered <- meanload[names(meanload) %in% tree_species]
s.error_filtered <- s.error[names(s.error) %in% tree_species]

# Verify data alignment
match_indices <- match(tree_species, names(meanload_filtered))
meanload_filtered <- meanload_filtered[match_indices]
s.error_filtered <- s.error_filtered[match_indices]
```




# Plot tree and trait values

```{r}
set.seed(42)

map <- contMap(pruned.tree_load,meanload_filtered,
        res = 800,
        lwd = 4,
        outline = FALSE,
        sig = 0)

# Adjust the color scale limits
map$lims <- c(0, max(meanload_filtered, na.rm = TRUE)) 

# Reverse the color scale
map_load <- setMap(map, invert = TRUE,lwd = 4)

png(here::here("output/figures/phylo/phylomap_load.png"), width = 3600, height = 4800, res = 600)
plot(map_load,  outline = FALSE, lwd = 6)
dev.off()

```


# Blomberg's K

```{r}
K_nemaload <- phylosig(pruned.tree_load, meanload_filtered, method = 'K', test = TRUE, se = s.error_filtered)
K_nemaload
plot.phylosig(K_nemaload)

png(here::here("output/figures/phylo/k_test_load.png"), width = 2800, height = 2600, res = 600) 
# Create the plot
plot.phylosig(K_nemaload)

# Close the graphics device and save the file
dev.off()

```

# Lambda

```{r}

lambda_nemaload <- phylosig(pruned.tree_load, meanload_filtered, method = 'lambda', test = TRUE) 
lambda_nemaload
plot.phylosig(lambda_nemaload)

```

# diversity

Prune tree, discarting those species not assessed for load

```{r}
pruned.tree_div<-drop.tip(tree, setdiff(tree$tip.label, faith_pd_table$species_name))

plot(pruned.tree_div)
```


Also discard those species in load that are not present in the tree

```{r}
# Filter data to include only species present in the phylogenetic tree
tree_species <- pruned.tree_div$tip.label
faith_pd_filtered <- faith_pd[names(faith_pd) %in% tree_species]

# Verify data alignment
match_indices <- match(tree_species, names(faith_pd_filtered))
faith_pd_filtered <- faith_pd_filtered[match_indices]
```




# Plot tree and trait values

```{r}
set.seed(42)

map <- contMap(pruned.tree_div,faith_pd_filtered,
        res = 800,
        lwd = 4,
        outline = FALSE,
        sig = 0)

map$lims <- c(0, max(faith_pd_filtered, na.rm = TRUE)) 

# Reverse the color scale
map_div <- setMap(map, invert = TRUE,lwd = 4,border = NA)

png(here::here("output/figures/phylo/phylomap_div.png"), width = 3600, height = 4800, res = 600)
plot(map_div,  outline = FALSE, lwd = 6)
dev.off()

```
# Blomberg's K

```{r}
K_nemadiv <- phylosig(pruned.tree_div, faith_pd_filtered, method = 'K', test = TRUE)
K_nemadiv


png(here::here("output/figures/phylo/k_test_div.png"), width = 2800, height = 2600, res = 600) 
# Create the plot
plot.phylosig(K_nemadiv)

# Close the graphics device and save the file
dev.off()

```

# Lambda

```{r}

lambda_nemadiv <- phylosig(pruned.tree_div, faith_pd_filtered, method = 'lambda', test = TRUE) 
lambda_nemadiv
plot.phylosig(lambda_nemadiv)

```