---
title: "R Notebook"
---


The RAD sequencing showed that there are several cryptic species and the taxonomy of the group needs to be revisited.

When sampling nematode interactions in the dry collection, the snail species' names were annotated according to the classic taxonomy and doesnt follow the level of detailed differenciation reconstructed by the rad sequencing tree. 

However, for those species for which cryptic lineages are identified and now appear as different nodes in the tree, we can approximate the taxa identity that was assessed for the load based on the location of the sampled populations. I therefore discard those lineages among a cryptic species complex for which I dont measure individuals of its location (based on individuals' ID):

- **Invalidus 2**
- **Wolfi 2, Wolfi 3**
- **sculpturatus 2, sculpturatus 3**
- **Simrothi 2**

(guess) decision justified by proximity of sampling location:

- **Perspectivus**: perspectivus 1 dicarded. I only measure perspectivus from 2005 and locations SL05-01 and SL05-02. The closest to this location are cf. perspectivus and perspectivus 2, so I keep these.


A completely arbitrary decision:

- **Canaliferous**: canaliferous individuals measured are not from neither canaliferous 1 or canaliferous 2 areas, so it can be assigned to any of them. 

I need to repeat analysis with different options for the decision on canalifrous (one discarding canaliferous 1 and another canaliferous 2) to make sure it doesnt impact the overall results.


```{r setup-packages}
#| message: false
#| warning: false
library(tidyverse)
library(readxl)
library(ggpubr)
library(phytools)
library(ape)
library(ggtree)

```




# Generate tree

### tree from raw rad seq data

```{r}

#tree <- read.nexus(here::here("data/phylogeny_2024/Species_tree_labelled.nex"))
tree <- read.newick(here::here("data/phylogeny_2024/tree2024.nwk"))
df_rads <- read_excel(here::here("data/phylogeny_2024/New_tree_taxa_28Aug24.xlsx"))
species_names_df_raw <- read.csv(here::here("data/species_names_checks.csv"))

species_names_df <- species_names_df_raw %>% 
  rename(species = annotated_name)  %>% 
  filter(species_name != "")

tree$tip.label <- gsub("'", "", tree$tip.label)

tree_data <- fortify(tree)
ggtree(tree) +
  geom_tiplab() +  # Add tip labels for species
  theme_tree() +  # Remove legend
  xlim(0, max(tree_data$x) + 0.3)  

```


Eliminate accents

```{r}

tree$tip.label[grepl("sp._nov._Volcán_Wolf_1", tree$tip.label)] <- "sp. nov. Volcan Wolf 1"
tree$tip.label[grepl("sp._nov._Volcán_Wolf_2", tree$tip.label)] <- "sp. nov. Volcan Wolf 2"
tree$tip.label[grepl("sp._nov._Volcán_Wolf_3", tree$tip.label)] <- "sp. nov. Volcan Wolf 3"
tree$tip.label[which(tree$tip.label == "sp._nov._pinzón_1")]<- "sp. nov. pinzon 1"

tree_data <- fortify(tree)
ggtree(tree) +
  geom_tiplab() +  # Add tip labels for species
  theme_tree() +  # Remove legend
  xlim(0, max(tree_data$x) + 0.3) 

```


```{r}
# Create a named vector where names are `annotated_name` and values are `species_name`
name_mapping <- setNames(species_names_df_raw$species_name, species_names_df_raw$annotated_name)

# Initialize a logical vector to track which labels were changed
changed_labels <- rep(FALSE, length(tree$tip.label))

# Loop through the tip labels and apply the conditions
tree$tip.label <- sapply(tree$tip.label, function(label) {
  if (label %in% names(name_mapping)) {
    # Get the corresponding species_name
    new_name <- name_mapping[[label]]
    
    # Condition 2: If species_name is blank, keep the current label
    if (new_name == "") {
      return(label)
    } else {
      # Condition 3: If species_name is not blank, change the label and mark it as changed
      changed_labels[tree$tip.label == label] <<- TRUE
      return(new_name)
    }
  } else {
    # Condition 1: If label is not found in annotated_name, keep the current label
    return(label)
  }
})

# Ensure tree_data is created from the tree
tree_data <- fortify(tree)

# Apply `changed_labels` only to the tips (i.e., `isTip == TRUE`)
tip_data <- tree_data %>% filter(isTip)

# Make sure `changed_labels` is only as long as the tip labels
tip_data$changed_labels <- changed_labels

set.seed(42)

# Plot the tree using only the tip data for labeling
ggtree(tree) +
  geom_tiplab(aes(label = label, color = changed_labels), data = tip_data) +  # Only apply to tips
  scale_color_manual(values = c("black", "red")) +  # Black for unchanged, red for changed
  theme_tree() +
  xlim(0, max(tree_data$x) + 0.3)  # Adjust x-axis limits if necessary



```

Tree formatting: delete duplicate names and homogenize names format

```{r}
# Identify the positions of the duplicates in the tip labels
amastroides_tips <- which(tree$tip.label == "amastroides")
perspectivus_tips <- which(tree$tip.label == "perspectivus")

# Append "delete" to one instance of each duplicated name
tree$tip.label[amastroides_tips[1]] <- "amastroidesdelete"
tree$tip.label[perspectivus_tips[1]] <- "perspectivusdelete"

# Drop the tips with "delete" in their names
tree <- drop.tip(tree, c("amastroidesdelete", "perspectivusdelete"))

# Apply modifications to `tip_data$label`
tip_data <- tip_data %>%
  mutate(
    label = gsub("_", "-", label),    # Replace underscores with hyphens (if needed)
    label = gsub("\\.", "", label),   # Remove periods
    label = gsub("-", "", label),     # Remove hyphens after previous step (to fully remove)
    label = gsub(" ", "", label)      # Remove spaces
  )


set.seed(42)

# Plot the tree using only the tip data for labeling
ggtree(tree) +
  geom_tiplab(aes(label = label, color = changed_labels), data = tip_data) +  # Only apply to tips
  scale_color_manual(values = c("black", "red")) +  # Black for unchanged, red for changed
  theme_tree() +
  xlim(0, max(tree_data$x) + 0.3)  # Adjust x-axis limits if necessary

#write.tree(tree, file = here::here("BAMM/cleannames_tree_doublenames_041024.txt"))
#write.tree(tree, file = here::here("BAMM/cleannames_tree_doublenames_041024.newick"))
```


## Add island information

```{r}
df_load_raw <- read_excel(path = here::here("data/18sept24_datalab.xlsx"), sheet = 3)

# Join df_load_raw with species_names_df to correct species names
df_corrected <- df_load_raw %>%
  left_join(species_names_df, by = c("spp" = "species")) %>%
  # Replace spp with the corrected species_name from species_names_df
  mutate(corrected_species = species_name) %>%
  # dplyr::select relevant columns and remove duplicates
  dplyr::select(corrected_species, island) %>%
  distinct()

# Rename columns for clarity
df_corrected <- df_corrected %>%
  rename(species = corrected_species)

# Filter to get only unique species-island combinations
species_island_table <- df_corrected %>%
  distinct(species, island) 

# Apply the same formatting transformations to the species names
species_island_table <- species_island_table %>%
  mutate(
    species = gsub("_", "-", species), # Replace underscores with hyphens
    species = gsub("\\.", "", species), # Remove periods
    species = gsub("-", "", species),    # Remove hyphens
     species = gsub(" ", "", species)  # Remove spaces
  )

species_island_table

```


# Plot species sampled

```{r}
# List of species to color in red (from species_island_table$species)
species_to_color <- species_island_table$species

# Plot the tree with ggtree
p <- ggtree(tree) +
  geom_tiplab(aes(label = label, color = label %in% species_to_color), 
              hjust = -0.2,          # Adjusts horizontal spacing
              size = 2.5,            # Adjusts the size of the labels
              angle = 0) +           # Rotates labels to horizontal
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black")) +
  theme_tree2() +
  xlim(NA, max(node.depth.edgelength(tree)) * 1.2) # Extends x-axis limit for more spacing

# Display the plot
print(p)
```

# Plot island clades

```{r}

# Create a data frame from tree tip labels to use for merging
tree_data <- data.frame(label = tree$tip.label)

tree_data <- tree_data %>%
  left_join(species_island_table, by = c("label" = "species")) %>%
  mutate(island = factor(island, levels = c(unique(na.omit(island)), "Not Assigned")))

# Replace NA values in island with "Not Assigned"
tree_data$island[is.na(tree_data$island)] <- "Not Assigned"

# Define custom colors for islands, with "Not Assigned" as white
island_colors <- c(
  viridis::viridis_pal(option = "turbo")(length(unique(tree_data$island)) - 1), # Turbo colors for islands
  "white"  # White for "Not Assigned"
)
names(island_colors) <- levels(tree_data$island)

# Plot the tree with points colored by island
p <- ggtree(tree) %<+% tree_data +
  geom_tiplab(hjust = -0.2, size = 2.5) +
  geom_tippoint(aes(color = island), size = 3) +  # Adds colored points at tips based on island
  scale_color_manual(values = island_colors) +  # Apply custom color palette
  theme_tree2() +
  xlim(NA, max(node.depth.edgelength(tree)) * 1.4) + # Extend x-axis for spacing
  theme(legend.position = "right")

# Display the plot
print(p)

#ggsave("tree_islandclades.png", height = 10, width = 8)
```



# combine both

```{r}

library(ggnewscale)


tree_data <- data.frame(label = tree$tip.label)

# List of species to color in red
species_to_color <- species_island_table$species

# Merge the tree data with the island information
tree_data <- tree_data %>%
  left_join(species_island_table, by = c("label" = "species")) %>%
  mutate(island = factor(island, levels = c(unique(na.omit(island)), "Not Assigned")),
         is_red = label %in% species_to_color)  # Indicates which labels should be red

# Replace NA values in island with "Not Assigned"
tree_data$island[is.na(tree_data$island)] <- "Not Assigned"

# Define custom colors for islands, with "Not Assigned" as white
island_colors <- c(
  viridis::viridis_pal(option = "turbo")(length(unique(tree_data$island)) - 1), # Turbo colors for islands
  "white"  # White for "Not Assigned"
)
names(island_colors) <- levels(tree_data$island)

# Plot the tree with red-colored tip labels and points for islands
p <- ggtree(tree) %<+% tree_data +
  geom_tiplab(aes(label = label, color = is_red), 
              hjust = -0.2, size = 2.5, show.legend = FALSE) + # Red for specific species
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"), guide = "none") +  # Red for matching species, black otherwise
  new_scale_color() +  # Allows separate color scale for the island points
  geom_tippoint(aes(color = island), size = 3) +  # Island-colored points
  scale_color_manual(values = island_colors, name = "Island") +  # Apply custom island color palette
  theme_tree2() +
  xlim(NA, max(node.depth.edgelength(tree)) * 1.4) + # Extend x-axis for spacing
  theme(legend.position = "right")

# Display the plot
print(p)

#ggsave("tree_islandclades.png", height = 10, width = 8)
```

sculpturatus2 -> Santiago
alethorhytidus -> Santa_Cruz
chatoblack -> Santa_Cruz
cfwolfi -> Santa_Cruz
wolfi2 -> Santa_Cruz
wolfi3 -> Santa_Cruz
littleBlackGarrapatero -> Santa_Cruz
sculpturatus3 -> Santiago
cfhirsutus -> Santa_Cruz
SC20070101 -> Santa_Cruz
spnovDarwin2 -> Isabela
perrus 1 - Isabela
spnovAlcedo1 -> Isabela
hemaerodes -> Isabela
sp nov Volcan Wolf 1 -> Isabela
sp nov Volcan Wolf 2 -> Isabela
sp nov Volcan Wolf 3 -> Isabela
elaeodes -> Isabela
simrothi2 -> Isabela
Brantsnail -> Isabela
spnovPinta1 -> Pinta
sp nov Pinta 2 -> Pinta
jacobi -> Santiago
sp nov Pinzon 1 -> Pinzon
perspectivus1 -> San_Cristobal
canaliferus2 -> San_Cristobal
cfbasiplicatus -> San_Cristobal
achatellinus -> San_Cristobal
SF15020104 -> Santa_Fe
SF15020103 -> Santa_Fe
SF15020101 -> Santa_Fe
caryonis -> Floreana
cfinvalidus -> Floreana
cfnux -> Floreana
invalidus2 -> Floreana
sulcatus -> Floreana
Brownguys -> Floreana

```{r}
island_colors <- c(
  "Santa_Cruz" = "#66c2a5",
  "San_Cristobal" = "#fc8d62",
  "Santiago" = "#8da0cb",
  "Isabela" = "#e78ac3",
  "Fernandina" = "#a6d854",
  "Floreana" = "#ffd92f",
  "Rabida" = "#e5c494",
  "Santa_Fe" = "#b3b3b3",
  "Espanola" = "#d73027",
  "Pinzon" = "#4575b4",
  "Pinta" = "#74add1",
  "Champion" = "#f46d43",
  "Gardner" = "#3288bd",
  "Not Assigned" = "grey45"
)
```


```{r}
# Add missing species-island mappings
missing_species <- data.frame(
  species = c("sculpturatus2", "alethorhytidus", "chatoblack", "cfwolfi", "wolfi2", 
              "wolfi3", "littleBlackGarrapatero", "sculpturatus3", "cfhirsutus", 
              "SC20070101", "spnovDarwin2", "perrus 1", "spnovAlcedo1", 
              "hemaerodes", "spnovVolcanWolf1", "spnovVolcanWolf2", 
              "spnovVolcanWolf3", "elaeodes", "simrothi2", "Brantsnail", 
              "spnovPinta1", "sp nov Pinta 2", "jacobi", "spnovpinzon1", 
              "perspectivus1", "canaliferus2", "cfbasiplicatus", "achatellinus", 
              "SF15020104", "SF15020103", "SF15020101", "caryonis", "cfinvalidus", 
              "cfnux", "invalidus2", "sulcatus", "Brownguys"),
  island = c("Santiago", "Santa_Cruz", "Santa_Cruz", "Santa_Cruz", "Santa_Cruz", 
             "Santa_Cruz", "Santa_Cruz", "Santiago", "Santa_Cruz", "Santa_Cruz", 
             "Isabela", "Fernandina", "Isabela", "Isabela", "Isabela", "Isabela", 
             "Isabela", "Isabela", "Isabela", "Isabela", "Pinta", "Pinta", 
             "Santiago", "Pinzon", "San_Cristobal", "San_Cristobal", 
             "San_Cristobal", "San_Cristobal", "Santa_Fe", "Santa_Fe", 
             "Santa_Fe", "Floreana", "Floreana", "Floreana", "Floreana", 
             "Floreana", "Floreana")
)

# Update the `species_island_table`
species_island_table <- species_island_table %>%
  bind_rows(missing_species)

```


```{r}
tree_data <- data.frame(label = tree$tip.label) %>%
  left_join(species_island_table, by = c("label" = "species")) %>%
  mutate(
    island = factor(island, levels = names(island_colors)),
    is_red = label %in% species_to_color  # TRUE for sampled species
  )

# Replace any missing values for `island` (just in case)
tree_data <- tree_data %>%
  mutate(island = replace_na(island, "Not Assigned"))



```


```{r}

# Define a robust Mode function
Mode <- function(x, na.rm = FALSE) {
  if (na.rm) x <- x[!is.na(x)]
  if (length(x) == 0) return(NA)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}



tree_islands <- tree_data %>%
  dplyr::select(label, island, is_red) %>%
  right_join(as_tibble(tree), by = c("label" = "label")) %>%
  mutate(island = if_else(is.na(island), "Not Assigned", as.character(island)))

# Directly assign "Pinta" to the specific parent node
tree_islands <- tree_islands %>%
  mutate(island = ifelse(parent == 93, "Pinta", island))


tree_islands <- tree_islands %>%
  mutate(is_red = ifelse(is.na(is_red), FALSE, is_red))  # Ensure no NA values




```



```{r}

set.seed(42)  # Set seed for reproducibility
p <- ggtree(tree) %<+% tree_islands +
  # Color branches by island
  geom_tree(aes(color = island), size = 0.8) +  
  scale_color_manual(
    values = island_colors, 
    name = "Island", 
    breaks = setdiff(names(island_colors), "Not Assigned")  # Exclude "Not Assigned" from legend
  ) +  # Use island colors for branches
  
  # Explicitly set tip labels to black and align them
  geom_tiplab(aes(label = label), 
              color = "black",  # Force labels to be black
              align = TRUE,  # Align all labels
              offset = 0.01,  # Add space between branches and labels
              size = 2.5, show.legend = FALSE) +  # Label size
  
  # Color tip points by island
  geom_tippoint(aes(color = island), size = 3) +  
  scale_color_manual(
    values = island_colors, 
    name = "Island", 
    breaks = setdiff(names(island_colors), "Not Assigned"),  # Exclude "Not Assigned" from legend
    guide = "none"
  ) +  # Use island colors again for tips
  
  theme_tree2() +
  xlim(NA, max(node.depth.edgelength(tree)) * 1.45) +  # Extend x-axis for labels
  theme(legend.position = "right")

p



#ggsave(here::here("output/figures/phylo/phylo_byisland.png"), height = 10, width = 8)
```

### compute species code id

```{r}
species_names_df_raw <- read.csv(here::here("data/species_names_checks.csv"))

# we're dropping the species with no name in this database. 
# this indicates morphospecies which have no formal name
species_names_df <- species_names_df_raw %>% 
  dplyr::rename(species = annotated_name)  %>% 
  filter(species_name != "") |> 
  mutate(species_code = as.numeric(factor(species_name)))

species_code_df <- species_names_df |> 
  dplyr::select(species_name, species_code) |> 
  distinct() |> 
  arrange(species_code)

species_code_df
```



### diversity data

```{r}
faith_pd_table <- read.delim(here::here("../sequencing/OTUs clustered/alpha-diversity.tsv"), 
                             header = TRUE, 
                             row.names = 1)

faith_pd_table <- rownames_to_column(faith_pd_table, var = "species")
faith_pd_table

```



```{r}
# correct species names
faith_pd_table <- faith_pd_table %>%
  left_join(species_names_df |> dplyr::select(species, species_code,species_name),
            by = "species") 
```



## Add load data



## island info

```{r}
df_island <- read_excel(path = here::here("data/18sept24_datalab.xlsx"), sheet = "island_characteristics")

df_island$min_emergence <- as.numeric(df_island$min_emergence)
df_island$max_emergence <- as.numeric(df_island$max_emergence)

# Calculate the mean difference between min_emergence and max_emergence
mean_diff <- df_island %>%
  filter(!is.na(min_emergence) & !is.na(max_emergence)) %>%
  mutate(diff = max_emergence - min_emergence) %>%
  summarise(mean_diff = mean(diff, na.rm = TRUE)) %>%
  pull(mean_diff)

# Estimate max_emergence for Champion by adding the mean difference to its min_emergence
df_island <- df_island %>%
  mutate(max_emergence = ifelse(is.na(max_emergence), min_emergence + mean_diff, max_emergence))

# Rename the column "mean_emergence" to "island_age" in df_island
# add an ID column
df_island <- df_island %>% dplyr::rename(island_area = area) |> 
  mutate(island_code = as.numeric(as.factor(island)))


# increase max. emergence of Sants Fe by a small amount - min. and max for this island are the same and this gives error in the model when trying to estimate the approximate age.

df_island <- df_island |> 
  dplyr::mutate(max_emergence = ifelse(island == "Santa_Fe", max_emergence + 0.1, max_emergence))


df_island 

```


```{r}
df_load_raw <- read_excel(path = here::here("data/18sept24_datalab.xlsx"), sheet = 3)

df_load_raw$nematode_count[which(df_load_raw$nematode_count == ">100")] <- "100"
df_load_raw$nematode_count <- as.numeric(df_load_raw$nematode_count)

df_load_rename <- df_load_raw %>% 
  dplyr::select(id, spp, nematode_count, island) %>% 
  dplyr::rename(species = spp)

# correct species names
df_load <- df_load_rename %>%
  left_join(species_names_df |> dplyr::select(species, species_code,species_name),
            by = "species") |> 
  left_join(df_island |> dplyr::select(island, island_code),
            by = "island")


df_load
```

Calculate mean load

```{r}
# Compute mean load per species and island
mean_load_per_species <- df_load %>%
  group_by(species_name, island) %>%  # Include island in the grouping
  summarize(mean_load = mean(nematode_count, na.rm = TRUE), .groups = "drop")


```

Match species names with tree tips.

```{r}
mean_load_per_species <- mean_load_per_species %>%
  mutate(
    species_name = gsub("_", "-", species_name), # Replace underscores with hyphens
    species_name = gsub("\\.", "", species_name), # Remove periods
    species_name = gsub("-", "", species_name),    # Remove hyphens
     species_name = gsub(" ", "", species_name)  # Remove spaces
  )

faith_pd_table <- faith_pd_table %>%
  mutate(
    species = gsub("_", "-", species), # Replace underscores with hyphens
    species = gsub("\\.", "", species), # Remove periods
    species = gsub("-", "", species),    # Remove hyphens
     species = gsub(" ", "", species)  # Remove spaces
  )
```



```{r}
# Standardize species names and integrate mean load with island information
tree_data <- tree_data %>%
  left_join(mean_load_per_species |> dplyr::select(-island), by = c("label" = "species_name")) %>%
  left_join(faith_pd_table |> dplyr::select(species_name, faith_pd), by = c("label" = "species_name")) |> 
  mutate(
    mean_load = ifelse(is.na(mean_load), 0, mean_load),  # Replace NA loads with 0
    island = factor(island, levels = names(island_colors))  # Standardize island levels
  )
```

```{r}
# Extract tree plot data and join with tree_data
tree_plot_data <- ggtree(tree)$data  # Extract plotting data from ggtree
tree_data <- tree_data %>%
  left_join(tree_plot_data %>% dplyr::select(label, y), by = "label")  # Add y positions to tree_data


tip_data <- left_join(tip_data, tree_data %>% dplyr::select(label, y), by = "label")
```







```{r}
# Define a gradient for mean load
load_colors <- scales::col_numeric(palette = viridis::viridis(100), domain = NULL)


# Generate ggtree plot data
 p <- ggtree(tree)

# # Extract the plotting data (with y values)
 tree_plot_data <- p$data
 
 tree_plot_data <- tree_plot_data %>%
  mutate(
    label = gsub("_", "-", label), # Replace underscores with hyphens
    label = gsub("\\.", "", label), # Remove periods
    label = gsub("-", "", label),    # Remove hyphens
     label = gsub(" ", "", label)  # Remove spaces
  )


# # Merge mean load and island information
#  tree_plot_data <- tree_plot_data %>%
#    left_join(tree_data %>% dplyr::select(label, mean_load, island, faith_pd), by = "label") %>%
#    mutate(mean_load = ifelse(is.na(mean_load), 0, mean_load))  # Replace NA loads with 0
 
  tree_plot_data <- tree_plot_data %>%
  left_join(mean_load_per_species, by = c("label" = "species_name")) %>%
  left_join(faith_pd_table |> dplyr::select(species_name, faith_pd), by = c("label" = "species_name")) |> 
  mutate(
    mean_load = ifelse(is.na(mean_load), 0, mean_load),  # Replace NA loads with 0
    island = factor(island, levels = names(island_colors))  # Standardize island levels
  )
```
Add island information for unasessed species

```{r}
missing_island_species <- read.csv(here::here("data/missing_island_species.csv"))

tree_plot_data1 <- tree_plot_data %>%
  mutate(island = as.character(island))  # This prevents level mismatches

# First, decode islands properly
tree_plot_data1 <- tree_plot_data1 %>%
  mutate(island = as.character(island)) %>%
  left_join(missing_island_species, by = "label") %>%
  mutate(island = ifelse(is.na(island.x), island.y, island.x)) %>%
  select(-island.x, -island.y)

# Restore factor levels to ensure consistency
tree_plot_data1$island <- factor(tree_plot_data1$island, levels = unique(tree_plot_data1$island))

# Check if any species are still missing an island
tree_plot_data1 %>% filter(isTip == TRUE & is.na(island))

tree_plot_data1 <- tree_plot_data1 %>%
  mutate(label = ifelse(label == "littleBlackGarrapatero", "spnovGarrap", label))
```



```{r}
plot_tree_load_fpd <- ggtree(tree) %<+% tree_plot_data1 +
  geom_tree(aes(color = island), size = 0.8) +  # Color branches by island
  
  # Explicitly use tree_plot_data for tip labels
  geom_tiplab(data = tree_plot_data1, aes(label = label), 
              color = "black",  # Force labels to be black
              align = TRUE,  # Align all labels
              offset = 0.01,  # Add space between branches and labels
              size = 2.5, show.legend = FALSE) +  # Label size
  
  geom_tippoint(aes(color = island), size = 3) +  # Add colored tip points
  scale_color_manual(values = island_colors, name = "Island") +  # Custom color scale
  theme_tree2() +
  xlim(NA, max(node.depth.edgelength(tree)) * 1.6) +  # Extend space for mean load bars
  theme(legend.position = "right")+


  
  new_scale_color() +  # Allow a separate color scale for Faith PD
  geom_segment(
    data = tree_plot_data1 %>% filter(faith_pd > 0),  # Only species with non-zero Faith PD
    aes(x = max(node.depth.edgelength(tree)) * 1.41,  # Start of the bar
        xend = max(node.depth.edgelength(tree)) * 1.46,  # End of the bar
        y = y, yend = y, color = faith_pd),  # Map Faith PD to color gradient
    linewidth = 3  # Adjust bar thickness
  ) +
  scale_color_gradientn(
    colors = viridis::turbo(100),  # Use a gradient color palette
    name = "Faith PD"  # Legend title
  ) +
  theme(legend.position = "right")+



  new_scale_color() +  # Allow a separate color scale for Mean Load
  geom_segment(
    data = tree_plot_data1 %>% filter(mean_load > 0),  # Only non-zero values
    aes(x = max(node.depth.edgelength(tree)) * 1.35,  # Start of the bar
        xend = max(node.depth.edgelength(tree)) * 1.40,  # End of the bar
        y = y, yend = y, color = mean_load),  # Map Faith PD to color gradient
    linewidth = 3  # Adjust bar thickness
  ) +
  scale_color_gradientn(
    colors = viridis::turbo(100),  # Use a different gradient for Mean Load
    name = "Mean Load"
  ) +
#  scale_size_continuous(range = c(1, 5), name = "Mean Load") +  # Adjust size range
  theme(legend.position = "right")

#ggsave(here::here("output/figures/phylo/phylo_faith_pd.png"), height = 10, width = 8)
```

```{r}
plot_tree_load_fpd <- ggtree(tree) %<+% tree_plot_data1 +
  geom_tree(aes(color = island), size = 0.8) +  # Color branches by island
  
  # Explicitly use tree_plot_data for tip labels
  geom_tiplab(data = tree_plot_data1, aes(label = label), 
              color = "black",  # Force labels to be black
              align = TRUE,  # Align all labels
              offset = 0.01,  # Add space between branches and labels
              size = 2.5, show.legend = FALSE) +  # Label size
  
  geom_tippoint(aes(color = island), size = 3) +  # Add colored tip points
  
  # Island colors legend (TWO ROWS at the BOTTOM)
  scale_color_manual(
    values = island_colors, 
    name = "Island",
    guide = guide_legend(nrow = 2, order = 3)  # Set to two rows, order third
  ) +
  
  theme_tree2() +
  xlim(NA, max(node.depth.edgelength(tree)) * 1.6) +  # Extend space for mean load bars
  theme(legend.position = "bottom") +  # Move Island legend to the bottom
  

  new_scale_color() +  # Allow a separate color scale for Faith PD
  geom_segment(
    data = tree_plot_data1 %>% filter(faith_pd > 0),  # Only species with non-zero Faith PD
    aes(x = max(node.depth.edgelength(tree)) * 1.41,  
        xend = max(node.depth.edgelength(tree)) * 1.46,  
        y = y, yend = y, color = faith_pd),  
    linewidth = 3  
  ) +
  scale_color_gradientn(
    colors = viridis::turbo(100),  
    name = "Faith PD",
    guide = guide_colorbar(order = 2)  # Keep Faith PD legend on the right
  ) + 
  theme(legend.position = "right") +  # Keep Faith PD legend on the right
  

  new_scale_color() +  # Allow a separate color scale for Mean Load
  geom_segment(
    data = tree_plot_data1 %>% filter(mean_load > 0),  
    aes(x = max(node.depth.edgelength(tree)) * 1.35,  
        xend = max(node.depth.edgelength(tree)) * 1.40,  
        y = y, yend = y, color = mean_load),  
    linewidth = 3  
  ) +
  scale_color_gradientn(
    colors = viridis::turbo(100),  
    name = "Mean Load",
    guide = guide_colorbar(order = 1)  # Keep Mean Load legend on the right
  ) +
  theme(legend.position = "right")  # Keep Mean Load legend on the right

# Print the plot
print(plot_tree_load_fpd)

#ggsave(here::here("output/figures/phylo/delete.png"), height = 10, width = 8)
```



########################################################
########################################################


# Make co-phylo figure


## OTUS tree

```{r}
plot(phylo_tree)
```

```{r}
# saveRDS(phylo_tree_otus, here::here("output/temporal_output/phylo_tree_otus.rds"))
# saveRDS(tree, here::here("output/temporal_output/tree.rds"))
# saveRDS(tree_plot_data, here::here("output/temporal_output/tree_plot_data.rds"))
```


```{r}
library(phytools)
library(ape)

# Ensure trees are in the correct format
phylo_tree_otus <- ape::ladderize(phylo_tree_otus)
# ???? Ensure the tree is ladderized for better visualization
tree <- ape::ladderize(tree)

corrected_tree_labels <- tree_plot_data$label  # Use formatted



# ???? Step 2: Prune the tree to include only tips in `corrected_tree_labels`
tips_to_remove <- setdiff(tree$tip.label, corrected_tree_labels)

# Print removed tips
if (length(tips_to_remove) > 0) {
  cat("Removing tips not present in corrected_tree_labels:\n")
  print(tips_to_remove)
} else {
  cat("No tips were removed. All tree tips match corrected_tree_labels.\n")
}

# Prune the tree
tree <- drop.tip(tree, tips_to_remove)

# ???? Step 3: Ensure tree tip labels match corrected names
if (!all(tree$tip.label %in% corrected_tree_labels)) {
  stop("Some tree tips still do not match corrected_tree_labels! Check for inconsistencies.")
}

# ???? Step 4: Create an association matrix using the filtered tree labels
bipartite_matrix_filtered <- bipartite_matrix_filtered[rownames(bipartite_matrix_filtered) %in% tree$tip.label, , drop = FALSE]

# Check if any snails were removed from bipartite matrix
removed_snails <- setdiff(rownames(bipartite_matrix_filtered), tree$tip.label)

if (length(removed_snails) > 0) {
  cat("Removed snail species from bipartite matrix (not in tree):\n")
  print(removed_snails)
}

# Create the association matrix
association <- which(bipartite_matrix_filtered > 0, arr.ind = TRUE)
association_df <- data.frame(
  snails = rownames(bipartite_matrix_filtered)[association[, 1]],  # Snail species
  otus = colnames(bipartite_matrix_filtered)[association[, 2]]     # Nematode OTUs
)

# ???? Step 5: Verify if all snails in `association_df` exist in `tree$tip.label`
snail_mismatches <- setdiff(association_df$snails, tree$tip.label)
otu_mismatches <- setdiff(association_df$otus, phylo_tree_otus$tip.label)

# Print any mismatches
if (length(snail_mismatches) > 0) {
  cat("Mismatched snails (not found in tree):\n")
  print(snail_mismatches)
}
if (length(otu_mismatches) > 0) {
  cat("Mismatched OTUs (not found in phylo_tree_otus):\n")
  print(otu_mismatches)
}

# Stop execution if mismatches exist
if (length(snail_mismatches) > 0 || length(otu_mismatches) > 0) {
  stop("There are unmatched labels! Fix these before proceeding with cophyloplot.")
}

# ???? Step 6: Generate the Cophylogenetic Plot
par(mar = c(0, 0, 0, 0))  # Remove margins for better visualization
cophyloplot(
  tree, phylo_tree_otus, 
  assoc = association_df, 
  length.line = 0.5,  # Adjusts connection line lengths
  space = 20,  # Space between trees
  gap = 2,  # Adjusts line placement
  rotate = TRUE,  # Rotate one tree if needed
  col = "red",  # Color of connecting lines
  link.type="curved",
  link.lty="solid",fsize=c(0.8,0.8)
)


```






########################################################
########################################################


Prune the tree to keep only assessed species

```{r}
# Extract the species names to keep (those that have an assigned name)
species_to_keep <- tree_data$label[tree_data$has_name]

# Step 1: Filter out any empty names from species_to_keep
species_to_keep <- species_to_keep[species_to_keep != ""]

# Step 2: Prune the tree to keep only these species
pruned_tree <- drop.tip(tree, setdiff(tree$tip.label, species_to_keep))

ggtree(pruned_tree) +
  geom_tiplab(aes(label = label), size = 3) + 
  xlim(0, max(pruned_tree$edge.length) + 3)  # Adjust x-axis limits
```



```{r}
df_load_raw <- read_excel(path = here::here("data/19july24_datalab.xlsx"), sheet = 3)

df_load_raw$nematode_count[which(df_load_raw$nematode_count == ">100")] <- "100"
df_load_raw$nematode_count <- as.numeric(df_load_raw$nematode_count)

df_load_rename <- df_load_raw %>% 
  dplyr::select(id, spp, nematode_count, island) %>% 
  rename(species = spp)

# correct species names
df_load <- df_load_rename %>%
  left_join(species_names_df |> dplyr::select(species, species_name),
            by = "species") 
df_load
```

Compute mean and se load per species

```{r}
df_meanload <- df_load %>%
  group_by(species_name) %>%
  summarise(n = n(),
            mean_load = mean(nematode_count, na.rm = TRUE),
            se_load = sd(nematode_count, na.rm = TRUE)/sqrt(length(nematode_count)), 
            sd_load = sd(nematode_count, na.rm = TRUE)) %>%
  arrange(desc(mean_load))

df_meanload
```


```{r}
meanload<-setNames(df_meanload$mean_load, df_meanload$species_name)
s.error<-setNames(df_meanload$se_load, df_meanload$species_name)
```


discard those species in load that are not present in the tree

```{r}
# Filter data to include only species present in the phylogenetic tree
tree_species <- pruned_tree$tip.label
meanload_filtered <- meanload[names(meanload) %in% tree_species]
s.error_filtered <- s.error[names(s.error) %in% tree_species]

# Verify data alignment
match_indices <- match(tree_species, names(meanload_filtered))
meanload_filtered <- meanload_filtered[match_indices]
s.error_filtered <- s.error_filtered[match_indices]
```


# Blomberg's K

```{r}
K_nemaload <- phylosig(pruned_tree, meanload_filtered, method = 'K', test = TRUE, se = s.error_filtered)
K_nemaload
plot.phylosig(K_nemaload)

```


# Lambda

```{r}

lambda_nemaload <- phylosig(pruned_tree, meanload_filtered, method = 'lambda', test = TRUE, se = s.error_filtered) 
lambda_nemaload <- phylosig(pruned_tree, meanload_filtered, method = 'lambda', test = TRUE) 
lambda_nemaload
plot.phylosig(lambda_nemaload)

```



# Plot tree and load

```{r}

contMap(pruned_tree,
        meanload_filtered,
        res = 400,
        lwd = 4,
        outline = FALSE,
        sig = 0)

```



```{r}
# Step 1: Create a mapping of species to islands
species_island_map <- df_load %>%
  dplyr::select(species_name, island) %>%
  distinct() %>%
  group_by(species_name) %>%
  summarize(island = paste(unique(island), collapse = ", "))  # Handle cases where species occur on multiple islands


tree_data1 <- fortify(pruned_tree)

# Merge the tree data with the island information
tree_data1 <- tree_data1 %>%
  left_join(species_island_map, by = c("label" = "species_name"))  # Merge based on the species names

# Create a column indicating whether each tip label has an assigned name
tree_data1$has_name <- !is.na(tree_data1$island)

# Plot the tree with conditional styling
ggtree(pruned_tree) +
  geom_tiplab(data = tree_data1, aes(label = label, 
                                    
                                    color = island),  # Use island information to color labels
              size = 3, 
              align = TRUE,  # Align labels
              linetype = NA,  # Remove lines connecting labels to tree
              offset = 0.01,  # Adjust the position of the labels to avoid going out of margin
              inherit.aes = FALSE) +
  theme(legend.position = "bottom")+  # Remove legend
  xlim(0, max(tree_data$x) + 3)  # Adjust the x-axis limit to control width   # Adjust the x-axis limit to control width

```


# Analysis with diversity


Keep species names for which nematode amplification was successful

```{r}
diversity_per_snailspp <- read.csv(here::here("output/tables/diversity_per_snailspp.csv"))[,-1]

# correct species names
diversity_per_snailspp <- diversity_per_snailspp %>%
  left_join(species_names_df |> dplyr::select(species, species_name),
            by = "species") 
```


Prune the tree to keep only assessed species

```{r}
# Extract the species names to keep from your dataset
species_assessed_div <- unique(diversity_per_snailspp$species_name)

# Check which species names are not found in the tree's tip labels
mismatched_species <- setdiff(species_assessed_div, pruned_tree$tip.label)

mismatched_species
```

Filter data

```{r}
diversity_per_snailspp_filtered <- diversity_per_snailspp %>%
  filter(!(species_name %in% mismatched_species))
  
```


```{r}
div_nem_species <-setNames(diversity_per_snailspp_filtered$nematode_species, diversity_per_snailspp_filtered$species_name)

if(length(setdiff(unique(pruned_tree$tip.label),diversity_per_snailspp_filtered$species_name)) != 0){
"problem- mismatch in names"
}else{
"all good - species in data and tree match"
}

```
How to get standard error? maybe from sequence data 


# Blomberg's K

```{r}
K_nemadiv <- phylosig(pruned_tree, div_nem_species, method = 'K', test = TRUE)
K_nemadiv
plot.phylosig(K_nemadiv)

```


# Lambda

```{r}

lambda_nemadiv <- phylosig(pruned_tree, div_nem_species, method = 'lambda', test = TRUE) 
lambda_nemadiv <- phylosig(pruned_tree, div_nem_species, method = 'lambda', test = TRUE) 
lambda_nemadiv
plot.phylosig(lambda_nemadiv)

```

# Plot tree and nem diversity


```{r}
matching_species <- intersect(pruned_tree$tip.label, names(div_nem_species))

# Filter and reorder the phenotypic data to match the tree tip labels
div_nem_species_matched <- div_nem_species[matching_species]
div_nem_species_matched <- div_nem_species_matched[match(pruned_tree$tip.label, names(div_nem_species_matched))]

if (!all(names(div_nem_species_matched) == pruned_tree$tip.label)) {
  stop("Error: The names in the phenotypic data do not match the tip labels in the tree.")
}

# Use contMap to plot the tree with the matched data
contMap(pruned_tree,
        div_nem_species_matched,
        res = 400,
        lwd = 4,
        outline = FALSE,
        sig = 0)
```


--------------------------------------------------


```{r}
heatmap_data <- data.frame(
  species_name = names(div_nem_species_matched),
  div = div_nem_species_matched
)

# Aggregate the data to make species names unique
heatmap_data <- heatmap_data %>%
  group_by(species_name) %>%
  summarise(div = sum(div, na.rm = TRUE)) %>%  # Adjust to your preferred method (sum, mean, etc.)
  ungroup()

# Ensure there are no missing values
heatmap_data <- heatmap_data %>%
  filter(!is.na(div))  # Remove rows with NA values

# Set the row names to the unique species names
rownames(heatmap_data) <- heatmap_data$species_name

# Plot the tree
p <- ggtree(pruned_tree) +
  geom_tiplab(align = TRUE, linesize = 0.5, offset = 0.01) +
  xlim(0, 3)  # Adjust the x-axis limits as needed

# Add the heatmap next to the tree
p1 <- gheatmap(p, heatmap_data[, "div", drop = FALSE], offset = 0.8, width = 0.1, 
               colnames = FALSE, low = "white", high = "red", legend_title = "div") +
  scale_fill_gradient(low = "white", high = "red")  # Ensures a color gradient for the fill

# Print the plot
print(p1)
```