---
title: "R Notebook"
---

## load and combine data

```{r}
library(tidyverse)
library(readxl)
library(naniar)
# library(ape)
# library(ggpubr)
# library(ade4)
library(vegan)
library(stringr)
library(brms)
library(cmdstanr)
library(bayesplot)
library(tidybayes)
library(posterior)
```

# load csv with species names for checks

Note that all species without data for load will be asigned a "2" in the code as they wont have any entry in the "species_name" column. This allows to filter out these species and only work with those for which load has been measured.

```{r}
species_names_df <- read.csv(here::here("data/species_names_checks.csv"))
species_names_df <- species_names_df %>% 
  rename(species = annotated_name)  %>%
  mutate(species_code = as.numeric(factor(species_name)))

species_names_df
```

## island info

```{r}
df_island <- read_excel(path = here::here("data/19july24_datalab.xlsx"), sheet = "island_characteristics")

# Rename the column "mean_emergence" to "island_age" in df_island
df_island <- df_island %>% rename(island_age = mean_emergence,
                                  island_area = area)

df_island

```

## Load dataset

```{r}
df_load_raw <- read_excel(path = here::here("data/19july24_datalab.xlsx"), sheet = 3)

# df_load |> glimpse()


df_load_raw$nematode_count[which(df_load_raw$nematode_count == ">100")] <- "100"
df_load_raw$nematode_count <- as.numeric(df_load_raw$nematode_count)

df_load <- df_load_raw %>% 
  select(id, spp, nematode_count, island) %>% 
  rename(species = spp)

# correct species names
df_load <- df_load %>%
  left_join(species_names_df |> select(species, species_code), by = "species")

# add island characteristics

df_load <- df_load |> 
  left_join(df_island |> select(island, island_age, island_area), by = "island")

df_load

```

Create dataframe linking species codes to island data for reference. This will be used to add island data to the rest of the datasets based on species code.

```{r}
df_load_spp_island_reference <- df_load %>%
  select(species_code, island, island_age, island_area) %>%
  distinct(species_code, .keep_all = TRUE)
```



# brightness data

```{r}

df_bright <- read.csv(here::here("data/kraemer2019_brightness_scores.csv"))
df_bright <- df_bright %>% 
  select(location, spp, brightness, microhab) %>% 
  rename(id = location,
         species = spp) %>%
  left_join(species_names_df |> select(species, species_code), by = "species") %>% # correct species names
  filter(species_code != 2) %>% # filter out species for with no load assessed 
  left_join(df_load_spp_island_reference, by = "species_code") # add island data

df_bright

```


# habitat data


```{r}
df_rad <- read_excel(here::here("data/GPS RAD snails.xlsx"))
df_hab <- df_rad %>% 
  select(Sample, Taxon, 
         habitat = Microhabitat, 
         vegetation = `Vegetation Zone`)%>% 
  rename(species = Taxon) %>% 
  glimpse() |> 
  left_join(species_names_df |> select(species, species_code), by = "species") |> # correct species names
  mutate(habitat = str_to_lower(habitat)) %>% 
  mutate(habitat = str_replace(habitat, "semi-arboreal", "arboreal")) |> # Semi-Arboreal is considered arboreal
  filter(habitat %in% c("arboreal", "terrestrial")) |> 
  select(species_code, microhab = habitat) |> 
  bind_rows(df_bright |> select(microhab, species_code), df_hab)|> # add habitat data from brightness dataset 
  count(microhab, species_code) |> 
  pivot_wider(values_from = n, names_from = microhab, values_fill = 0) |> 
  mutate(total_hab = terrestrial + arboreal) |> 
  filter(species_code != 2) %>% # filter out species for with no load assessed
  left_join(df_load_spp_island_reference, by = "species_code") # add island data

df_hab

```


# Vegetation data

```{r}

df_veg <- df_rad %>% 
  select(Sample, Taxon, 
         habitat = Microhabitat, 
         vegetation = `Vegetation Zone`)%>% 
  rename(species = Taxon) %>% 
  glimpse() |> 
  left_join(species_names_df |> select(species, species_code), by = "species") |> # correct species names
  mutate(vegetation = str_to_lower(vegetation)) %>% 
  filter(vegetation %in% c("humid", "arid")) |> ## filter for only legal values -- others are typos
  count(species_code, vegetation) |> 
  pivot_wider(values_from = n, names_from = vegetation, values_fill = 0) |> 
  mutate(total_veg = humid + arid)|> 
  filter(species_code != 2) %>% # filter out species for with no load assessed
  left_join(df_load_spp_island_reference, by = "species_code") # add island data

df_veg

```



Datasets, with same species_codes:

df_island
df_load
df_bright
df_hab
df_veg 



Only consider species with data for all the measurements:

```{r}

# Get the common species names present in all datasets
common_species <- intersect(df_bright$species_code, df_load$species_code) |> 
  intersect(df_hab$species_code) |> 
  intersect(df_veg$species_code) |> 
  setdiff("")

# Filter each dataframe to include only the common species
df_bright <- df_bright |> 
  filter(species_code %in% common_species)

df_load <- df_load |> 
  filter(species_code %in% common_species)

df_hab <- df_hab |> 
  filter(species_code %in% common_species)

df_veg <- df_veg |> 
  filter(species_code %in% common_species)

```


## create a species-codes mapping table

```{r}
# Combine all species codes from the datasets
all_species_codes <- unique(c(
  df_bright$species_code,
  df_load$species_code,
  df_hab$species_code,
  df_veg$species_code
))

# Ensure there are no more than 21 unique species codes
if(length(all_species_codes) > 21) {
  stop("There are more than 21 unique species codes. Adjust your mapping strategy.")
}

# Create a mapping table
species_code_mapping <- data.frame(
  species_code = all_species_codes,
  species_code_new = 1:length(all_species_codes)
)

species_code_mapping

```


Add new column to each dataset

```{r}
add_new_species_code <- function(df, mapping) {
  df <- merge(df, mapping, by = "species_code", all.x = TRUE)
  df
}

df_bright <- add_new_species_code(df_bright, species_code_mapping)
df_load <- add_new_species_code(df_load, species_code_mapping)
df_hab <- add_new_species_code(df_hab, species_code_mapping)
df_veg <- add_new_species_code(df_veg, species_code_mapping)
```




# Full model


Data

```{r}
# Prepare the data list for Stan
data_list <- list(
  N = nrow(df_bright_filtered),
  N_spp = length(common_species),
  N_islands = length(df_island),
  N_habitat = nrow(df_hab_filtered),
  N_veg = nrow(df_veg_filtered),
  N_nem = nrow(df_load_filtered),
  
  spp_id_bright = df_bright_filtered$species_code_new,
  island_id = df_bright_filtered$island,
  spp_id_hab = df_hab_filtered$species_code_new,
  spp_id_veg = df_veg_filtered$species_code_new,
  spp_id_load = df_load_filtered$species_code_new,
  
  brightness = df_bright_filtered$brightness,
  island_age = df_island$island_age,
  island_area = df_island$island_area,
  
  habitat_arboreal = df_hab_filtered$arboreal,
  total_hab = df_hab_filtered$total_hab,
  
  vegetation_arid = df_veg_filtered$arid,
  total_veg = df_veg_filtered$total_veg,
  
  load = df_load_filtered$nematode_count
)

```


Run model

```{r}


 # Fit the model using cmdstanr
 stan_model <- cmdstan_model(
   stan_file = here::here("stan/full_model_commonspp.stan"), 
   pedantic = TRUE
 )
 
 fit <- stan_model$sample(
   data = data_list,
   seed = 123,             
   chains = 4,               
   parallel_chains = 4,    
   iter_warmup = 1000,       
   iter_sampling = 2000   
 )
```


