---
title: "R Notebook"
---

## load and combine data

```{r}
library(tidyverse)
library(readxl)
library(naniar)
# library(ape)
# library(ggpubr)
# library(ade4)
library(vegan)
library(stringr)
library(brms)
library(cmdstanr)
library(bayesplot)
library(tidybayes)
library(posterior)
```

# load csv with species names for checks

Note that all species without data for load will be asigned a "2" in the code as they wont have any entry in the "species_name" column. This allows to filter out these species and only work with those for which load has been measured.

```{r}
species_names_df <- read.csv(here::here("data/species_names_checks.csv"))
species_names_df <- species_names_df %>% 
  rename(species = annotated_name)  %>%
  mutate(species_code = as.numeric(factor(species_name)))

species_names_df
```

## island info

```{r}
df_island <- read_excel(path = here::here("data/19july24_datalab.xlsx"), sheet = "island_characteristics")

# Rename the column "mean_emergence" to "island_age" in df_island
df_island <- df_island %>% rename(island_age = mean_emergence,
                                  island_area = area)

df_island

```

## Load dataset

```{r}
df_load_raw <- read_excel(path = here::here("data/19july24_datalab.xlsx"), sheet = 3)

# df_load |> glimpse()


df_load_raw$nematode_count[which(df_load_raw$nematode_count == ">100")] <- "100"
df_load_raw$nematode_count <- as.numeric(df_load_raw$nematode_count)

df_load <- df_load_raw %>% 
  select(id, spp, nematode_count, island) %>% 
  rename(species = spp)

# correct species names
df_load <- df_load %>%
  left_join(species_names_df |> select(species, species_code), by = "species")

# add island characteristics

df_load <- df_load |> 
  left_join(df_island |> select(island, island_age, island_area), by = "island")

df_load

```

Create dataframe linking species codes to island data for reference. This will be used to add island data to the rest of the datasets based on species code.

```{r}
df_load_spp_island_reference <- df_load %>%
  select(species_code, island, island_age, island_area) %>%
  distinct(species_code, .keep_all = TRUE)
```



# brightness data

```{r}

df_bright <- read.csv(here::here("data/kraemer2019_brightness_scores.csv"))
df_bright <- df_bright %>% 
  select(location, spp, brightness, microhab) %>% 
  rename(id = location,
         species = spp) %>%
  left_join(species_names_df |> select(species, species_code), by = "species") %>% # correct species names
  filter(species_code != 2) %>% # filter out species for with no load assessed 
  left_join(df_load_spp_island_reference, by = "species_code") # add island data

df_bright

```


# habitat data


```{r}
df_rad <- read_excel(here::here("data/GPS RAD snails.xlsx"))
df_hab <- df_rad %>% 
  select(Sample, Taxon, 
         habitat = Microhabitat, 
         vegetation = `Vegetation Zone`)%>% 
  rename(species = Taxon) %>% 
  glimpse() |> 
  left_join(species_names_df |> select(species, species_code), by = "species") |> # correct species names
  mutate(habitat = str_to_lower(habitat)) %>% 
  mutate(habitat = str_replace(habitat, "semi-arboreal", "arboreal")) |> # Semi-Arboreal is considered arboreal
  filter(habitat %in% c("arboreal", "terrestrial")) |> 
  select(species_code, microhab = habitat)


df_hab <-  bind_rows(df_bright |> select(microhab, species_code), df_hab)|> # add habitat data from brightness dataset 
  count(microhab, species_code) |> 
  pivot_wider(values_from = n, names_from = microhab, values_fill = 0) |> 
  mutate(total_hab = terrestrial + arboreal) |> 
  filter(species_code != 2) %>% # filter out species for with no load assessed
  left_join(df_load_spp_island_reference, by = "species_code") # add island data

df_hab

```


# Vegetation data

```{r}

df_veg <- df_rad %>% 
  select(Sample, Taxon, 
         habitat = Microhabitat, 
         vegetation = `Vegetation Zone`)%>% 
  rename(species = Taxon) %>% 
  glimpse() |> 
  left_join(species_names_df |> select(species, species_code), by = "species") |> # correct species names
  mutate(vegetation = str_to_lower(vegetation)) %>% 
  filter(vegetation %in% c("humid", "arid")) |> ## filter for only legal values -- others are typos
  count(species_code, vegetation) |> 
  pivot_wider(values_from = n, names_from = vegetation, values_fill = 0) |> 
  mutate(total_veg = humid + arid)|> 
  filter(species_code != 2) %>% # filter out species for with no load assessed
  left_join(df_load_spp_island_reference, by = "species_code") # add island data

df_veg

```



Datasets, with same species_codes:

df_island
df_load
df_bright
df_hab
df_veg 



Only consider species with data for all the measurements:

```{r}

# Get the common species names present in all datasets
common_species <- intersect(df_bright$species_code, df_load$species_code) |> 
  intersect(df_hab$species_code) |> 
  intersect(df_veg$species_code) |> 
  setdiff("")

# Filter each dataframe to include only the common species
df_bright <- df_bright |> 
  filter(species_code %in% common_species)

df_load <- df_load |> 
  filter(species_code %in% common_species)

df_hab <- df_hab |> 
  filter(species_code %in% common_species)

df_veg <- df_veg |> 
  filter(species_code %in% common_species)

```


## create a species-codes mapping table

```{r}
# Combine all species codes from the datasets
all_species_codes <- unique(c(
  df_bright$species_code,
  df_load$species_code,
  df_hab$species_code,
  df_veg$species_code
))

# Ensure there are no more than 21 unique species codes
if(length(all_species_codes) > 21) {
  stop("There are more than 21 unique species codes. Adjust your mapping strategy.")
}

# Create a mapping table
species_code_mapping <- data.frame(
  species_code = all_species_codes,
  species_code_new = 1:length(all_species_codes)
)

species_code_mapping

```


Add new column to each dataset

```{r}
add_new_species_code <- function(df, mapping) {
  df <- merge(df, mapping, by = "species_code", all.x = TRUE)
  df
}

df_bright <- add_new_species_code(df_bright, species_code_mapping)
df_load <- add_new_species_code(df_load, species_code_mapping)
df_hab <- add_new_species_code(df_hab, species_code_mapping)
df_veg <- add_new_species_code(df_veg, species_code_mapping)
```


Create island mapping and assign island codes

```{r}
# Extract unique islands from one of the datasets that contain all island names
unique_islands <- unique(c(df_load$island, df_bright$island, df_hab$island, df_veg$island))

# Create a dataframe that maps island names to unique integer codes
island_codes <- data.frame(
  island = unique_islands,
  island_code = seq_along(unique_islands)
)

island_codes
```


```{r}
# Add island_code to df_load
df_load <- df_load %>%
  left_join(island_codes, by = "island")

# Add island_code to df_bright
df_bright <- df_bright %>%
  left_join(island_codes, by = "island")

# Add island_code to df_hab
df_hab <- df_hab %>%
  left_join(island_codes, by = "island")

# Add island_code to df_veg
df_veg <- df_veg %>%
  left_join(island_codes, by = "island")
```



# Full model


Data

```{r}
# Prepare the data list for Stan
data_list <- list(
  N_load = nrow(df_load),
  N_bright = nrow(df_bright),
  N_spp = length(common_species),
  N_islands = length(island_codes$island_code),
  N_habitat = nrow(df_hab),
  N_veg = nrow(df_veg),
  
  spp_id_bright = df_bright$species_code_new,
  island_id_load = df_load$island_code,
  island_id_bright = df_bright$island_code,
  spp_id_hab = df_hab$species_code_new,
  spp_id_veg = df_veg$species_code_new,
  spp_id_load = df_load$species_code_new,
  
  brightness = df_bright$brightness,
  island_age_load = df_load$island_age,
  island_age_bright = df_bright$island_age,
  island_area_load = df_load$island_area,
  island_area_bright = df_bright$island_area,
  
  habitat_arboreal = df_hab$arboreal,
  total_hab = df_hab$total_hab,
  
  vegetation_arid = df_veg$arid,
  total_veg = df_veg$total_veg,
  
  load = df_load$nematode_count
)

```


Run model

```{r}
# Fit the model using cmdstanr
full_model_commonspp_slopepos_load <- cmdstan_model(
  stan_file = here::here("stan/full_model_commonspp_slopepos_load.stan"), 
  pedantic = TRUE
)
```


```{r, eval=FALSE}
full_model_commonspp_slopepos_load_samples <- 
  full_model_commonspp_slopepos_load$sample(
    data = data_list,
    seed = 123,             
    chains = 5,               
    parallel_chains = 5,    
    iter_warmup = 200,       
    iter_sampling = 200,
    output_dir = here::here("model_object/full_model_commonspp_slopepos_load")
  )

```


```{r}
# read in the model from above
full_model_commonspp_slopepos_load_samples <- as_cmdstan_fit(dir("model_object/full_model_commonspp_slopepos", full.names = TRUE))

full_model_commonspp_slopepos_load_samples$summary()

```


```{r}
# Extract the posterior samples
posterior_samples <- full_model_commonspp_slopepos_load_samples$draws()


posterior_df <- as_draws_df(posterior_samples)

full_model_commonspp_slopepos_load_samples$diagnostic_summary()

full_model_commonspp_slopepos_load_samples$summary()

mcmc_areas(
  posterior_df,
  pars = vars(starts_with("slope")),
  prob = 0.8  
)
```


# # generated quantities - linear regressions


## aridity


```{r}
sp_sample_arid <- df_veg |> 
  select(species_code, arid, total_veg) |> 
  mutate(p_arid = arid/total_veg)


sp_sample_load <- df_load |> 
  group_by(species_code) |> 
  summarize(mean_load = mean(nematode_count),
            n_sample = n())

arid_load_df <- sp_sample_load |> 
  left_join(sp_sample_arid, by = join_by(species_code)) 

arid_load_df |> 
  ggplot(aes(x = p_arid, y = mean_load, size = n_sample, colour = total_veg)) + 
  geom_point(alpha = .7)
```

```{r}
# calculate averages
# log brightness

df_bright |> 
  group_by(species_code) |> 
  summarize(mean_log_bright = mean(log(brightness))) |> 
  pluck("mean_log_bright") |> 
  mean()

df_hab |> 
  mutate(p_arbor = arboreal/total_hab) |> 
  pluck("p_arbor") |> mean()

df_island |> 
  pluck("island_age") |> mean()

df_island$island_area |> log() |>  mean()

```

compile GQ program

```{r}
gq_trendline_arid <- cmdstan_model(here::here("stan/gq_trendline_full_model_commonspp_slopepos_load.stan"))

arid_line_samples <- gq_trendline_arid$generate_quantities(
  fitted_params = full_model_commonspp_slopepos_load_samples,
  data =data_list)
```


```{r}
arid_line_samples |> 
  gather_rvars(lambda[i]) |> 
  mutate(p_arid = c(0, .2, .5, .7, 1), 
         mean_load = exp(.value)) |> 
  ggplot(aes(x = p_arid, ydist = mean_load)) + 
  stat_lineribbon() + 
  scale_fill_brewer(type = "seq", palette = 2, direction = -1) + 
  geom_point(aes(y = mean_load), data = arid_load_df)
```




# generated quantities

```{r}

# Fit the model using cmdstanr
gq_slopepos_load <- cmdstan_model(
  stan_file = here::here("stan/gq_full_model_commonspp_slopepos_load.stan"), 
  pedantic = TRUE
)


gq <- gq_slopepos_load$generate_quantities(
  fitted_params = full_model_commonspp_slopepos_load_samples,
  data = data_list
)
```
