---
title: "R Notebook"
output: html_notebook
---

```{r setup-packages}
#| message: false
#| warning: false
library(tidyverse)
library(readxl)
library(naniar)
library(cmdstanr)
library(bayesplot)
library(tidybayes)
library(posterior)
library(ggpubr)
library(phytools)
library(ape)

```

parameters for plotting

```{r}
my_theme<-theme(axis.text=element_text(size=12),
        axis.title = element_text(size = 14),
        legend.text=element_text(size=10),
        legend.title = element_text(size=12),
        plot.title = element_text(face="bold",size=14,margin=margin(0,0,20,0),hjust = 0.5),
        axis.title.y = element_text(hjust = 0.5),
        axis.title.x = element_text(hjust = 0.5))
```


# diversity nematodes df

```{r}
diversity_per_snailspp <- read.csv(here::here("output/tables/diversity_per_snailspp.csv"))[,-1]
```


## compute species code id

```{r}
species_names_df_raw <- read.csv(here::here("data/species_names_checks.csv"))

# we're dropping the species with no name in this database. 
# this indicates morphospecies which have no formal name
species_names_df <- species_names_df_raw %>% 
  rename(species = annotated_name)  %>% 
  filter(species_name != "") 
```


```{r}
diversity_per_snailspp <- read.csv(here::here("output/tables/diversity_per_snailspp.csv"))[,-1]

# correct species names
diversity_per_snailspp <- diversity_per_snailspp %>%
  left_join(species_names_df |> select(species, species_name),
            by = "species")
```


Filter species names based on the ones assessed for diversity

```{r}
species_names_filtered <- species_names_df_raw %>%
  filter(species_name %in% diversity_per_snailspp$species_name)
```

Create species code

```{r}
species_names_df <- species_names_filtered |> 
   mutate(species_code = as.numeric(factor(species_name)))

 species_code_df <- species_names_df |> 
   select(species_name, species_code) |> 
   distinct() |> 
   arrange(species_code)
 
 species_code_df
```


## island info

```{r}
df_island <- read_excel(path = here::here("data/19july24_datalab.xlsx"), sheet = "island_characteristics")

# Rename the column "mean_emergence" to "island_age" in df_island
# add an ID column
df_island <- df_island %>% rename(island_age = mean_emergence,
                                  island_area = area) |> 
  mutate(island_code = as.numeric(as.factor(island)))

df_island

```



I need to add the island information. Perhaps the best is to do the mapping before, in the diversity_per_snailspp table, so that I can use the island df after with it.

```{r}
metadata <- read_excel(here::here("../sequencing/metadata.xlsx"))[-1,]

metadata <- metadata |> 
  mutate(annotated_name = species)|> 
  left_join(species_names_df |> select(annotated_name, species_name), by = "annotated_name")
```



# Nematode diversity dataset

```{r}
diversity_per_snailspp <- read.csv(here::here("output/tables/diversity_per_snailspp.csv"))[,-1]

# correct species names
diversity_per_snailspp <- diversity_per_snailspp %>%
  mutate(annotated_name = species) |> 
  left_join(species_names_df |> select(annotated_name, species_name, species_code),
            by = "annotated_name") |> 
  left_join(metadata |> select(species_name, island), by = "species_name") |> 
  left_join(df_island |> select(island, island_code),
            by = "island")
diversity_per_snailspp
```



Previously we merged all the data on islands into the load dataset, but in the last bit of code above we depart from this.
Here we are adding only the island code. 
Island level data will instead be added to the species data, since it is used in calculations at the species level:

HOWEVER, species-island associations are only found in the diversity dataset

```{r}
sp_code_island_df <- diversity_per_snailspp |> 
  select(species, species_code, island, island_code) |> 
  distinct() |> 
  left_join(df_island) |> 
  left_join(species_names_df |> select(species_name, species_code)) |> 
  distinct()

sp_code_island_df
sp_code_island_df |> 
  head() |> 
  knitr::kable()
```


## Brightness data

```{r}

df_bright_raw <- read_csv(here::here("data/kraemer2019_brightness_scores.csv"))
df_bright <- df_bright_raw %>% 
  select(location, spp, brightness, microhab) %>% 
  rename(id = location,
         species = spp) %>%
  rename(annotated_name = species)|> 
  left_join(species_names_df |> select(annotated_name, species_code), by = "annotated_name")

df_bright


```

```{r}
df_bright_noNA <- df_bright |> 
  drop_na(species_code)
```


```{r}
df_rad_raw <- read_excel(here::here("data/GPS RAD snails.xlsx"))
df_rad_rename <- df_rad_raw %>% 
  select(Sample, Taxon, 
         habitat = Microhabitat, 
         vegetation = `Vegetation Zone`)%>% 
  rename(annotated_name = Taxon) %>% 
  glimpse() |> 
  left_join(species_names_df |> select(annotated_name, species_code), by = "annotated_name") |> # correct species names
  mutate(habitat = str_to_lower(habitat)) %>% 
  mutate(habitat = str_replace(habitat, "semi-arboreal", "arboreal")) |> # Semi-Arboreal is considered arboreal
  filter(habitat %in% c("arboreal", "terrestrial"))

df_rad_rename
```
```{r}
df_rad_rename_noNA <- df_rad_rename |> 
  drop_na(species_code)
```


```{r}
df_hab <-  bind_rows(df_bright_noNA |> 
                       select(microhab, species_code),
                     df_rad_rename_noNA |> 
                       select(microhab = habitat, species_code)
) |> # add habitat data from brightness dataset 
  count(microhab, species_code) |> 
  pivot_wider(values_from = n, names_from = microhab, values_fill = 0) |> 
  mutate(total_hab = terrestrial + arboreal)

df_hab
```

### Vegetation data

```{r}

df_veg <- df_rad_rename_noNA %>% 
  select(Sample, annotated_name, 
         habitat, vegetation)%>%
  # glimpse() |> 
  left_join(species_names_df |> 
              select(annotated_name, species_code), by = "annotated_name") |> # correct species names
  # glimpse() |> 
  mutate(vegetation = str_to_lower(vegetation)) |> 
  # filter(vegetation %in% c("humid", "arid")) |> ## filter for only legal values -- others are typos
  count(species_code, vegetation) |> 
  pivot_wider(values_from = n, names_from = vegetation, values_fill = 0) |> 
  mutate(total_veg = humid + arid)

df_veg

```


## Negative binomial model with informed priors on brightness and load


```{r}
#| class-output: stan
all_spp_load_constr_bright <- cmdstan_model(here::here("stan/all_spp_load_nbinom_constr_bright.stan"),
                                          pedantic = TRUE)
```


load here is the diversity. Using the same name of dataset df_load to ease the amount of names in the model

```{r}
df_load <- diversity_per_snailspp
```


```{r}

# Data code including phylogenetic covariance matrix
constr_brightload_samp <- all_spp_load_constr_bright$sample(
  data = list(
  N_spp = nrow(sp_code_island_df),  # Number of species (47)
  N_island = max(sp_code_island_df$island_code),
  sp_id = sp_code_island_df$species_code,
  island_area = sp_code_island_df$island_area,
  island_age = sp_code_island_df$island_age,
  island_index_spp = sp_code_island_df$island_code,
  
  # Habitat data
  N_habitat = nrow(df_hab),
  habitat_arboreal = df_hab$arboreal,
  total_hab = df_hab$total_hab,
  sp_index_hab = df_hab$species_code,
  
  # Vegetation zone data
  N_veg = nrow(df_veg),
  veg_arid = df_veg$arid,
  total_veg = df_veg$total_veg,
  sp_index_veg = df_veg$species_code,
  
  # Brightness data
  N_bright = nrow(df_bright_noNA), 
  brightness_obs = df_bright_noNA$brightness,
  sp_index_bright = df_bright_noNA$species_code,
  
  # Nematode load data
  N_load = nrow(df_load),  # Total observations for nematode load
  nematode_count = df_load$nematode_species,
  sp_index_load = df_load$species_code,  # Use species_code from df_load
  island_index_load = df_load$island_code  # Use island_code from df_load
  
),
  chains = 10, parallel_chains = 10, refresh = 0,
iter_sampling = 500,
output_dir = here::here("output/models/missing_data_load_nbinom_constr_brightdiv")
)



```

```{r}
# read in the model from above
constr_brightload_samp <- as_cmdstan_fit(dir(here::here("output/models/missing_data_load_nbinom_constr_brightdiv"), full.names = TRUE))
```


```{r}
# Extract the posterior samples
posterior_samples <- constr_brightload_samp$draws()


posterior_df <- as_draws_df(posterior_samples)

#all_spp_load_samp$diagnostic_summary()
#all_spp_load_samp$summary()

mcmc_areas(
  posterior_df,
  pars = vars(starts_with("slope")),
  prob = 0.8  
)
```