---
title: "Model load"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---

```{r setup-packages}
#| message: false
#| warning: false
library(tidyverse)
library(readxl)
library(naniar)
library(cmdstanr)
library(bayesplot)
library(tidybayes)
library(posterior)
library(ggpubr)
library(phytools)
library(ape)

```

parameters for plotting

```{r}
my_theme<-theme(axis.text=element_text(size=12),
        axis.title = element_text(size = 14),
        legend.text=element_text(size=10),
        legend.title = element_text(size=12),
        plot.title = element_text(face="bold",size=14,margin=margin(0,0,20,0),hjust = 0.5),
        axis.title.y = element_text(hjust = 0.5),
        axis.title.x = element_text(hjust = 0.5))
```



## compute species code id

```{r}
species_names_df_raw <- read.csv(here::here("data/species_names_checks.csv"))

# we're dropping the species with no name in this database. 
# this indicates morphospecies which have no formal name
species_names_df <- species_names_df_raw %>% 
  rename(species = annotated_name)  %>% 
  filter(species_name != "") |> 
  mutate(species_code = as.numeric(factor(species_name)))

species_code_df <- species_names_df |> 
  select(species_name, species_code) |> 
  distinct() |> 
  arrange(species_code)

species_code_df
```



## island info

```{r}
df_island <- read_excel(path = here::here("data/19july24_datalab.xlsx"), sheet = "island_characteristics")

# Rename the column "mean_emergence" to "island_age" in df_island
# add an ID column
df_island <- df_island %>% rename(island_age = mean_emergence,
                                  island_area = area) |> 
  mutate(island_code = as.numeric(as.factor(island)))

df_island

```

## Parasite Load dataset

```{r}
df_load_raw <- read_excel(path = here::here("data/19july24_datalab.xlsx"), sheet = 3)

df_load_raw$nematode_count[which(df_load_raw$nematode_count == ">100")] <- "100"
df_load_raw$nematode_count <- as.numeric(df_load_raw$nematode_count)

df_load_rename <- df_load_raw %>% 
  select(id, spp, nematode_count, island) %>% 
  rename(species = spp)

# correct species names
df_load <- df_load_rename %>%
  left_join(species_names_df |> select(species, species_code),
            by = "species") |> 
  left_join(df_island |> select(island, island_code),
            by = "island")
df_load

```

## Island + species information

Previously we merged all the data on islands into the load dataset, but in the last bit of code above we depart from this.
Here we are adding only the island code. 
Island level data will instead be added to the species data, since it is used in calculations at the species level:

HOWEVER, species-island associations are only found in the load dataset

```{r}
sp_code_island_df <- df_load |> 
  select(species, species_code, island, island_code) |> 
  distinct() |> 
  left_join(df_island) |> 
  left_join(species_names_df |> select(species_name, species_code)) |> 
  distinct()

sp_code_island_df |> 
  head() |> 
  knitr::kable()
```

## Brightness data

```{r}

df_bright_raw <- read_csv(here::here("data/kraemer2019_brightness_scores.csv"))
df_bright <- df_bright_raw %>% 
  select(location, spp, brightness, microhab) %>% 
  rename(id = location,
         species = spp) %>%
  left_join(species_names_df |> select(species, species_code), by = "species")

df_bright


```

In this dataset it seems that the MAJORITY of observations don't have a species name. 
Is that correct??

```{r}
df_bright |> count(species_code) |> 
  arrange(desc(n))
```

That's a lot of data to lose! 

```{r}
df_bright_noNA <- df_bright |> 
  drop_na(species_code)
```



```{r}
df_rad_raw <- read_excel(here::here("data/GPS RAD snails.xlsx"))
df_rad_rename <- df_rad_raw %>% 
  select(Sample, Taxon, 
         habitat = Microhabitat, 
         vegetation = `Vegetation Zone`)%>% 
  rename(species = Taxon) %>% 
  glimpse() |> 
  left_join(species_names_df |> select(species, species_code), by = "species") |> # correct species names
  mutate(habitat = str_to_lower(habitat)) %>% 
  mutate(habitat = str_replace(habitat, "semi-arboreal", "arboreal")) |> # Semi-Arboreal is considered arboreal
  filter(habitat %in% c("arboreal", "terrestrial"))
```


#### warning! 

There should be no species in this list that are not found in the master species list! 
However we are missing the following in the species list

```{r}
df_rad_rename |> 
  filter(is.na(species_code))
```


for now we are dropping them

```{r}
df_rad_rename_noNA <- df_rad_rename |> 
  drop_na(species_code)
```



```{r}
df_hab <-  bind_rows(df_bright_noNA |> 
                       select(microhab, species_code),
                     df_rad_rename_noNA |> 
                       select(microhab = habitat, species_code)
) |> # add habitat data from brightness dataset 
  count(microhab, species_code) |> 
  pivot_wider(values_from = n, names_from = microhab, values_fill = 0) |> 
  mutate(total_hab = terrestrial + arboreal)

df_hab
```

### Vegetation data

the Rad dataset is where the data on vegetation come from 

Previously we filtered for only the vegetation types spelled "humid" or "arid", but now (19 August 2024)
there seems to be no different spellings of these in the dataset. 
Were they corrected in the original spreadsheet?


```{r}

df_veg <- df_rad_rename_noNA %>% 
  select(Sample, species, 
         habitat, vegetation)%>% 
  # glimpse() |> 
  left_join(species_names_df |> 
              select(species, species_code), by = "species") |> # correct species names
  # glimpse() |> 
  mutate(vegetation = str_to_lower(vegetation)) |> 
  # filter(vegetation %in% c("humid", "arid")) |> ## filter for only legal values -- others are typos
  count(species_code, vegetation) |> 
  pivot_wider(values_from = n, names_from = vegetation, values_fill = 0) |> 
  mutate(total_veg = humid + arid)

df_veg

```

## Summary of datasets

We now have all the datasets that we need to build the model. 

#### Vegetation data

```{r}
df_veg |> head() |> knitr::kable()
```

This dataset contains `r sum(df_veg$total_veg)` observations of `r nrow(df_veg)` species.

#### Habitat data

```{r}
df_hab |> head() |> knitr::kable()
```

This dataset contains `r sum(df_hab$total_hab)` observations of `r nrow(df_hab)` species.

#### species and island data

This is the only dataset to contain one row per species with no missing species. It gives us the 

```{r}
sp_code_island_df |> 
  head() |> 
  knitr::kable()
```

#### brightness data

```{r}
df_bright_noNA |> head() |> knitr::kable()
```


#### load data


```{r}
# Ensure df_load has the species_code and island_code columns
df_load <- df_load %>%
  left_join(sp_code_island_df %>% select(species_code, island_code), 
            by = "species_code")

df_load |> head() |> knitr::kable()
```



# Models

Building up slowly from a simple model

## Model of microhabitat

```{r}
#| class-output: stan
all_spp_arbor <- cmdstan_model(here::here("stan/all_spp_arbor.stan"))

all_spp_arbor
```


```{r}
all_spp_arbor_samp <- all_spp_arbor$sample(
  data = list(
    N_spp = nrow(sp_code_island_df),
    sp_id = sp_code_island_df$species_code,
    N_habitat = nrow(df_hab),
    habitat_arboreal = df_hab$arboreal,
    total_hab = df_hab$total_hab,
    sp_index_hab = df_hab$species_code
  ),
chains = 4, parallel_chains = 4, refresh = 0
)
```

## Model of microhabitat and veg zone


```{r}
#| class-output: stan
all_spp_arbor_veg <- cmdstan_model(here::here("stan/all_spp_arbor_veg.stan"))

all_spp_arbor_veg
```


```{r}
all_spp_arbor_veg_samp <- all_spp_arbor_veg$sample(
  data = list(
    N_spp = nrow(sp_code_island_df),
    sp_id = sp_code_island_df$species_code,
    # microhabitat
    N_habitat = nrow(df_hab),
    habitat_arboreal = df_hab$arboreal,
    total_hab = df_hab$total_hab,
    sp_index_hab = df_hab$species_code,
    # veg zone
    N_veg = nrow(df_veg),
    veg_arid = df_veg$arid,
    total_veg = df_veg$total_veg,
    sp_index_veg = df_veg$species_code
  ),
  chains = 4, parallel_chains = 4, refresh = 0
)
```

### Quick parameter visualization

The second model uses both arid and arboreal data. I want to extract those species-specific probabilities and examine them.

```{r}
#| fig-cap: po
all_spp_arbor_veg_samp |> 
  tidybayes::gather_rvars(arid_prob[sp], arboreal_prob[sp]) |> 
  ggplot(aes(x = sp, ydist = .value)) + 
  stat_pointinterval() + 
  facet_wrap(~.variable)
```

```{r}
all_spp_arbor_veg_samp |> 
  tidybayes::spread_rvars(arid_prob[sp], arboreal_prob[sp]) |> 
  mutate_all(.funs = median) |> 
  ggplot(aes(x = arid_prob, y = arboreal_prob)) + geom_point()
```

A few things to note:

* species which are undersampled get probabilities equal to the prior. This is because the model has no phylogenetic structure and no correlation structure either
* species which lack data for one of these two variables usually lack it for the other as well
* most species are 100% or 0% on these two variables, with relatively little in between.

## Model of Brightness, microhab, veg zone

```{r}
#| class-output: stan
all_spp_arbor_veg_bright <- cmdstan_model(here::here("stan/all_spp_arbor_veg_bright.stan"),
                                          pedantic = TRUE)

all_spp_arbor_veg_bright
```


```{r}
all_spp_arbor_veg_bright_samp <- all_spp_arbor_veg_bright$sample(
  data = list(
    N_spp = nrow(sp_code_island_df),
    N_island = max(sp_code_island_df$island_code),
    sp_id = sp_code_island_df$species_code,
    island_area = sp_code_island_df$island_area,
    island_age = sp_code_island_df$island_age,
    island_index_spp = sp_code_island_df$island_code,
    # microhabitat
    N_habitat = nrow(df_hab),
    habitat_arboreal = df_hab$arboreal,
    total_hab = df_hab$total_hab,
    sp_index_hab = df_hab$species_code,
    # veg zone
    N_veg = nrow(df_veg),
    veg_arid = df_veg$arid,
    total_veg = df_veg$total_veg,
    sp_index_veg = df_veg$species_code,
    # brightness data
    N_bright = nrow(df_bright_noNA), 
    brightness_obs = df_bright_noNA$brightness,
    sp_index_bright = df_bright_noNA$species_code
  ),
  chains = 4, parallel_chains = 4, refresh = 0
)
```

let's look now at species brightness predictions from this model

```{r}
bright_rvars <- all_spp_arbor_veg_bright_samp |> 
  tidybayes::gather_rvars(true_ln_bright[species_code]) |> 
  mutate(brightness = exp(.value),
         species_code = forcats::fct_reorder(as.character(species_code), .value))

##################
# Create a named vector for species names
species_code_to_name <- setNames(sp_code_island_df$species_name, sp_code_island_df$species_code)

# Replace species codes with species names in bright_rvars
bright_rvars <- bright_rvars %>%
  mutate(species_name = species_code_to_name[as.character(species_code)])

# Replace species codes with species names in df_bright_noNA
df_bright_noNA <- df_bright_noNA %>%
  mutate(species_name = species_code_to_name[as.character(species_code)])
#################

post_bright <- bright_rvars |> 
  ggplot(aes(x = species_name, ydist = brightness)) + 
  stat_pointinterval() + 
  scale_y_log10()

post_bright <- post_bright + 
  geom_point(aes(x = species_name, y = brightness), inherit.aes = FALSE, 
             data = df_bright_noNA, 
             colour = "red",
             size = .5, alpha = .5,
             position = position_jitter(width = .3)) +
  theme_minimal() +
  my_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  xlab(NULL)




```



# Adding phylogenetic information

```{r}
pruned.tree <- readRDS(here::here("output/tree/tree_spp_load.rds"))
plot(pruned.tree)
```


### Create a phylogenetic covariance matrix

- Phylogenetic Tree and Brownian Motion Assumption: The vcv function assumes that traits evolve under a Brownian motion model along the branches of the phylogenetic tree. In this model, the amount of evolutionary change that occurs is proportional to the amount of time since species shared a common ancestor.

- Shared Evolutionary History: The VCV matrix represents how much of the evolutionary history is shared between pairs of species. The diagonal elements represent the total amount of evolutionary change for each species (i.e., their own branch lengths), while the off-diagonal elements represent the shared evolutionary history between pairs of species (i.e., the branch lengths they share up to their most recent common ancestor).

Matrix Construction:

- Diagonal Elements: Each diagonal element of the matrix corresponds to the total branch length from the root of the tree to a given species. This reflects the total evolutionary variance that has accumulated along that lineage.

- Off-diagonal Elements: The off-diagonal elements correspond to the covariance between two species, which is proportional to the length of the branch they share from their most recent common ancestor.

The resulting matrix is a symmetric matrix where the diagonal elements are the variances (representing the evolutionary path length of each species), and the off-diagonal elements are the covariances (representing shared evolutionary history between species).


```{r}
# Calculate the phylogenetic covariance matrix
phylo_cov_matrix <- vcv(pruned.tree, corr = TRUE)
```


We only have phylogenetic data for a subset of species. To make the model run, we need to have the same number of species in the matrix than we are modeling. Species without phylogenetic data are added to the matrix with all 0:

```{r}
 # Match species in the tree to the species in your data
 tree_species <- colnames(phylo_cov_matrix)
 dataset_species <- unique(sp_code_island_df$species_name)
 
 # Ensure the order matches your dataset
 species_match <- match(dataset_species, tree_species)
 phylo_cov_matrix_final <- phylo_cov_matrix[species_match, species_match]
 
 colnames(phylo_cov_matrix_final) <- dataset_species
 rownames(phylo_cov_matrix_final) <- dataset_species
 
 # Replace NA values in the final covariance matrix with zeros
 phylo_cov_matrix_final[is.na(phylo_cov_matrix_final)] <- 0

```


change species names to species codes in the covariance matrix:

```{r}
 # Create a named vector to map species names to species codes
 name_to_code <- setNames(sp_code_island_df$species_code, sp_code_island_df$species)
 # Replace the row and column names in phylo_cov_matrix_final with species codes
 rownames(phylo_cov_matrix_final) <- name_to_code[rownames(phylo_cov_matrix_final)]
 colnames(phylo_cov_matrix_final) <- name_to_code[colnames(phylo_cov_matrix_final)]

```


The covariance matrix must be positive definite. To guarantee this, we add a small value to the diagonal elements:

```{r}
 epsilon <- 1e-6  # Regularization constant (can be adjusted)
 phylo_cov_matrix_final <- phylo_cov_matrix_final + diag(epsilon, nrow(phylo_cov_matrix_final))
# 
# # ensure all egeinvalues are epositive
 eigen_values <- eigen(phylo_cov_matrix_final)$values
 print(eigen_values)
```


# model brightness adding phylogenetic information


In the model, we treat the phylogenetic effects as random effects that influence the species-level traits (e.g., brightness or nematode load). We assume that these random effects are distributed according to a multivariate normal distribution with a covariance structure given by the phylogenetic covariance matrix.


```{r}
#| class-output: stan
all_spp_arbor_veg_bright_phylo <- cmdstan_model(here::here("stan/all_spp_arbor_veg_bright_phylo.stan"),
                                          pedantic = TRUE)

all_spp_arbor_veg_bright_phylo
```

data

```{r}
# Data code including phylogenetic covariance matrix
all_spp_arbor_veg_bright_phylo_samp <- all_spp_arbor_veg_bright_phylo$sample(
  data = list(
    N_spp = nrow(sp_code_island_df),  # Number of species (47)
    N_island = max(sp_code_island_df$island_code),
    sp_id = sp_code_island_df$species_code,
    island_area = sp_code_island_df$island_area,
    island_age = sp_code_island_df$island_age,
    island_index_spp = sp_code_island_df$island_code,
    # microhabitat
    N_habitat = nrow(df_hab),
    habitat_arboreal = df_hab$arboreal,
    total_hab = df_hab$total_hab,
    sp_index_hab = df_hab$species_code,
    # veg zone
    N_veg = nrow(df_veg),
    veg_arid = df_veg$arid,
    total_veg = df_veg$total_veg,
    sp_index_veg = df_veg$species_code,
    # brightness data
    N_bright = nrow(df_bright_noNA), 
    brightness_obs = df_bright_noNA$brightness,
    sp_index_bright = df_bright_noNA$species_code,
    # phylogenetic covariance matrix
    phylo_cov_matrix = phylo_cov_matrix_final  # Include the phylogenetic covariance matrix
  ),
  chains = 4, parallel_chains = 4, refresh = 0
)

```



```{r}
bright_rvars_phylo <- all_spp_arbor_veg_bright_phylo_samp |> 
  tidybayes::gather_rvars(true_ln_bright[species_code]) |> 
  mutate(brightness = exp(.value),
         species_code = forcats::fct_reorder(as.character(species_code), .value))


##################
# Create a named vector for species names
species_code_to_name <- setNames(sp_code_island_df$species_name, sp_code_island_df$species_code)

# Replace species codes with species names in bright_rvars
bright_rvars_phylo <- bright_rvars_phylo %>%
  mutate(species_name = species_code_to_name[as.character(species_code)])

# Replace species codes with species names in df_bright_noNA
df_bright_noNA <- df_bright_noNA %>%
  mutate(species_name = species_code_to_name[as.character(species_code)])
#################

post_bright_phylo <- bright_rvars_phylo |> 
  ggplot(aes(x = species_name, ydist = brightness)) + 
  stat_pointinterval() + 
  scale_y_log10()

post_bright_phylo <- post_bright_phylo + 
  geom_point(aes(x = species_name, y = brightness), inherit.aes = FALSE, 
             data = df_bright_noNA, 
             colour = "red",
             size = .5, alpha = .5,
             position = position_jitter(width = .3)) +
  theme_minimal() +
  my_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  xlab(NULL)

```





compare predicted brightness with and without phylogenetic information

```{r}

ggarrange(
  post_bright,
  post_bright_phylo,
  ncol = 1,
  nrow = 2,
  labels = LETTERS[1:2]
)

```


- negligible changes in brightness when adding phylogenetic information













###################################################
###################################################









# model load adding phylogenetic information



```{r}
#| class-output: stan
all_spp_load_phylo <- cmdstan_model(here::here("stan/all_spp_load_phylo.stan"),
                                          pedantic = TRUE)

all_spp_load_phylo
```


```{r}

# Data code including phylogenetic covariance matrix
all_spp_load_samp <- all_spp_load_phylo$sample(
  data = list(
  N_spp = nrow(sp_code_island_df),  # Number of species (47)
  N_island = max(sp_code_island_df$island_code),
  sp_id = sp_code_island_df$species_code,
  island_area = sp_code_island_df$island_area,
  island_age = sp_code_island_df$island_age,
  island_index_spp = sp_code_island_df$island_code,
  
  # Habitat data
  N_habitat = nrow(df_hab),
  habitat_arboreal = df_hab$arboreal,
  total_hab = df_hab$total_hab,
  sp_index_hab = df_hab$species_code,
  
  # Vegetation zone data
  N_veg = nrow(df_veg),
  veg_arid = df_veg$arid,
  total_veg = df_veg$total_veg,
  sp_index_veg = df_veg$species_code,
  
  # Brightness data
  N_bright = nrow(df_bright_noNA), 
  brightness_obs = df_bright_noNA$brightness,
  sp_index_bright = df_bright_noNA$species_code,
  
  # Nematode load data
  N_load = nrow(df_load),  # Total observations for nematode load
  nematode_count = df_load$nematode_count,
  sp_index_load = df_load$species_code,  # Use species_code from df_load
  island_index_load = df_load$island_code,  # Use island_code from df_load
  
  # Phylogenetic covariance matrix
  phylo_cov_matrix = phylo_cov_matrix_final  # Phylogenetic covariance matrix
),
  chains = 4, parallel_chains = 4, refresh = 0,
output_dir = here::here("output/models/missing_data_load_phylo")
)



```



```{r}
# read in the model from above
all_spp_load_samp <- as_cmdstan_fit(dir(here::here("output/models/missing_data_load_phylo"), full.names = TRUE))
```


## Predictios brightness

```{r}
bright_rvars_phylo <- all_spp_load_samp |> 
  tidybayes::gather_rvars(true_ln_bright[species_code]) |> 
  mutate(brightness = exp(.value),
         species_code = forcats::fct_reorder(as.character(species_code), .value))


##################
# Create a named vector for species names
species_code_to_name <- setNames(sp_code_island_df$species_name, sp_code_island_df$species_code)

# Replace species codes with species names in bright_rvars
bright_rvars_phylo <- bright_rvars_phylo %>%
  mutate(species_name = species_code_to_name[as.character(species_code)])

# Replace species codes with species names in df_bright_noNA
df_bright_noNA <- df_bright_noNA %>%
  mutate(species_name = species_code_to_name[as.character(species_code)])
#################

post_bright_phylo <- bright_rvars_phylo |> 
  ggplot(aes(x = species_name, ydist = brightness)) + 
  stat_pointinterval() + 
  scale_y_log10()

post_bright_phylo <- post_bright_phylo + 
  geom_point(aes(x = species_name, y = brightness), inherit.aes = FALSE, 
             data = df_bright_noNA, 
             colour = "red",
             size = .5, alpha = .5,
             position = position_jitter(width = .3)) +
  theme_minimal() +
  my_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  xlab(NULL)

post_bright_phylo
```




- The errors for species with missing data are too high - this generates NaN and can't plot it.


## Predictios load

```{r}

predicted_load <- all_spp_load_samp$draws("predicted_nematode_load")

# Convert the extracted predictions to a format suitable for bayesplot
predicted_load <- as_draws_matrix(predicted_load)

# Posterior predictive check for nematode load
ppc_load <-  ppc_dens_overlay(y = df_load$nematode_count, yrep = predicted_load) +
  scale_color_manual(name = "Legend",
                     values = c("y" = "black", "yrep" = "blue"),
                     labels = c("Observed load", "Predicted load")) +
  scale_fill_manual(name = "Legend",
                    values = c("y" = "black", "yrep" = "blue"),
                    labels = c("Observed load", "Predicted load"))+
  my_theme


```


```{r}
# Extract the posterior samples
posterior_samples <- all_spp_load_samp$draws()


posterior_df <- as_draws_df(posterior_samples)

#all_spp_load_samp$diagnostic_summary()

#all_spp_load_samp$summary()

mcmc_areas(
  posterior_df,
  pars = vars(starts_with("slope")),
  prob = 0.8  
)
```


- curves are extremely flat, except for brightness which has an extremely narrow variance.


# Add constrains

Trying first with a couple of constrains on brightness supported by Kraemer et al. 2019. We know snails are brighter if they are arboreous, and that brightness decreases with island age. Constrains are added to slope priors:

slope_arbor_bright ~ normal(0, 0.5) T[0, ];  // Biased towards positive (Kraemer et al. 2019)
slope_age_bright ~ normal(0.5, 0) T[, 0];  // Biased towards negative (Kraemer et al. 2019)


```{r}
#| class-output: stan
all_spp_load_phylo_constrained_bright <- cmdstan_model(here::here("stan/all_spp_load_phylo_constrained_bright.stan"),
                                          pedantic = TRUE)

all_spp_load_phylo_constrained_bright
```


```{r}

# Data code including phylogenetic covariance matrix
all_spp_load_constrained_bright_samp <- all_spp_load_phylo_constrained_bright$sample(
  data = list(
  N_spp = nrow(sp_code_island_df),  # Number of species (47)
  N_island = max(sp_code_island_df$island_code),
  sp_id = sp_code_island_df$species_code,
  island_area = sp_code_island_df$island_area,
  island_age = sp_code_island_df$island_age,
  island_index_spp = sp_code_island_df$island_code,
  
  # Habitat data
  N_habitat = nrow(df_hab),
  habitat_arboreal = df_hab$arboreal,
  total_hab = df_hab$total_hab,
  sp_index_hab = df_hab$species_code,
  
  # Vegetation zone data
  N_veg = nrow(df_veg),
  veg_arid = df_veg$arid,
  total_veg = df_veg$total_veg,
  sp_index_veg = df_veg$species_code,
  
  # Brightness data
  N_bright = nrow(df_bright_noNA), 
  brightness_obs = df_bright_noNA$brightness,
  sp_index_bright = df_bright_noNA$species_code,
  
  # Nematode load data
  N_load = nrow(df_load),  # Total observations for nematode load
  nematode_count = df_load$nematode_count,
  sp_index_load = df_load$species_code,  # Use species_code from df_load
  island_index_load = df_load$island_code,  # Use island_code from df_load
  
  # Phylogenetic covariance matrix
  phylo_cov_matrix = phylo_cov_matrix_final  # Phylogenetic covariance matrix
),
  chains = 4, parallel_chains = 4, refresh = 0,
output_dir = here::here("output/models/missing_data_load_phylo_constrained_bright")
)



```



```{r}
# read in the model from above
all_spp_load_constrained_bright_samp <- as_cmdstan_fit(dir(here::here("output/models/missing_data_load_phylo_constrained_bright"), full.names = TRUE))
```




```{r}
# Extract the posterior samples
posterior_samples <- all_spp_load_constrained_bright_samp$draws()


posterior_df <- as_draws_df(posterior_samples)

#all_spp_load_constrained_samp$diagnostic_summary()

#all_spp_load_constrained_samp$summary()

mcmc_areas(
  posterior_df,
  pars = vars(starts_with("slope")),
  prob = 0.8  
)
```



# Add further constrains

We add further constrains based on the nematode's biology (supported by studies, but not referenced here).

slope_arid_load ~ normal(0.5, 0) T[0, ];  // Biased towards negative
slope_arbor_load ~ normal(0.5, 0) T[, 0];  // Biased towards negative

```{r}
#| class-output: stan
all_spp_load_phylo_constrained_bright_load <- cmdstan_model(here::here("stan/all_spp_load_phylo_constrained_bright_load.stan"),
                                          pedantic = TRUE)

all_spp_load_phylo_constrained_bright_load
```


```{r}

# Data code including phylogenetic covariance matrix
all_spp_load_constrained_bright_load_samp <- all_spp_load_phylo_constrained_bright_load$sample(
  data = list(
  N_spp = nrow(sp_code_island_df),  # Number of species (47)
  N_island = max(sp_code_island_df$island_code),
  sp_id = sp_code_island_df$species_code,
  island_area = sp_code_island_df$island_area,
  island_age = sp_code_island_df$island_age,
  island_index_spp = sp_code_island_df$island_code,
  
  # Habitat data
  N_habitat = nrow(df_hab),
  habitat_arboreal = df_hab$arboreal,
  total_hab = df_hab$total_hab,
  sp_index_hab = df_hab$species_code,
  
  # Vegetation zone data
  N_veg = nrow(df_veg),
  veg_arid = df_veg$arid,
  total_veg = df_veg$total_veg,
  sp_index_veg = df_veg$species_code,
  
  # Brightness data
  N_bright = nrow(df_bright_noNA), 
  brightness_obs = df_bright_noNA$brightness,
  sp_index_bright = df_bright_noNA$species_code,
  
  # Nematode load data
  N_load = nrow(df_load),  # Total observations for nematode load
  nematode_count = df_load$nematode_count,
  sp_index_load = df_load$species_code,  # Use species_code from df_load
  island_index_load = df_load$island_code,  # Use island_code from df_load
  
  # Phylogenetic covariance matrix
  phylo_cov_matrix = phylo_cov_matrix_final  # Phylogenetic covariance matrix
),
  chains = 4, parallel_chains = 4, refresh = 0,
output_dir = here::here("output/models/missing_data_load_phylo_constrained_bright_load")
)



```



```{r}
# read in the model from above
all_spp_load_constrained_bright_load_samp <- as_cmdstan_fit(dir(here::here("output/models/missing_data_load_phylo_constrained_bright_load"), full.names = TRUE))
```




```{r}
# Extract the posterior samples
posterior_samples <- all_spp_load_constrained_bright_load_samp$draws()


posterior_df <- as_draws_df(posterior_samples)

#all_spp_load_constrained_samp$diagnostic_summary()

#all_spp_load_constrained_samp$summary()

mcmc_areas(
  posterior_df,
  pars = vars(starts_with("slope")),
  prob = 0.8  
)
```


